---
title: "Graphs for Animal Crrossing Presentation"
author: "Sidney"
date: "2023-03-05"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
library(shiny)
library(shinydashboard)
library(lubridate)
```

## Loading the data
```{r message=FALSE, warning=FALSE}
bugs <- clean_names(read_csv(here("data","insects.csv")))
fish <- clean_names(read_csv(here("data","fish.csv")))
```

## Fixing error in data
```{r}
bugs[65,34] <- "Ins45"
```

## Cleaning up the "bugs" data frame
```{r}
bugs <- bugs %>% 
  select(-"critterpedia_filename", -"furniture_filename", -contains("id"), -contains("sh"), -contains("color"), -"number") %>% 
  mutate(category = "bugs")
```

## Cleaning up the "fish" data frame
```{r}
fish <- fish %>% 
  select(-"critterpedia_filename", -"furniture_filename", -contains("id"), -contains("sh"), -contains("color"), -"number", -"lighting_type", -"rain_snow_catch_up", -"size") %>% 
  mutate(category = "fish")
```

## Combining "fish" and "bugs" into the "museum" data frame
```{r message=FALSE, warning=FALSE}
museum <- full_join(fish, bugs)
```

## Pivoting the months & times into 2 columns
```{r}
 museum <- museum %>%
  pivot_longer(nh_jan:nh_dec,
               names_to = "month",
               names_prefix = "nh_",
               values_to = "times")
```

## Separating the "times" column into "start times" and "end times"
```{r message=FALSE, warning=FALSE}
museum <- museum %>% 
  filter(times != "NA") %>% 
  mutate_all(~ gsub("\\ ", " ", .)) %>% 
  replace(. == "All day", "0 AM – 24 AM") %>% 
  separate(times, into = c("start_times","end_times"), sep = " – ") %>% 
  separate(start_times, into = c("start_times","start_td"), sep = " ") %>% 
  separate(end_times, into = c("end_times","end_td"), sep = " ")
```

## Converting "start times" and "end times" to numerics
```{r}
museum$start_times <- as.numeric(museum$start_times)
museum$end_times <- as.numeric(museum$end_times)
```

## Converting all times to 24 hours
```{r}
museum <- museum %>% 
  mutate(start_times = ifelse(start_times == 12, 0, start_times)) %>% 
  mutate(start_times = ifelse(start_td == "PM",start_times+12,start_times)) %>%
  mutate(end_times = ifelse(end_times == 12, 0, end_times)) %>% 
  mutate(end_times = ifelse(end_td == "PM",end_times+12,end_times)) %>% 
  select(-"start_td",-"end_td")
```

## Converting "months" to factors based on the built in "month.abb" levels
```{r}
month.abb.l <- tolower(month.abb)
museum$month <- factor(museum$month, levels = month.abb.l)
```

## Adding ".png" to the every "icon_filename" value
```{r}
museum <- museum %>% 
  mutate(icon_filename = paste(museum$icon_filename, ".png", sep = ""))
```



##Graphs



#Graph of Avg Sell Price in Each Month + Hour
```{r}

museum$sell <- as.numeric(museum$sell)

sum_museum <- museum %>% 
  mutate(end_times = ifelse(museum$start_times > museum$end_times,museum$end_times + 24,museum$end_times)) %>% 
  group_by(name) %>% 
  summarize(n_months = mean(n_distinct(month)),
            n_hours = mean(end_times - start_times),
            sell = mean(sell)) 

sum_museum$n_hours <-as.factor(sum_museum$n_hours)
sum_museum$n_months <-as.factor(sum_museum$n_months)

sum_museum %>%
  ggplot(aes(x=n_months,y=sell, fill=n_months))+
  geom_boxplot(aes(group=n_months))+
  theme_classic()+
  theme(legend.position="none")+
  labs(title = "Average Sell Price Within a Number of Months",
       x = "Months",
       y = "Sell Price",
       color=NULL)
  

sum_museum %>% 
  ggplot(aes(x=n_hours,y=sell, fill=n_hours))+
  geom_boxplot(aes(group=n_hours))+
  theme_classic()+
  theme(legend.position="none")+
  labs(title = "Average Sell Price Within Hour Availability",
       x = "Hours",
       y = "Sell Price",
       color=NULL)
```

#Density Curve Representing the Selling Prices of Each Location For Fish
```{r}
museum <-museum %>% 
  filter(category=="fish") %>% 
  mutate(where=case_when(where_how == "River (clifftop)"~"river",
                        where_how =="River (mouth)" ~ "river" ,
                        where_how == "River" ~ "river",
                        where_how == "Sea" ~ "sea",
                        where_how == "Sea (rainy days)" ~ "sea",
                        where_how == "Pier" ~ "sea",
                        where_how == "Pond" ~ "pond"))

  ggplot(museum, aes(x=sell))+
  geom_density(data=filter(museum, where_how =="River" | where_how == "River (clifftop)"), color = "slateblue3")+
  geom_density(data=filter(museum, where_how =="River" | where_how == "River (mouth)"), color = "salmon1")+
  geom_density(data=filter(museum, where_how =="River"), color = "gray12")+
  geom_density(data=filter(museum, where_how =="Sea") , color = "gray12")+
  geom_density(data=filter(museum, where_how =="Sea" | where_how == "Sea (rainy days)"), color = "mediumvioletred")+
    geom_density(data=filter(museum, where_how =="Sea" | where_how == "Pier"), color = "darkgoldenrod1")+
    geom_density(data=filter(museum, where_how =="Pond"), color = "gray12")+
  facet_wrap(~where, ncol=3)+
  theme_classic()+
  theme(legend.position="none")+
  labs(title = "Density of Fish Sell Price by Location",
       x = "Sell Price",
       y = NULL,
       color=NULL) 
#River and River (clifftop) should overlap -- river fish still appear in river clifftop and river mouth
#same goes for sea and sea rainy days because there are certain fish that appear only in rainy days 
```






