---
title: "Data Structures"
author: "Joel Ledford"
date: "Winter 2020"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. 

## Review


## Data Structures
In addition to the types or classes of data, R also organizes data in specific ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. For now, we are going to focus on `vectors`.  

## Vectors
Vectors are a common way of organizing data in R. The key detail is that they contain data *of the same type*. We create vectors using the `c` command.  

A numeric vector.
```{r}
my_vector <- c(10, 20, 30)
```

We can make character vectors as well.  
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thrusday", "Friday", "Saturday", "Sunday")
```

What if our numeric vector above had a missing value?  
```{r}
my_vector <- c(10, NA, 30)
anyNA(my_vector)
```

Certain functions require us to specify how we want to deal with NAs. Often if NA is included in an operation, the result is NA.
```{r}
mean(my_vector)
```

Let's remove the NA value.
```{r}
mean(my_vector, na.rm = TRUE)
```

Another way.  
```{r}
my_new_vector <- na.omit(my_vector)
my_new_vector
```

A convenient trick for creating a vector is to generate a sequence of numbers.
```{r}
my_vector_sequence <- c(1:10)
my_vector_sequence
```

`seq` allows more control.
```{r}
my_vector_sequence2 <- seq(1, 4, by=0.25)
my_vector_sequence2
```

## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.  
```{r}
days_of_the_week[3]
```

## Practice
Work through these examples and make adjustments to the values to experiment.

1. Create a new numeric vector of values 1 through 100
```{r}
my_new_vector_sequence <- c(1:100)
my_new_vector_sequence
```

2. We can evaluate all numerics in our vector to determine which has a value of 15.
```{r}
my_new_vector_sequence==15
```

3. We can also use operators such as <, >, ==, <==, etc. Which values are less than or equal to 15?  
```{r}
my_new_vector_sequence<=15
```

4. If you use `[]` then you only get the values, not the logical evaluation of the entire vector.
```{r}
my_new_vector_sequence[my_new_vector_sequence<=15]
```

5. Using the `&` symbol allows us to refine our elements of interest.
```{r}
my_new_vector_sequence[my_new_vector_sequence <= 15 & my_new_vector_sequence >= 1]
```

## That's it, let's take a break!  

-->[Home](https://jmledford3115.github.io/datascibiol/)