---
title: "Getting Started"
author: "Joel Ledford"
date: "Winter 2020"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

## Background
In nearly every field of science, our ability to generate data has exceeded our capacity for analysis. For me, this means that there is the potential for loss to science; many important discoveries may go unnoticed because we are unable to efficiently analyze data.

There is also a growing problem of curation and reproducibility. As the analyses and data become more complex, research labs have a problem of managing data and analysis such that data can be used in future analyses or shared with colleagues. As students, postdocs, and visiting scientists come and go in labs, or even as an individual moves between projects, there is often a mess left behind where data are inefficiently stored, analysis was not documented, or in the worse case scenario, sloppy copy/paste type errors cause wasted days to months. Effectively stored data along with a documented, reproducible, work flow will allow for high-quality analysis, error prevention and time savings, and smooth and efficient collaboration with others or even with yourself months to years in the future.  

Lastly, with the widespread availability of data online there is the potential for new scientific discovery or insight through exploratory data analysis. In many cases a clean, organized workflow will yield new perspectives and promising results.  

## Objectives
The goal for this course is to help get you started learning to manage, transform, and visualize data using the R programming language. You will learn to clearly and neatly organize messy data, transform it in ways that address your questions, and communicate results in a variety of formats. The course is designed for people with **no prior programming experience**. There is a substantial learning curve but, working together, we should be able to make learning R easier, interesting, and fun.

Why spend the time? If you are pursuing any research path in science, this class will give you a competitive advantage and save you countless hours as most people struggle to [learn R on their own](https://www.wired.com/2017/03/biologists-teaching-code-survive/). Even if you are not pursuing a research career, companies routinely hire data scientists (or people with skills in data science) to analyze a variety of data.

## Resources
Given that R is open source, many resources are available online. We will use a combination of resources in the class, but key items are listed below.

- [R for data science](https://r4ds.had.co.nz/)
- [R cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
- [RStudio keyboard shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)

## Data Science at UC Davis
Our campus has a wealth of expertise in data science, from working groups to full initiatives. Should your interests progress, here are some links. They offer regular workshops and maintain archives.

- [Data Science Initiative, UC Davis](http://dsi.ucdavis.edu/)  
- [Davis R Users Group](http://d-rug.github.io/) 

## Lab 1
The goal of lab 1 is to get everyone started using R, Rstudio, and GitHub. All of our work will be done in RStudio and uploaded to the class GitHub repository.  

At the start of each lab session, I will randomly pull someone's code from the class GitHub repository and use it as an example. There are many, many ways to solve the same problems in R and seeing other people's code is a great way to learn. This is not about shaming people who make mistakes!  

This first lab is tedious, so please be patient. It is important that everyone is set-up correctly before they leave today. In the spirit of the R universe, our class is a community. If you see someone struggling, please give them some help.  

## Setup your own computer
You are welcome to use your own computer, but in order to do so you need to install R, RStudio, and GitHub Desktop. If you would like to set-up your own computer, please follow the directions [here](https://jmledford3115.github.io/datascibiol/setup.html). If you are unsure about troubleshooting your own computer, **please** use the lab computers.  

## GitHub
[**GitHub**](www.github.com) is a file storage and management site used mostly by programmers. Programmers upload code to repositories (folders) and make it publicly available. We will use Github to store our work in BIS 15.  

## Create a GitHub account
Since we will use GitHub as a repository for all assignments, we need to make an account for each student and build a new project through RStudio. Navigate to www.github.com and create a free account. [Email](mailto:jmledford@ucdavis.edu) your username to me. You will receive an email to join the repository for our course.  

I have made a repository for our class titled: [BIS15-W20-DataScienceBiologists](https://github.com/jmledford3115/BIS15-W20-DataScienceBiologists). Once you have accepted the invitation to join the repository, you should be able to login to GitHub and see this organization or click on the link above.  

## Make Your Own Repository  
1. Go to [GitHub](www.github.com) and login.  
2. Create a new repository.  
3. Under repository name title it `BIS15W2020_your first initial and lastname`. Mine would look like this: BIS15W2020_jledford  
4. Default settings are fine, but do click "Initialize this repository with a README"  
5. Once the repository is made, click on settings in the top right  
6. On the left toolbar click "Collaborators"  
7. Add "rlgodwin28" and "jmledford3115" as collaborators.  

## GitHub Desktop
The easiest way to manage your GitHub repositories is to use [GitHub desktop](https://desktop.github.com/). This is a  helpful tool that will allow you to manage all of your files. Highly recommended! Once installed, open the program and login using your GitHub account.  

## Clone `BIS15-W20-DataScienceBiologists`
In our class GitHub, you will see a directory called `BIS15-W20-DataScienceBiologists`. From within GitHub Desktop, clone this folder to your desktop. Also clone your new personal repository `BIS15W2020_your first initial and lastname`. At the start of each class you will need to repeat this process.  

## RStudio
Inside the `BIS15-W20-DataScienceBiologists` folder you will see a folder called `class files`. Within this there are several files. Copy and paste all of these files into your personal repository `BIS15W2020_your first initial and lastname`.  

RStudio is an interactive working environment for R. When you first open it, you should see your screen divided into four quadrants. I will demonstrate each of these in class, but for now it is enough to know that they display different information and are helpful as you work on code.  

Please run the following code to check your current working directory.  
```{r}
getwd()
```

If you find that you are not in the correct folder, please navigate to it: Session>Set Working Directory>Choose Directory. You want to be in your personal repository `BIS15W2020_your first initial and lastname`.   

Re-run the code below to confirm.
```{r}
getwd()
```

## Push files to GitHub
You can use your repository to store all of your notes and work. If you setup GitHub Desktop correctly, then you will see that it has automatically tracked your changes. Clicking the `Commit to master` button in the bottom left corner means that you are recording a waypoint in your edits locally. Once this is done, you need to upload your changes to GitHub. This is done by cliking `Push origin`. Once you have pushed your changes you can double check by looking on GitHub.  

## Practice
Since you will use your GitHub repository for your homework, we need to practice.  

1. Navigate to the `BIS15-W20-DataScienceBiologists` folder on your computer.  
2. Open the folder called `class_files`.  
3. Copy all of the files in this folder.  
4. Paste the files into **your** repository.  
5. Open `dummy_push` in your repository and run the code.  
6. Add your name under `author` at the top of the page.  
8. Save this file.  
9. Knit following class instructions.  
10. Use GitHub Desktop to first `Commit` and then `push` to your GitHub repository.    

## GitHub Issues
We may encounter  issues over the next few weeks with GitHub. These will be caused mostly by user error, but sometimes weird things happen. If you are stuck, get my attention and we will sort things out.  

## That's it, let's take a break!  

--> On to [part 2](https://jmledford3115.github.io/datascibiol/lab1_2.html)  