---
title: "Data Types and Structures"
author: "Joel Ledford"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

## Essential Information about R
R is an open source, **scripting** language. RStudio is a GUI (graphical user interface) that is frequently used to interact with R. You need R installed in order for RStudio to work.

Because R is a scripting language you don't point and click to run commands. Instead, you write instructions that are used by the computer to complete an operation. This makes for excellent repeatability but makes it hard to learn at first because the instructions are specific. R is **not** tolerant of typos, puntuation errors, errant spaces, or other types of mistakes.

## RMarkdown
Experienced programmers often write their code with minimal annotations. This is great for them because they read R fluently. Since we are programming novices we will write our code in RMarkdown format. RMarkdown allows us to embed code in annotated chunks, including display of graphical output. The RMarkdown file can then be exported in a number of formats. If you have questions, the definitive guide is [here](https://bookdown.org/yihui/rmarkdown/).  

## Help
R is a flexible programming language that is used by scientists worldwide. Thousands of packages are available to perform specific analyses and there is a large community of people willing to help if you have problems. Just be sure to thoroughly look for help before posting questions to stack overflow or related forums.

Adding a '?' before a command will bring up the associated help document.
```{r}
?mean
```

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Solve arithmetic operations in R.  
2. Define an object in R; assign values to objects.  
3. Learn best practices in the naming of objects and other R programming conventions.  
4. Learn the differences between the types of data used in R.  
5. Perform simple operations using vectors.  

## Arithmetic
You can use R as a calculator; just enter the operation into a new code chunk. Order of operations applies and you don't need to add an '=' sign.
```{r}
4*12
```

```{r}
(4*12)/2
```

Statistics are what R was originally written for, so it will pretty much do anything you need and there are thousands of packages for specialized statistics depending on your needs.
```{r}
mean(4, 6, 8, 5, 6, 7, 7, 7)
```

## Objects
In order to access the potential of R we need to assign values or other types of data to `objects`. There is a specific format that I want you to follow, so please pay close attention.  

Assign a value to object 'x'. The '<-' symbol is read as 'gets'. In this case, x gets 42. Make sure that you are in the environment panel and you should see the value associated with 'x'. On a mac, you can push `option` and `-` to automatically generate the gets symbol.
```{r}
x<-42
```

To print the object to the screen, just type x. Alternatively, you can use the print command.
```{r}
x
print(x)
```

Once an object has been created, you can do things with them.
```{r}
treatment<-36
control<-38
```

```{r}
my_experiment<-treatment+control
my_experiment
```

## Nomenclature
We need to be careful about nomenclature when we write code. R allows us to give almost any name we want to an object, but there are exceptions. For example, we don't want to give a name to an object that is the same as a function in R.

```{r eval=FALSE, include=TRUE}
else<-12
```

We get an error here because `else` is a function in R. You also don't want to give names that might get confused with functions; i.e. you can assign a value to 'mean' but this could become confusing because mean is used as a function.  

## Practice
1. Create three new objects, `venom_GT`, `chiron`, and `veyron`. These are the fastest cars in the world. Assign each car to its top speed. The venom_GT can go 270, chiron is 261, and veyron is 268.
```{r}
venom_GT <- 270
chiron <- 261
veyron <- 268
```

2. Use arithmetic to calculate the mean top speed for the cars.
```{r}
(270+261+268)/3
```

3. Use the function `mean()` to calculate the mean top speed for the cars.
```{r}
mean(venom_GT, chiron, veyron)
```

## Types of Data
There are five frequently used `classes` of data: 1. numeric, 2. integer, 3. character, 4. logical, 5. complex.
```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe"
my_logical <- FALSE
my_complex <- 2+4i
```

To find out what type of data you are working with, use the `class()` function. This is important because sometimes we will need to change the type of data to perform certain analyses.
```{r}
class(my_numeric)
```

You can use the `is()` and `as()` functions to clarify or specify a type of data.
```{r}
is.integer(my_numeric) #is my_numeric an integer?
```

```{r}
my_integer <- 
  as.integer(my_numeric) #create a new object specified as an integer
```

```{r}
is.integer(my_integer) #is my_numeric an integer?
```

## Data Structures
In addition to the types or classes of data, R also organizes data in specific ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. For now, we are going to focus on `vectors`.  

## Vectors
Vectors are a common way of organizing data in R. The key detail is that they contain data *of the same type*. We create vectors using the `c` command.  

A numeric vector.
```{r}
my_vector <- c(10, 20, 30)
```

We can make character vectors as well.
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thrusday", "Friday", "Saturday", "Sunday")
```

A convenient trick is to generate a string of numbers.
```{r}
my_vector_string <- c(1:10)
my_vector_string
```

`seq` allows more control.
```{r}
my_vector_string2 <- seq(1, 4, by=0.25)
my_vector_string2
```

## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.
```{r}
days_of_the_week[3]
```

## Practice
Work through these examples and make adjustments to the values to experiment.

1. Create a new numeric vector of values 1 through 100
```{r}
new_vector <- 1:100
```

2. We can evaluate all numerics in our vector to determine which has a value of 15.
```{r}
new_vector == 15
```

3. We can also use operators such as <, >, ==, <==, etc.
```{r}
new_vector < 10
```

4. If you use `[]` then you only get the values, not the logical evaluation of the entire vector.
```{r}
new_vector[new_vector <=10]
```

5. Using the `&` symbol allows us to refine our elements of interest.
```{r}
new_vector[new_vector <= 10 & new_vector >= 1]
```

6. There is a distinction between the `&` symbol and the `|` symbol.
```{r}
new_vector[new_vector <= 10 | new_vector == 1]
```

## Wrap-up
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  

-->[Home](https://jmledford3115.github.io/datascibiol/)