---
title: "Data Frames"
author: "Joel Ledford"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Setup
At the beginning of each class, please make sure that you navigate to the class [GitHub](https://github.com/FRS417-DataScienceBiologists) and download the `class_files` folder. In our class GitHub, you will see a directory called `class_files`. Click on the `Clone or download` button and download the files as a `.zip`. You should now have a folder called `class_files` on your computer inside of which are all of the files that we will use for today's class.  

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Combine a series of vectors into a data frame.  
2. Name columns and rows in a data frame.  
3. Select columns and rows and use summary functions.   
4. Change data classes of columns in the data frame.  
5. Write your data frame to a csv file!  

## Data Frame
Last time, you learned how to work with vectors. Today we will organize the vectors into a new type of data structure called a data frame. The data frame is the most common way to store data within R. You can think of a data frame as similar to an Excel spreadsheet. A data frame can store data of different types. 

Let's build three separate vectors of length (in), weight (oz), and sex of three different ruby throated hummingbirds.
```{r}
Length <- c(3.2, 3.7, 3.4)
Weight <- c(2.9, 4.0, 3.1)
Sex <- c("Male", "Female", "Male")
```

Here we combine our three vectors to create a data frame with the function `data.frame()`.
```{r}
hbirds <- data.frame(Length, Weight, Sex)
hbirds
```

Notice that not only are the data neat and clean looking, there is also information provided about the type of data in the frame. `dbl` means that the value is a type of numeric [double precision floating point](http://uc-r.github.io/integer_double/).

What are the column names of our data frame? Notice that R defaulted to using the names of our vectors, but we could name them something else when creating the data frame, or rename them later.
```{r}
names(hbirds)
```

Let's use lowercase names when we create the data frame. We just changed to lowercase here, but we could use any names we wish.
```{r}
hbirds <- data.frame(length = Length, weight = Weight, sex = Sex)
names(hbirds)
```

## Renaming Column Names
We should rename the length and width column so we can remember the units. Recall from lab 1 that we can index a vector with `[]`. We don't need to rename sex, so we will index the column names for only the first and second column.
```{r}
names(hbirds)[1:2] <- c("length_in", "weight_oz")
names(hbirds)
```

## Accessing Data Frame Columns and Rows 

The same methods of selecting elements in a vector apply to data frames. We use `[]`. But now we have two positions where the first applies to the rows, and the second to the columns. 
```{r}
hbirds[1,]
hbirds[,1]
```

We can also select values in an entire column using the `$` sign. This can be useful for calculations. Here we calculate the mean of all of the birds weights. 
```{r}
w <- hbirds$weight_oz
mean(w)
```

## Summaries

We can also summarize our entire data frame with the`summary()` function.
```{r}
summary(hbirds)
```

## Adding Rows

We can add a new bird's data with the `rbind()` function, known as row bind. It will bind our new vector to our data frame row-wise.
```{r}
new_bird <- c(3.6, 3.9, "Female")
hbirds <- rbind(hbirds, new_bird)
hbirds
```


## Adding Columns

What if we come across more data and we want to add it to our data frame? We can do that with the `$` operator.
```{r}
hbirds$neighborhood <- c("Lakewood", "Brentwood", "Lakewood", "Scenic Heights")
hbirds
summary(hbirds)
```

## Changing Column Classes

Great, we added our new column, but it's a character and not a factor. A factor is a special type of character vector with predefined levels. 

```{r}
class(hbirds$neighborhood)
hbirds$neighborhood <- as.factor(hbirds$neighborhood)
class(hbirds$neighborhood)
levels(hbirds$neighborhood)
summary(hbirds)
```


## Writing Data to File

We should save our hbirds data frame so we can use it again later! There are many ways to save data in R, here we write our data frame to a csv file. We use  `row.names = FALSE` to avoid row numbers from printing out. 
```{r}
write.csv(hbirds, "hbirds_data.csv", row.names = FALSE)
```

## Practice
1. Below are data collected by three scientists (Jill, Steve, Susan in order) measuring temperatures of three hot springs near Mono Lake.
```{r}
temp <- c(36.25, 35.40, 35.30, 35.15, 35.35, 33.35, 30.70, 29.65, 29.20)
name <- c("Jill","Susan", "Steve", "Jill","Susan", "Steve", "Jill","Susan", "Steve")
spring <- c("Buckeye", "Buckeye", "Buckeye", "Benton", "Benton", "Benton", "Travertine", "Travertine", "Travertine")

```

2. Build a data frame called `hsprings` with the above data. Name the temperature column `temp_C`. Print out the data frame. 
```{r}
hsprings <- data.frame(temp_C = temp, name, spring)
hsprings
```

3. Change the column titled `name` to `scientist`, leave the other column names the same. Print out the data frame names.
```{r}
names(hsprings)[2] <- c("scientist")
names(hsprings)
```

4. Our scientists forgot to record the depth data for each spring. Here it is as a vector, add it as a new column called depth_ft: `c(4.15, 4.13, 4.12, 3.21, 3.23, 3.20, 5.67, 5.65, 5.66)`. Print out the data frame.
```{r}
hsprings$depth_ft <- c(4.15, 4.13, 4.12, 3.21, 3.23, 3.20, 5.67, 5.65, 5.66)
hsprings
```

5. Calculate the mean temperature of all of the temperature measurements.
```{r}
mean(hsprings$temp_C)
```

6. What is the class of the scientist column. Use an R command to see all the unique levels.
```{r}
class(hsprings$scientist)
levels(hsprings$scientist)
```

7. Save your hot springs data as a csv file! Do not allow row names. 
```{r}
write.csv(hsprings, "hsprings_data.csv", row.names = FALSE)
```

  

## That's it, let's take a break!   

--> On to [part 2](https://jmledford3115.github.io/datascibiol/lab2_2.html)  