---
title: "Dplyr"
author: "Joel Ledford"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Review
In the last section you were introduced to the **tidyverse** and the function `filter()`. `filter()` works by looking through rows, but what about columns? Recall that each column is a vector, so we need the new function `select()` to separate columns of interest. As you may expect, we can use `filter()` and `select()` together to build data frames focused on our variables of interest.

## Resources
- [R for Data Science](https://r4ds.had.co.nz/)

## Learning Goal
*At the end of this exercise, you will be able to:*    
1. Use the `filter()`, `select()`, and `arrange()` functions of dplyr to organize and sort data frames. 

## Libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Load the fish scale data
Here we load a .csv file with data on fish scales. We open the file using `read_csv()`. 
```{r message=FALSE, warning=FALSE}
fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")
```

## Dplyr: select()
Select allows you to build a new data frame by selecting your columns (variables) of interest. Our fish data only has six columns, but this should give you some ideas especially when you have large data frames with lots of columns.  

We are only interested in lakeid and scalelength.
```{r}
select(fish, lakeid, scalelength)
```

The `-` operator is useful in select. It allows us to select everything except the specified variables.
```{r}
select(fish, -fish_id, -annnumber, -length, -radii_length_mm)
```

For very large data frames with lots of variables, `select()` uses many different operators to make things easier. Let's say we are only interested in the variables that deal with length.
```{r}
select(fish, contains("length"))
```

Some additional options to select columns based on a specific criteria include:  
1. ends_with() = Select columns that end with a character string  
2. contains() = Select columns that contain a character string  
3. matches() = Select columns that match a regular expression  
4. one_of() = Select columns names that are from a group of names  

## Practice
1. What are the names of the columns in the `fish` data?


2. We are only interested in the variables `lakeid`, `length`, and `scalelength`. Use `select()` to build a new dataframe focused on these variables.


## Dplyr: Can we combine filter() and select()?
Absolutely. This is one of the strengths of the tidyverse, it uses the same grammar to specify commands.
```{r}
fish2 <- select(fish, lakeid, scalelength)
```

```{r}
filter(fish2, lakeid=="AL")
```

## Dplyr: Pipes %>% 
The code above works fine but there is a more efficient way. We need to learn pipes `%>%`. Pipes allow you to feed the output from one function to the input of another function. We are going to use pipes from here on to keep things cleaner. (command+shift+m)
```{r}
fish%>%
  select(lakeid, scalelength)%>%
  filter(lakeid=="AL")
```

## Practice
1. Build a dataframe focused on the lakes `AL` and `AR` and looking at radii length between 2 and 4 only. Use pipes!


## Dplyr: arrange()
The `arrange()` command allows us to sort values in a column.
```{r}
fish %>% 
  arrange(desc(scalelength))
```

It can be very helpful in combination with the other functions.
```{r}
fish %>% 
  select(lakeid, length, fish_id, scalelength) %>% 
  filter(lakeid=="AL") %>% 
  arrange(fish_id)
```

## Practice
1. Build a dataframe focused on the scalelengths of `fish_id` 300 and 301. Use `arrange()` to sort from smallest to largest scalelength. Use pipes!


## Wrap-up
Please review the learning goals and be sure to use the code here as a reference when completing the homework.

See you next time!