---
title: "Tidy Data 2 and Summarize"
author: "Joel Ledford"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Group Project
Let's take 10 minutes to decide on groups and think about a theme for our projects. What data interest you? Make a plan on doing some  internet searches and selecting a data set that will work for your group. Think about the kinds of questions you might ask.  

## Resources
- [dplyr-tidyr tutorial](http://tclavelle.github.io/dplyr-tidyr-tutorial/)
- [Data wrangling cheatsheet (`dplyr`,`tidyr`)](http://ucsb-bren.github.io/refs/cheatsheets/data-wrangling-cheatsheet.pdf)
- [Tidyr, R for Data Science](https://r4ds.had.co.nz/tidy-data.html#unite)

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Explain the difference between tidy and messy data.  
2. Evaluate a dataset as tidy or untidy.  
3. Use the `spread()` function of tidyr to transform messy data to tidy data. 
4. Use `summarize()` and `group_by()` to produce statistical summaries of data.

## Review
Last time we learned the fundamentals of tidy data and used the `gather()` function to wrangle one type of frequently encountered untidy data. Today we will use the `spread()` function to deal with a second type of untidy data.  

## Load the tidyverse
```{r}
library("tidyverse")
```

## Tidyr
+ `gather()` and `spread()` convert data between wide and long format  
+ `separate()` and `unite()` separates or unites information in columns 

## gather()
Recall that we use `gather()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
?USArrests
```

```{r}
arrests <- 
  USArrests %>% 
  mutate(State=rownames(USArrests)) %>% 
  select(State, Murder, Assault, Rape)
arrests
```

## Practice
1. Are these data tidy? Please use `gather()` to tidy the data.
```{r}
arrests_tidy <- 
  arrests %>% 
  gather(Murder, Assault, Rape, key="Crime", value="per100k")
```

2. Restrict the data to assualt only. Sort in ascending order.
```{r}
arrests_tidy %>%
  filter(Crime=="Assault") %>% 
  arrange(per100k)
```

## spread()
The opposite of `gather()`. You use `spread()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.
```{r}
country <- c("Afghanistan", "Afghanistan", "Afghanistan", "Afghanistan", "Brazil", "Brazil", "Brazil", "Brazil", "China", "China", "China", "China")
year <- c("1999", "1999", "2000", "2000", "1999", "1999", "2000", "2000", "1999", "1999", "2000", "2000")
key <- c("cases", "population", "cases", "population", "cases", "population", "cases", "population", "cases", "population", "cases", "population")
value <- c(745, 19987071, 2666, 20595360, 37737, 172006362, 80488, 174504898, 212258, 1272915272, 213766, 1280428583)

tb_data <- data.frame(country=country, year=year, key=key, value=value)
tb_data
```

When using `spread()` the `key` is the variable that you are spreading.
```{r}
tb_data %>% 
  spread(key=key, value=value)
```

## Practice
1. Run the following to build the `gene_exp` data frame.
```{r}
id <- c("gene1", "gene1", "gene2", "gene2", "gene3", "gene3")
type <- c("treatment", "control", "treatment", "control","treatment", "control")
L4_values <- rnorm(6, mean = 20, sd = 3)
```

```{r}
gene_exp <- 
  data.frame(gene_id=id, type=type, L4_values=L4_values)
gene_exp
```

2. Are these data tidy? Please use `spread()` to tidy the data.
```{r}
gene_exp %>% 
  spread(key=type, value=L4_values)
```

## summarize()
summarize() will produce summary statistics for a given variable in a data frame. For example, in homework 2 you were asked to calculate the mean of the sleep total column for large and small mammals. We did this using a combination of tidyverse and base R commands, which isn't very efficient or clean. It also took two steps.
```{r}
?msleep
```

From homework 2.
```{r}
large <- msleep %>% 
  select(name, genus, bodywt, sleep_total) %>% 
  filter(bodywt>=200) %>% 
  arrange(desc(bodywt))
```

```{r}
mean(large$sleep_total)
```

We can accomplish the same task using the `summarize()` function in the tidyverse and make things cleaner.
```{r}
msleep %>% 
  filter(bodywt>=200) %>%
  summarize(mean_sleep_lg=mean(sleep_total))
```

You can also combine functions to make useful summaries for multiple variables.
```{r}
msleep %>% 
    filter(bodywt>=200) %>% 
    summarize(mean_sleep_lg = mean(sleep_total), 
              min_sleep_lg = min(sleep_total),
              max_sleep_lg = max(sleep_total),
              total = n())
```

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of vector), first() (returns first value in vector), last() (returns last value in vector) and n_distinct() (number of distinct values in vector).

## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>%
  summarize(ngenera=n_distinct(genus))
```

2. What are the min, max, and mean body weight for all of the mammals? Be sure to include the total n.
```{r}
msleep %>%
  summarize(min_bodywt=min(bodywt),
            max_bodywt=max(bodywt),
            mean_bodywt=mean(bodywt),
            total=n())
```

##group_by()
The `summarize()` function is most useful when used in conjunction with `group_by()`. Although producing a summary of body weight for all of the mammals in the dataset is helpful, what if we were interested in body weight by feeding ecology?
```{r}
msleep %>%
  group_by(vore) %>% #we are grouping by feeding ecology
  summarize(min_bodywt=min(bodywt),
            max_bodywt=max(bodywt),
            mean_bodywt=mean(bodywt),
            total=n())
```

## Practice
1. Calculate mean brain weight by taxonomic order in the msleep data.
```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brainwt=mean(brainwt))
```

2. What does `NA` mean?

3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea?
```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brainwt=mean(brainwt, na.rm=TRUE))
```

```{r}
msleep %>% 
  filter(order=="Cetacea")
```

## That's it, let's take a break!   

--> On to [part 2](https://jmledford3115.github.io/datascibiol/lab4_2.html)  