---
title: "Data Visualization 2"
author: "Joel Ledford"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Review
Now that you have been introduced to `ggplot`, we need to practice a few more plot types and learn how to manipulate the aesthetics of plots to better suit our needs. Aesthetics can make a significant difference, but you can take it too far so remember that the goal is to produce clean plots that are not too distracting.  

##Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [R for Data Science](https://r4ds.had.co.nz/)
- [R Cookbook](http://www.cookbook-r.com/)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(skimr)
```

## Load the data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1
```{r message=FALSE, warning=FALSE}
homerange <- 
  readr::read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", na = c("", " ", "NA", "#N/A", "-999"))
```

Notice that by including common options for NA's, we improve the way data are read into R.
```{r}
glimpse(homerange)
```

## Box Plots
Boxplots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range. In the case below, we are plotting `log10.mass` by taxonomic class in the homerange data. `geom_boxplot()` is the geom type for a standard boxplot. The center line in each box represents the median, not the mean.
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10.mass))+
  geom_boxplot()
```

## Practice
1. There are more herbivores than carnivores in the homerange data, but how do their masses compare? Make a boxplot that compares their masses. Use `log10.mass`.
```{r}
homerange %>% 
  ggplot(aes(x=trophic.guild, y=log10.mass))+
  geom_boxplot()
```

2. Practice your dplyr skills and filter out carnivorous mammals. 
```{r}
carni_mammals <- 
  homerange %>% 
  filter(taxon=="mammals", trophic.guild=="carnivore")
carni_mammals
```

3. Now use a boxplot to visualize the range of body mass by family of carnivore.
```{r}
carni_mammals %>% 
  ggplot(aes(x=family, y=log10.mass))+
  geom_boxplot()+
  coord_flip()
```

## Aesthetics: Labels
Now that we have practiced scatterplots, barplots, and boxplots we need to learn how to adjust their appearance to suit our needs. Let's start with labelling x and y axes.  

In this example from part 1, we explored the relationship between body mass and homerange.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra)) +
  geom_point()
```

The plot looks clean, but it is incomplete. A reader unfamiliar with the data might have a difficult time interpreting the current labels. To add custom labels, we use the `labs` command.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra)) +
  geom_point()+
  labs(title = "Mass vs. Homerange",
       x = "Mass (log10)",
       y = "Homerange (log10)")
```

We can improve the plot further by adjusting the size and face of the text. We do this using `theme()`.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra)) +
  geom_point()+
  labs(title = "Mass vs. Homerange",
       x = "Mass (log10)",
       y = "Homerange (log10)")+
  theme(plot.title=element_text(size=18, face="bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` allows control of title position.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra)) +
  geom_point()+
  labs(title = "Mass vs. Homerange",
       x = "Mass (log10)",
       y = "Homerange (log10)")+ 
  theme(plot.title = element_text(size = rel(2), hjust = 0.5))
```

## Practice
1. Make a barplot that shows the number of individuals per locomotion type. Be sure to provide a title and label the axes appropriately.
```{r}
homerange %>% 
  ggplot(aes(x=locomotion))+
  geom_bar()+
  labs(title = "Locomotion Type",
       x = "Locomotion Type",
       y = "# Individuals")+ 
  theme(plot.title = element_text(size = rel(1.5)))
```

## Other Aesthetics
There are lots of options for aesthtics. An aesthetic can be assigned to either numeric or categorical data. `color` is a common option; notice that an appropriate key is displayed when you use one of these options.
```{r}
homerange %>% 
  ggplot(aes(x=log10.mass, y=log10.hra, color=locomotion))+
  geom_point()
```

`size` adjusts the size of points relative to a continuous variable.
```{r}
options(scipen = 999) #disable scientific notation

homerange %>% 
  ggplot(aes(x=log10.mass, y=log10.hra, size=mean.mass.g))+
  geom_point()
```

Here I am plotting `class` on the x-axis and `log10.mass` on the y-axis. I use `group` to make individual box plots for each taxon. I also use `fill` so I can associate the different taxa with a color coded key.
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10.mass, group=taxon, fill=taxon))+
  geom_boxplot()
```

## Practice
1. Make a barplot that shows counts of ectotherms and endotherms. Label the axes, provide a title, and fill by thermoregulation type.
```{r}
homerange %>% 
  ggplot(aes(x=thermoregulation, fill=thermoregulation))+
  geom_bar()+
  labs(title = "Thermoregulation Counts",
       x = "Thermoregulation Type",
       y = "# Individuals")+ 
  theme(plot.title = element_text(size = rel(1.25)))
```

2. Make a boxplot that compares thermoregulation type by log10.mass. Group and fill by class. Label the axes and provide a title.
```{r}
homerange %>% 
  ggplot(aes(x=thermoregulation, y=log10.mass, group=class, fill=class))+
  geom_boxplot()+
  labs(title = "Thermoregulation vs. Log10 Mass by Taxonomic Class",
       x = "Thermoregulation Type",
       y = "log10.mass")+ 
  theme(plot.title = element_text(size = rel(1.25)))
```

## Wrap-up
Please review the learning goals and be sure to use the code here as a reference when completing the homework.

-->[Home](https://jmledford3115.github.io/datascibiol/)