---
title: "Labs 1 and 2 Review"
author: "Joel Ledford"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

## Progress
Everyone is doing great so far. Remember, learning R is like learning a new language so it is hard at first but it will get easier. I will continue to make adjustments to make the labs flow. Please don't be discouraged, if you need help reach out to me!

## Git and GitHub
These programs are glitchy and sensitive. Remember, first commit then push and always provide a name. Although they are frustrating, both of these programs are the primary way that code is shared online. Please make an appointment with me so I can help you. I will be at our classroom each Friday at 12:00 and will stay after class to answer questions and provide help.

## Notes
You should build your own notes as we continue through the quarter, copying bits of my code and adding your own. Use RMarkdown to take notes so you can practice. I would take some time after each week to review the lesson, refine your notes, and make sure that you understand the main points. Post the notes to your GitHub repository!

## Recap
If git and GitHub are not working for you, [email me](mailto:jmledford@ucdavis.edu).  

1. **Arithmetic.**R can be used as a fancy calculator of sorts. Be sure to put expressions in code chunks. The mac shortcut is option + command + i. The PC shortcut is Ctrl + Alt + i.
```{r}
9*5
```

2. **Objects.** We use objects in R to store information of multiple types. Make sure that you give your objects reasonable names that reflect the information they contain. Once an object is present, they are easy to work with.
```{r}
my_age <- 45
```

```{r}
my_age + 1
```

3. **Data Classes.** R can work with data of many types of classes. It is sometimes important to know which class of data you are working with.
```{r}
integer <- 5L
numeric <- 88.9
character <- "Moose"
```

```{r}
class(integer)
class(numeric)
class(character)
```

It is also easy to make a logical test of class type.
```{r}
is.character(numeric) # does the object character belong to class numeric?
```

4. **Data Structures.** R organizes data into a few different structures. So far, we have worked with vectors and data matrices.  Vectors are important because they are fundamental to R and built using the `c` command. Vectors contain data of the same class.
```{r}
happy_vector <- c(5, 10, 15, 20, 25 ,30)
```

```{r}
unhappy_vector <- c(5, 10, 15, "green", "blue")
unhappy_vector
```

Why does `unhappy_vector` work? R changes the content of the vector into what it thinks is the best class. If you use this in an analysis, it will often result in problems so it is best to keep vectors to a single class of data.
```{r}
class(unhappy_vector)
```

Vectors may contain data of many different class types.
```{r}
options(scipen=999) #disables scientific notation
organism<-c("Human","Mouse","Fruit Fly", "Roundworm","Yeast")
genomeSizeBP<-c(3000000000,3000000000,135600000,97000000,12100000)
estGeneCount<-c(30000,30000,13061,19099,6034)
```

We can combine them into a single vector.
```{r}
genome_size <- c(genomeSizeBP, estGeneCount)
genome_size
```

And provide names.
```{r}
names(genome_size) <- c("genome size BP", "genome size BP", "genome size BP", "genome size BP", "genome size BP", "gene count", "gene count", "gene count", "gene count", "gene count")
genome_size
```

You can also select elements of a vector using `[]`
```{r}
humans <- genome_size[c(1, 6)]
humans
```

5. **Data Matrices.** Vectors may seem inefficient, but they are a fundamental data structure in R. A data matrix is a little better at organizing the data.
```{r}
genome_matrix <- matrix(genome_size, nrow=2, byrow=T)
genome_matrix
```

Let's name the columns and rows.
```{r}
colnames(genome_matrix) <- organism
rownames(genome_matrix) <- c("genome size BP", "gene count")
genome_matrix
```

Once we have the matrix, we can use functions to make some calculations.
```{r}
means <- rowMeans(genome_matrix)
```

Add them to the matrix
```{r}
genome_matrix2 <- cbind(genome_matrix, means)
genome_matrix2
```

6. **Data Frames.**Most data that you will encounter will be structured as a data matrix. Notice that each column is a vector!
```{r}
genome_frame <- data.frame(organism=organism, genomeSizeBP=genomeSizeBP, estGeneCount=estGeneCount)
genome_frame
```

## Last Hints
If you understand the code here, then you are ready for the next step. If you are stuck, please work through the lesson again on the class webpage. If you need help, please [email me](mailto:jmledford@ucdavis.edu) and I will arrange time.

