---
title: "Homework 5"
author: "Joel Ledford"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Develop your dplyr superpowers so you can easily and confidently manipulate dataframes.  
2. Learn helpful new functions that are part of the `janitor` package.  

## Instructions
For the second part of lab today, we are going to spend time practicing the dplyr functions we have learned and add a few new ones. This lab doubles as your homework. Please complete the lab and push your final code to GitHub.  

## Load the libraries
```{r}
library("tidyverse")
library("janitor")
```

## Load the superhero data
These are data taken from comic books and assembled by fans. The include a good mix of categorical and continuous data.  Data taken from: https://www.kaggle.com/claudiodavi/superhero-set  

Check out the way I am loading these data. If I know there are NAs, I can take care of them at the beginning. But, we should do this very cautiously. At times it is better to keep the original columns and data intact.  
```{r}
superhero_info <- read_csv("data/heroes_information.csv", na = c("", "-99", "-"))
superhero_powers <- read_csv("data/super_hero_powers.csv", na = c("", "-99", "-"))
```

## Data tidy
1. Some of the names used in the `superhero_info` data are problematic so you should rename them here. Before you do anything, first have a look at the names of the variables. You can use `rename()` or `clean_names()`.    
```{r}
names(superhero_info)
```

```{r}
names(superhero_powers)
```

```{r}
superhero_powers <- clean_names(superhero_powers)
```

```{r}
superhero_info <- clean_names(superhero_info)
```

## `tabyl`
The `janitor` package has many awesome functions that we will explore. Here is its version of `table` which not only produces counts but also percentages. Very handy! Let's use it to explore the proportion of good guys and bad guys in the `superhero_info` data.  
```{r}
tabyl(superhero_info, alignment)
```

1. Who are the publishers of the superheros? Show the proportion of superheros from each publisher. Which publisher has the highest number of superheros?  
```{r}
tabyl(superhero_info, publisher)
```

2. Notice that we have some neutral superheros! Who are they? List their names below.  
```{r}
neutral <- superhero_info %>% filter(alignment=="neutral")
```

```{r}
neutral$name
```

## `superhero_info`
3. Let's say we are only interested in the variables name, alignment, and "race". How would you isolate these variables from `superhero_info`?
```{r}
superhero_info %>% select("name", "alignment", "race")
```

## Not Human
4. List all of the superheros that are not human.
```{r}
not_human <- superhero_info %>% filter(race!="Human")
```

```{r}
not_human$name
```

## Good and Evil
5. Let's make two different data frames, one focused on the "good guys" and another focused on the "bad guys".
```{r}
good_guys <- 
  superhero_info %>% 
  filter(alignment=="good")
```

```{r}
bad_guys <- 
  superhero_info %>% 
  filter(alignment=="bad")
```

6. For the good guys, use the `tabyl` function to summarize their "race".
```{r}
good_guys %>% tabyl(race)
```

7. Among the good guys, Who are the Vampires?
```{r}
good_guys %>% filter(race=="Vampire")
```

8. Among the bad guys, who are the male humans over 200 inches in height?
```{r}
bad_guys %>% filter(race=="Human" & gender=="Male" & height>=150)
```

9. Are there more good guys or bad guys with green hair?  
```{r}
good_guys %>% 
  filter(hair_color=="Green")
```

```{r}
bad_guys %>% 
  filter(hair_color=="Green")
```

```{r}
table(superhero_info$hair_color, superhero_info$alignment)
```

```{r}
superhero_info %>% 
  tabyl(hair_color, alignment)
```

```{r}
superhero_info %>% 
  filter(hair_color=="Green") %>% 
  group_by(alignment) %>% 
  summarise(n=n())
```

```{r}
superhero_info %>% 
  filter(hair_color=="Green") %>% 
  count(hair_color, alignment)
```

10. Let's explore who the really small superheros are. In the `superhero_info` data, which have a weight less than 50? Be sure to sort your results by weight lowest to highest.  
```{r}
superhero_info %>% 
  filter(weight<50) %>% 
  arrange(weight)
```

11. Let's make a new variable that is the ratio of height to weight. Call this variable `height_weight_ratio`.    
```{r}
superhero_info <- 
  superhero_info %>% 
  mutate(height_weight_ratio=height/weight)
```

12. Who has the highest height to weight ratio?  
```{r}
superhero_info %>% 
  arrange(desc(height_weight_ratio))
```

## `superhero_powers`
Have a quick look at the `superhero_powers` data frame.  

13. How many superheros have a combination of agility, stealth, super_strength, stamina?
```{r}
superhero_powers %>%
  filter(agility & stealth & super_strength & stamina) %>%
  select(hero_names, agility, stealth, super_strength, stamina)
```

```{r}
superhero_powers %>% 
  select(hero_names, agility, stealth, super_strength, stamina) %>% 
  filter(agility==TRUE & stealth==TRUE & super_strength==TRUE & stamina==TRUE)
```

## Your Favorite
14. Pick your favorite superhero and let's see their powers!  
```{r}
superhero_powers %>%
  filter(hero_names == "Darth Vader") %>%
  select_if(all)  # Selects all columns where all values are TRUE
```

15. Can you find your hero in the superhero_info data? Show their info!  
```{r}
superhero_info %>% 
  filter(name=="Darth Vader")
```

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.  