---
title: "GOBY Explore"
author: "Gabe"
date: "`r Sys.Date()`"
output: 
  html_document:
    keepmd: yes
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Libraries**
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
```

**Loading Data**
```{r}
data <- read_csv("Potential Data/Final Data .csv")
data <- data%>%
  na_if("-999")
```
**Separating Dates**
```{r}
data <-data %>%
  separate(Date, into = c("day", "month", "year"), sep="/")
```

**Refining Data**
```{r}
value_data <- data%>%
  select("day", "month", "year", "SiteName", ammNO3f_Val, ammSO4f_Val, ECf_Val, OMCf_Val, SOILf_Val, SVR_Val)
```

**Mean Data per Year**
```{r}
mean_data_by_year <- value_data%>%
  mutate(SiteName=paste(SiteName, year, sep="_"))%>%
  group_by(SiteName)%>%
  summarise(mean_ammNO3=mean(ammNO3f_Val, na.rm=T),
            mean_ammSO4=mean(ammSO4f_Val, na.rm=T),
            mean_EC=mean(ECf_Val, na.rm=T),
            mean_OMC=mean(OMCf_Val, na.rm=T),
            mean_SOIL=mean(SOILf_Val, na.rm=T),
            mean_SVR=mean(SVR_Val, na.rm=T))%>%
  as.data.frame()%>%
  separate(SiteName, into = c("SiteName", "year"), sep="_")
```

**Plots showing the mean Standard Visual Range (km) for each year**
```{r}
mean_data_by_year%>%
  ggplot(aes(year, mean_SVR))+
  geom_point()+
  facet_wrap(~SiteName)+
  theme(axis.text.x = element_text(angle=90, hjust = 1))
```

Some things to note:

- Bliss SP is separated into TRPA for years <2021 and TRPA/RHTS for years =>2021. 

- Death Valley and Wrightwood only have data from years <= 2013. 

- Hoover, CA has the best average visibility per year, while Fresno has the worst. 

- Trinity, CA has no data from 2016. 

- Lava Beds NM has a steady decline in SVR from 2019-2021 (See 2021 Antelope Fire)


