---
title: "Data Exploration Oonagh"
author: "Oonagh Pretorius"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(naniar)
library(devtools)
library(tidyverse)
library(here)
library(janitor)
library(ggmap)
library(ggplot2)
library(lubridate)
library(dplyr)
```

Loading data

```{r}
visibility <- read_csv("data/CA_visibility_data.csv", na = "-999") %>% clean_names()
```
Split date by year:
```{r}
visibility_tidy <- visibility %>% 
  separate(date, into= c("month", "day", "year"), sep = "/")
```

#### **Data Cleaning**

```{r}
glimpse(visibility_tidy)
```

```{r}
head(visibility_tidy) # the data are tidy
```

Any NAs?
```{r}
naniar::miss_var_summary(visibility_tidy)
summary(visibility_tidy)
```

Importantly, this shows us that there are no missing values for "site_code", "site_name", "date", "longitude", "latitude", "elevation", and "state". The values that do have NAs are all below 10%.


#### **Exploration**

```{r}
names(visibility_tidy)
```
##############################################################################
##############################################################################
##### 1. How many sites and where are they? Check "site_code" vs. "site_name"
######**Don't need to include this section in final code doc**
*There are 23 distinct sites names which we will use for analysis*


```{r}
visibility_tidy %>%
  summarize(distinct_site_code = n_distinct(site_code), #interestingly, there is one more site_code (24) than site_name (23).
            distinct_site_name = n_distinct(site_name),
            distinct_state = n_distinct(state)) #this confirms there is only data for CA in the dataset
```
OR
```{r}
visibility_tidy %>%
  summarize(across(c(site_code, site_name, state), n_distinct))
```

```{r}
visibility_tidy %>%
    tabyl(site_code, site_name)
```

```{r}
visibility_tidy %>% 
  tabyl(site_code, site_name) %>% 
  adorn_percentages() %>%
  adorn_pct_formatting(digits = 2) #cannot find the extra site_code. As all site_name variables contain 100% of their site codes, I think we should use the site_name variable going forward.
```
##############################################################################
##############################################################################


##### 2. where are the sites located in CA?
####### Step 1: Create a Base Map
a) Find the range of latitude and longitude for our data:
```{r}
visibility_tidy %>% 
  select(latitude, longitude) %>% #note that the lat and long are in degrees decimal
  summary()
```

b) Set the bounding box range:
```{r}
lat <- c(33.46, 41.71)
long <- c(-124.1, -116.4)
bbox <- make_bbox(long, lat, f = 0.05)
```

Let's get a base map for our bounding box area. We will use the stamen maps because they are free. There are several different map types, including: `terrain-labels`, `terrain-lines`, `toner`, `toner-2011`, `toner-background`, `toner-hybrid`, `toner-lines`, `toner-lite`, and `watercolor.`
```{r}
map1 <- get_map(bbox, maptype = "terrain", source = "stamen")
ggmap(map1)
```

####### Step 2: Adding Points to Base Map
`ggmap` works well with `ggplot2`. To add our points we only need to specify the x and y location similar to how we made charts in previous labs. 

###### **I. Basic map with site locations**

```{r}
ggmap(map1) + 
  geom_point(data = visibility_tidy, aes(longitude, latitude), size=1.5) +
  labs(x= "Longitude", y= "Latitude", title="Air Quality Research Site Locations in CA")

  #ggsave("plot.png", width = 5, height = 5)
```

###### **II.Basic map with site location and by elevation**

```{r}
ggmap(map1) + 
  geom_point(data = visibility_tidy, aes(longitude, latitude, colour=elevation), size=2.5) +
  labs(x= "Longitude", y= "Latitude", title="Air Quality Research Site Locations in CA")
```

###### **III.Basic map with site location and colour name key** - not informative - too many sites
 
```{r}
ggmap(map1) + 
  geom_point(data = visibility_tidy, aes(longitude, latitude, colour=site_name), size=2.5) +
  labs(x= "Longitude", y= "Latitude", title="Air Quality Research Site Locations in CA")
```

###### **IV. Basic map with only ONE site location**
Example:
```{r}
visibility_yos <- visibility_tidy %>% 
  filter(site_name=="Yosemite NP")

  ggmap(map1) + 
  geom_point(data = visibility_yos, aes(longitude, latitude), size=2.5) +
  labs(x= "Longitude", y= "Latitude", title="Yosemite Air Quality Research Site")
```

###### **IV. Basic map with only ONE site location with key**
  Example:
```{r}
visibility_yos <- visibility_tidy %>% 
  filter(site_name=="Yosemite NP")

  ggmap(map1) + 
  geom_point(data = visibility_yos, aes(longitude, latitude, fill=site_name), col = "orange", size=2.5) +
  labs(x= "Longitude", y= "Latitude", title="Yosemite Air Quality Research Site")
```

**Saving maps**
```{r}
ggsave("plot.png", width = 5, height = 5) #Saves last plot as 5’ x 5’ file named "plot.png" in working directory.
#Matches file type to file extension.
```

#######################################################################################
#######################################################################################

##### 3. Comparing variables over time by site

First, summarise all variables by site and year:
```{r}
visibility2 <- visibility_tidy %>% 
  group_by(site_name, year) %>% 
  summarise(across(c(svr_val, amm_no3f_val, amm_so4f_val, e_cf_val, om_cf_val, soi_lf_val, longitude, latitude), mean, na.rm=T), total=n())
```
Now, we make a separate, faceted graph per variable:

**A) Graph Mean Standard Visual Range over time, faceted by site**
```{r}
visibility2 %>% 
  ggplot(aes(x=year, y=svr_val))+
  geom_point(na.rm=TRUE)+
  facet_wrap(~site_name, ncol=4)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Standard Visual Range over Time",
       x = "Year",
       y = "Average Standard Visual Range (km)",)

#ggsave("svr_time.png", width = 10, height = 10)
```
**B)Graph Mean Ammonium Nitrate Concentration Over Time** faceted by site
```{r}
visibility2 %>% 
  ggplot(aes(x=year, y=amm_no3f_val))+
  geom_point(na.rm=TRUE)+
  facet_wrap(~site_name, ncol=4)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Ammonium Nitrate Concentration over Time",
       x = "Year",
       y = "Ammonium Nitrate Concentration (ug/m^3)",)

#ggsave("amno3_time.png", width = 10, height = 10)
```
**C) Graph Mean Ammonium Sulfate Concentration Over Time** faceted by site
```{r}
visibility2 %>% 
  ggplot(aes(x=year, y=amm_so4f_val))+
  geom_point(na.rm=TRUE)+
  facet_wrap(~site_name, ncol=4)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Ammonium Sulfate Concentration over Time",
       x = "Year",
       y = "Ammonium Sulfate Concentration (ug/m^3)",) 

#ggsave("ammso4_time.png", width = 10, height = 10)
```

**D) Graph Mean Elemental Carbon Concentration Over Time** faceted by site
```{r}
visibility2 %>% 
  ggplot(aes(x=year, y=e_cf_val))+
  geom_point(na.rm=TRUE)+
  facet_wrap(~site_name, ncol=4)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Elemental Carbon Concentration over Time",
       x = "Year",
       y = "Elemental Carbon Concentration (ug/m^3)",)

#ggsave("ecf_time.png", width = 10, height = 10)
```
####*Notes: Interesting fluctuations for Lassen, Lava Beds, Point Reyes, Yosemite, Sequoia, Trinity. Link to wildfire data?*

**E) Graph Mean Organic Mass Carbon (fine) Over Time** faceted by site
```{r}
visibility2 %>% 
  ggplot(aes(x=year, y=om_cf_val))+
  geom_point(na.rm=TRUE)+
  facet_wrap(~site_name, ncol=4)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Organic Mass Carbon (fine) over Time",
       x = "Year",
       y = "Organic Mass Carbon (fine) Concentration (ug/m^3)",) #fill by season?

#ggsave("omcf_time.png", width = 10, height = 10)
```

**F) Graph Mean Fine Soil Particles Over Time** faceted by site
```{r}
visibility2 %>% 
  ggplot(aes(x=year, y=soi_lf_val))+
  geom_point(na.rm=TRUE)+
  facet_wrap(~site_name, ncol=4)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Fine Soil Particles over Time",
       x = "Year",
       y = "Fine Soil Particle Concentration (ug/m^3)",)
#ggsave("soilf_time.png", width = 10, height = 10)
```
##### 4. Plotting the Relationship between Average Standard Visual Range and Elevation

First, summarise SVR by site:
```{r}
visibility3 <- visibility_tidy %>% 
  group_by(site_name) %>% 
  summarise(across(c(svr_val, elevation), mean, na.rm=T), total=n()) %>% 
   arrange(desc(svr_val))
visibility3
```

Notes: Hoover has the highest overall SVR and Fresno has the lowest.

Now, we can graph the relationship:
```{r}
visibility3 %>% 
  ggplot(aes(svr_val, elevation, col=site_name))+
  geom_point()+
  labs(title = "Elevation vs. Average Standard Visual Range",
       x = "Average Standard Visual Range (km)",
       y = "Elevation (ft)",
       fill = "Site Name")

#ggsave("svr_elevation.png", width = 8, height = 5)
```

#### **Potential questions**

1.  How does visibility change with elevation?
2.  How does visibility change over time (overall and by site)?
3.  Can changes in visibility be linked with major pollution events such as wild fires?
4.  What are the most/least polluted sites?
