---
title: "Lab 9: Making slides and Outputs"
author: "Joel Ledford"
date: "Winter 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("jpeg")
```

## Repositories
- Take 5 minutes or so to double check your repository.  
- Is it neat, clean, and organized?  
- Did you add a README?  
- Did you write a description?  

## Presentations
- Showcase what you have learned.   
- Make sure there is a separate, complete repository.  
- 5-6 minute presentation.  
- Clear, logical, and with figures.   

## ioslides
- A simple way to make a presentation in R.   
- Allows you to embed code and figures.
- ioslides resource: https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html

## Adjusting output
- `include = FALSE` prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
- `echo = FALSE` prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.
- `message = FALSE` prevents messages that are generated by code from appearing in the finished file.
- `warning = FALSE` prevents warnings that are generated by code from appearing in the finished.

```{r include=FALSE}
# notice that this hides the code and messages but loads the data
life_history <- readr::read_csv(file = "data/mammal_lifehistories_v2.csv", 
                  na = c("NA", " ", ".", "-999"))
```

## Code example
We used the following libraries.
```{r echo=TRUE, message=FALSE}
library("tidyverse")
library("naniar")
```

## How many NA's?
```{r echo=TRUE}
naniar::miss_var_summary(life_history)
```

## Plot example | Gestation Time in Large Mammals
```{r warning=FALSE}
life_history %>% 
  filter(mass>=100000) %>% 
  ggplot(aes(x=order, y=gestation))+
  geom_boxplot()
```

## Exporting figures
```{r echo=T, eval=F, warning=FALSE}
plot <- life_history %>% 
  filter(mass>=100000) %>% 
  ggplot(aes(x=order, y=gestation))+
  geom_boxplot()

png("gestation.png")
plot
dev.off()
```