---
title: "Lab 11 Intro"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("jpeg")
```

## Seating & Set-up
1. Please make sure that you sit next to your group members for the rest of the quarter.
2. Please set-up your computer as normal.

## Warm-up
For this warm-up, please use the `heartrate` data. Be sure to label your plots!
```{r}
library("tidyverse")
```

```{r}
heartrate <- read_csv("data/heartrate.csv")
```

```{r}
heartrate
```

1. Make a plot that compares the heartrate of patients for drug a.
```{r}
heartrate %>% 
  ggplot(aes(x=patient, y=a, fill=patient))+
  geom_col()+
  labs(title="Comparison of drug A",
       x=NULL,
       y="Heartrate")
```

2. Make a plot that compares heartrate (as a range) for each drug.
```{r}
heartrate %>% 
  pivot_longer(-patient,
               names_to="drug",
               values_to = "heartrate") %>% 
  ggplot(aes(x=drug, y=heartrate, fill=drug))+
  geom_boxplot(alpha=0.7)+
  labs(title="Comparison of Heartrate Drugs",
       x="Drugs",
       y="Heartrate")
```

## Group project  
We will meet in our groups today to compare the data sets that we found. You should try to pick the data set that you think will be most appropriate; i.e. does it have a mix of variables, does it have a story to tell? Once you have a data set, think about the kinds of questions you might ask. Remember, in the end you will need to deliver a 5-6 minute presentation of your work. If you need help, please ask!  