---
title: "BIS15L_AI_hmwk"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
This homework is a bit less traditional this week. There are some errors in the code below. In addition add notes and titles to the code chunks that explains what the code chunck does. Use AI tools, notes from class, or any other resources to help fix the errors. Feel free to edit any plots to your preference. Don't forget to answer the two questions at the bottom. Upon completion, you should not have to '#' out any of the code chunks.

Push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.   

Again make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  


```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```


1.
```{r}
blood_count <- read_csv("data/blood_count.csv")
```

2. 
```{r}
glmpse(blood_count)

```

3. 
```{r}
blood_count %>% 
  group_by(gender) %>% 
  summarize(min=min(red_blood_cells)
            mean=maen(red_blood_cells),
            max=max(red_blood_cells)
```


4. 
```{r}
options(scipen = 999, digits = 2)

blood_count %>% 
  filter(Gender = "male") %>%
  select[3:9] %>%
  colMeans(na.rm = TRUE)

```

5. 
```{r}
bloood_count %>% 
  filter(Gender = "Female") />%
  select[3:9] %>%
  colMeans(na.rm = T)
```


6. 
```{r}

blood_count %>%
  pivot_wider(cols = [3:9], names_to = "Test", values_to = "Values") %>%
  ggplot(aes(x = Gender, y = Values, fill = Gender)) +
  geom_boxplot() +
  facet(~ Test, scales = "free") 
  labs(title = "Boxplot of Blood Count Tests",
       x = "Test",
       y = "Values")

```

7. 
```{r}
grouped_bc <- blood_count %>%
  change(Age_Group = case_when(
    Age >= 20 & Age <= 29 ~ "20s",
    Age >= 30 & Age <= 39 ~ "30s",
    Age >= 40 & Age <= 49 ~ "40s",
    Age >= 50 & Age <= 59 ~ "50s",
    Age >= 60 & Age <= 70 ~ "60s")) # there were only 4 people in 70s, so I combined it with 60s

head(grouped_bc)

```

8. 
```{r}
blood_count %>%
  pivot_longer(cols = 3:9, names_to = "Test", values_to = "Values") %>%
  ggplot(ase(x = Age_Group, y = values, fill = Age_Group)) +
  geom_boxplot() +
  facet_wrap( Test, scales = "free") +  # Facet by the "Test" variable
  labs(title = "Boxplot of Blood Count Tests by Age Group",
       x = "Age Group",
       y = "Values")

```
9. What are two drawbacks and two benefits of AI?

10. Do you think you will use AI for coding in the future? Why or why not?


## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 