---
title: "Lab 15.2"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*  
1. Use `shinydashboard` to improve ui aesthetics.  

## Load the Libraries  
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(shiny)
library(shinydashboard)
library(janitor)
```

## Load the Data
For this lab, we will use data on the gestation length of eutherian mammals. The data are from: Thodoris Danis, Antonis Rokas; The evolution of gestation length in eutherian mammals. Proc. Biol. Sci. 1 October 2024; 291 (2033): 20241412. https://doi.org/10.1098/rspb.2024.1412
```{r message=FALSE, warning=FALSE}
gestation <- read_csv("data/Data_S1.csv") %>% 
  clean_names() %>% 
  rename(gestationlength=gestationength)
```

```{r}
names(gestation)
```

## Aesthetics
There are many ways to adjust shiny apps for better visual appeal. Let's start with some more practice and then add details.  

1. Build an app that allows users to select a mammalian `order` and a variable of interest (body mass, longevity, or gestation length) and then plots the density distribution of that variable for the selected order.
```{r, eval=F, echo=T}

```

## ui Layout with `shinydashboard`
We have a functional app, but the layout doesn't look very nice. We can change that with many different methods (and specialized libraries), but shiny dashboards are simple to use and offer a lot of functionality for making apps look professional quickly.  

We won't use `fluidPage()` with `shinydashboard` but rather `dashboardPage()`. A shiny dashboard contains a minimum of a header, a sidebar, and a body.
```{r, eval=F, echo=T}
ui <- 
  dashboardPage(
    
  
  dashboardHeader(title = "This is the Header"),
 
  dashboardSidebar(title = "This is the Sidebar"),
  
  dashboardBody() # this is where the output will go
)

server <- function(input, output) { }

shinyApp(ui, server)
```

Let's add our ui and server elements to make a functional dashboard. We will put our ui elements in the `dashboardBody()`. For now, we will turn off the sidebar.
```{r, eval=F, echo=T}

```

Now let's add our server functions.
```{r, eval=F, echo=T}

```

What if you want to turn on the sidebar and put the inputs there? We can do that by moving the `selectInput()` functions to the `dashboardSidebar()`.
```{r, eval=F, echo=T}

```

Whew! We made it! We created a nice looking app with `shinydashboard`. And now we can easily see how even a basic shiny app can become messy and why it's always helpful to include lots of annotations. chatGPT can really help troubleshoot errors and is especially helpful with annotation.    

## Practice
Let's try to build an app that allows users to explore sex composition of wolf packs by `pop`. These data are from: Brandell, Ellen E (2021), Serological dataset and R code for: Patterns and processes of pathogen exposure in gray wolves across North America, Dryad, [Dataset](https://doi.org/10.5061/dryad.5hqbzkh51)  

1. Start by loading and exploring the data.
```{r message=FALSE, warning=FALSE}
wolves <- read_csv("data/wolves_data/wolves_dataset.csv") %>% clean_names()
```

```{r}
names(wolves)
```

2. For the app, here is a faceted version of what we are looking for. We want the reactive part to be `pop` and you should try to use `shinydashboard`.
```{r}
wolves %>% 
  filter(sex!="NA") %>% 
  ggplot(aes(x=sex, fill=sex))+
  geom_bar()+
  facet_wrap(~pop)
```

```{r, eval=F, echo=T}

```

-->[Home](https://jmledford3115.github.io/datascibiol/)