---
title: "Homework 11"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
``` 

```{r}
options(scipen = 999)
```

## Data
For this homework, we are going to use the [Gapminder](https://www.gapminder.org/data/) data on population (pop), GDP per capita (gdp), and life expectancy (lex). These data were downloaded and compiled using the `gapminder_compile.Rmd` script in the data folder.  

Let's focus our attention on data from the past 100 years, between 1925 and 2025.
```{r message=FALSE, warning=FALSE}
gapminder <- read_csv("data/gapminder.csv") %>% 
  filter(between(year, 1925, 2025))
```

1. Use the function(s) of your choice to get an idea of the overall structure of the data, including its dimensions, column names, classes, etc. 
```{r}
dim(gapminder)
names(gapminder)
glimpse(gapminder)
```

2. How many countries are represented in this dataset?
```{r}
gapminder %>% 
  summarize(n_countries = n_distinct(country))
```

3. Which five countries have had the largest absolute population growth between 1925-2025? Show this as a table.
```{r}
pop_growth_top5 <- gapminder %>% 
  filter(year %in% c(1925, 2025)) %>% 
  select(country, year, pop) %>% 
  pivot_wider(names_from = year, values_from = pop, names_prefix = "year_") %>% 
  mutate(pop_growth = year_2025 - year_1925,
         abs_pop_growth = abs(pop_growth)) %>% 
  arrange(desc(abs_pop_growth)) %>% 
  slice_head(n = 5)

pop_growth_top5
```

4. Make a plot that shows population growth over time for the top country you found in question #3.
```{r}
top_country <- pop_growth_top5 %>% 
  slice_head(n = 1) %>% 
  pull(country)

gapminder %>% 
  filter(country == top_country) %>% 
  ggplot(aes(year, pop)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = paste("Population Growth:", top_country),
    x = "Year",
    y = "Population"
  ) +
  theme_minimal()
```

5. Has global life expectancy changed between 1925 and 2025? Show the min, median, mean, and max for all countries (combined) in the dataset.
```{r}
gapminder %>% 
  filter(year %in% c(1925, 2025)) %>% 
  group_by(year) %>% 
  summarize(
    min_lex = min(lex, na.rm = TRUE),
    median_lex = median(lex, na.rm = TRUE),
    mean_lex = mean(lex, na.rm = TRUE),
    max_lex = max(lex, na.rm = TRUE),
    .groups = "drop"
  )
```

6. Make a plot that shows the distribution of life expectancy for all countries in 1925 and 2025 (try using `geom_density`). What do you notice about the shift in life expectancy over time?
```{r}
gapminder %>% 
  filter(year %in% c(1925, 2025)) %>% 
  ggplot(aes(lex, fill = factor(year))) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Life Expectancy Distribution in 1925 vs 2025",
    x = "Life Expectancy",
    y = "Density",
    fill = "Year"
  ) +
  theme_minimal()
```

7. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer in 2025? Show this as a plot. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gapminder %>% 
  filter(year == 2025, country != "Monaco") %>% 
  ggplot(aes(gdp, lex)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "GDP per Capita vs Life Expectancy (2025)",
    x = "GDP per Capita",
    y = "Life Expectancy"
  ) +
  theme_minimal()
```

8. Which five countries have had the highest absolute GDP per capita growth over the past 50 years? Show this as a table. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gdp_growth_top5 <- gapminder %>% 
  filter(year %in% c(1975, 2025), country != "Monaco") %>% 
  select(country, year, gdp) %>% 
  pivot_wider(names_from = year, values_from = gdp, names_prefix = "year_") %>% 
  mutate(gdp_growth = year_2025 - year_1975,
         abs_gdp_growth = abs(gdp_growth)) %>% 
  arrange(desc(abs_gdp_growth)) %>% 
  slice_head(n = 5)

gdp_growth_top5
```

9. How does per capita GDP growth compare between these same five countries over the past 50 years? Show this as a plot.
```{r}
top5_gdp_countries <- gdp_growth_top5 %>% 
  pull(country)

gapminder %>% 
  filter(year >= 1975, country %in% top5_gdp_countries) %>% 
  ggplot(aes(year, gdp, color = country)) +
  geom_line(linewidth = 1) +
  labs(
    title = "GDP per Capita Trends (Top 5 Absolute Growth Countries)",
    x = "Year",
    y = "GDP per Capita",
    color = "Country"
  ) +
  theme_minimal()
```

10. Do one analysis of your choice that includes a table and plot as outputs.
```{r}
lex_growth_table <- gapminder %>% 
  filter(year %in% c(1975, 2025)) %>% 
  select(country, year, lex) %>% 
  pivot_wider(names_from = year, values_from = lex, names_prefix = "year_") %>% 
  mutate(lex_growth = year_2025 - year_1975) %>% 
  arrange(desc(lex_growth)) %>% 
  slice_head(n = 10)

lex_growth_table
```

```{r}
lex_growth_table %>% 
  ggplot(aes(reorder(country, lex_growth), lex_growth)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Life Expectancy Gain (1975-2025)",
    x = "Country",
    y = "Life Expectancy Gain (years)"
  ) +
  theme_minimal()
```

## Additional Comprehensive EDA
The analyses below add common EDA checks that are usually done before modeling or interpretation.

11. Data quality checks: missingness, duplicate country-year rows, and year coverage by country.
```{r}
quality_summary <- gapminder %>% 
  summarize(
    n_rows = n(),
    n_columns = ncol(gapminder),
    n_countries = n_distinct(country),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  )

missing_summary <- gapminder %>% 
  summarize(across(everything(), ~sum(is.na(.)))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_n"
  ) %>% 
  mutate(missing_pct = 100 * missing_n / nrow(gapminder)) %>% 
  arrange(desc(missing_n))

duplicate_country_year <- gapminder %>% 
  count(country, year) %>% 
  filter(n > 1)

country_year_coverage <- gapminder %>% 
  group_by(country) %>% 
  summarize(n_years = n_distinct(year), .groups = "drop")

quality_summary
missing_summary
duplicate_country_year
country_year_coverage %>% summarize(min_years = min(n_years), max_years = max(n_years))
```

```{r}
missing_summary %>% 
  ggplot(aes(reorder(variable, missing_n), missing_n)) +
  geom_col(fill = "gray50") +
  coord_flip() +
  labs(
    title = "Missing Values by Variable",
    x = "Variable",
    y = "Number of Missing Values"
  ) +
  theme_minimal()
```

12. Numeric distribution summary and potential outliers in 2025 using the IQR rule.
```{r}
numeric_summary <- gapminder %>% 
  summarize(
    across(
      c(pop, gdp, lex),
      list(
        min = ~min(., na.rm = TRUE),
        q1 = ~quantile(., 0.25, na.rm = TRUE),
        median = ~median(., na.rm = TRUE),
        mean = ~mean(., na.rm = TRUE),
        q3 = ~quantile(., 0.75, na.rm = TRUE),
        max = ~max(., na.rm = TRUE),
        sd = ~sd(., na.rm = TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    cols = everything(),
    names_to = c("variable", ".value"),
    names_sep = "_"
  )

outliers_2025 <- gapminder %>% 
  filter(year == 2025) %>% 
  select(country, pop, gdp, lex) %>% 
  pivot_longer(cols = c(pop, gdp, lex), names_to = "variable", values_to = "value") %>% 
  group_by(variable) %>% 
  mutate(
    q1 = quantile(value, 0.25, na.rm = TRUE),
    q3 = quantile(value, 0.75, na.rm = TRUE),
    iqr = q3 - q1,
    lower = q1 - 1.5 * iqr,
    upper = q3 + 1.5 * iqr,
    is_outlier = value < lower | value > upper
  ) %>% 
  ungroup() %>% 
  filter(is_outlier) %>% 
  arrange(variable, desc(value)) %>% 
  select(variable, country, value, lower, upper)

numeric_summary
outliers_2025
```

```{r}
gapminder %>% 
  filter(year == 2025) %>% 
  pivot_longer(cols = c(pop, gdp, lex), names_to = "variable", values_to = "value") %>% 
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "lightblue") +
  scale_y_log10() +
  labs(
    title = "2025 Distributions by Variable (Log Scale)",
    x = "Variable",
    y = "Value (log10 scale)"
  ) +
  theme_minimal()
```

13. Global trends over time (population, weighted life expectancy, and median GDP per capita).
```{r}
global_yearly <- gapminder %>% 
  group_by(year) %>% 
  summarize(
    global_population = sum(pop, na.rm = TRUE),
    weighted_mean_lex = weighted.mean(lex, pop, na.rm = TRUE),
    median_gdp = median(gdp, na.rm = TRUE),
    .groups = "drop"
  )

global_yearly
```

```{r}
global_yearly %>% 
  ggplot(aes(year, weighted_mean_lex)) +
  geom_line(color = "firebrick", linewidth = 1) +
  labs(
    title = "Population-Weighted Global Life Expectancy Over Time",
    x = "Year",
    y = "Weighted Mean Life Expectancy"
  ) +
  theme_minimal()
```

```{r}
global_yearly %>% 
  ggplot(aes(year, median_gdp)) +
  geom_line(color = "navy", linewidth = 1) +
  scale_y_log10() +
  labs(
    title = "Global Median GDP per Capita Over Time",
    x = "Year",
    y = "Median GDP per Capita (log10 scale)"
  ) +
  theme_minimal()
```

14. Relationship diagnostics: how the GDP-life expectancy correlation changes over time.
```{r}
correlation_by_year <- gapminder %>% 
  filter(gdp > 0) %>% 
  group_by(year) %>% 
  summarize(
    cor_log10gdp_lex = cor(log10(gdp), lex, use = "complete.obs"),
    .groups = "drop"
  )

correlation_by_year
```

```{r}
correlation_by_year %>% 
  ggplot(aes(year, cor_log10gdp_lex)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Correlation Between log10(GDP) and Life Expectancy Over Time",
    x = "Year",
    y = "Pearson Correlation"
  ) +
  theme_minimal()
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! 
