---
title: "Data Visualization: `ggplot` part 3"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Build histograms and density plots.  
2. Create new categories with `case_when()`.

## Review
At this point you should feel comfortable making a variety of plots and adjusting aesthetics. This includes making bar plots, scatter plots, and box plots. Last time we explored grouping options like `color`, `fill`, and `group`. 


## Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(palmerpenguins)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

## Data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  
```{r}
homerange <- read_csv(here("lab10", "data", "Tamburelloetal_HomeRangeDatabase.csv"), na = c("", "NA", "\\"))
```

## Histograms and Density Plots
Histograms are frequently used by biologists; they show the distribution of continuous variables. As students, you almost certainly have seen histograms of grade distributions. Without getting into the statistics, a histogram `bins` the data and you specify the number of bins that encompass a range of observations. For something like grades, this is easy because the number of bins corresponds to the grades A-F. By default, R uses a formula to calculate the number of bins but some adjustment is often required.  

What does the distribution of body mass look like in the `homerange` data?
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram(alpha = 0.6, color = "black")+
  labs(title = "Distribution of Body Mass")
```

`Density plots` are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_density(fill="steelblue", alpha  =0.6, color = "black")+
  labs(title = "Distribution of Body Mass")
```

I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r message=FALSE}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(aes(y = ..density..), binwidth = .4, fill = "steelblue", alpha = 0.8, color = "black")+
  geom_density(color = "red")+
  labs(title = "Distribution of Body Mass")
```

## Practice
1. Make a histogram of `log10.hra`. Make sure to add a title.
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra)) +
  geom_histogram(fill="steelblue", alpha=0.8, color="black")+
  labs(title = "Distribution of Homerange")
```

2. Now plot the same variable using `geom_density()`.
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra)) +
  geom_density(fill="steelblue", alpha=0.8, color="black")+
  labs(title = "Distribution of Homerange")
```

3. Combine them both!
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra)) +
  geom_histogram(aes(y = ..density..), binwidth = .4, fill="steelblue", alpha=0.8, color="black")+
  geom_density(color="red")+
  labs(title = "Distribution of Homerange")
```

## Create Categories with mutate and case_when()
`case_when()` is a very handy function from `dplyr` which allows us to calculate a new variable from other variables. We use `case_when()` within `mutate()` to do this.`case_when()` allows us to specify multiple conditions. Let's reclassify the body mass variable into a new factor variable with small, medium, and large animals. 
```{r}
homerange <- homerange %>% 
             mutate(mass_category = case_when(log10.mass <= 1.75 ~ "small",
                                              log10.mass > 1.75 & log10.mass <= 2.75 ~ "medium",
                                              log10.mass > 2.75 ~ "large"))
```

Here we check how the newly created body mass categories compare across trophic guild. 
```{r}
homerange %>% 
  ggplot(aes(x = mass_category, fill = trophic.guild)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Observations by Taxon and Mass Category in Homerange Data",
       x = "Mass Category",
       fill = "Trophic Guild")
```

## Practice
1. Use case_when() to make a new column `range_category` that breaks down `log10.hra` into very small, small, medium, and large classes based on quartile.
```{r}
summary(homerange$log10.hra)
```

```{r}
#install.packages("gtools")
library(gtools)
quartiles <- quantcut(homerange$log10.hra)
table(quartiles)
```

```{r}
homerange <- 
  homerange %>% 
  mutate(range_category = case_when(log10.hra <3.65 ~ "very small",
                                    log10.hra >=3.65 & log10.hra <4.59 ~ "small",
                                    log10.hra >=4.59 & log10.hra <=6.02 ~ "medium",
                                    log10.hra >=6.02 ~ "large"))
```

2. How many and which taxonomic classes are represented in each of the `range_category` classes?
```{r}
homerange %>% 
  ggplot(aes(x=range_category, fill=class))+
  geom_bar(position="dodge")
```

3. Isolate the small `range_category` and plot the range of `log10.mass` by taxonomic class.
```{r}
homerange %>% 
  filter(range_category=="small") %>% 
  ggplot(aes(x=class, y=log10.mass, fill=class)) +
  geom_boxplot(alpha=0.75)
```

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)