---
title: "Lab 5: Visualizing Data, Part 3"
author: "Joel Ledford"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Project Update
Meet with your group and discuss the data that you found for your project. We will then take a few minutes to have each group briefly describe the data that they plan to use in their projects.  

## Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [R for Data Science](https://r4ds.had.co.nz/)
- [R Cookbook](http://www.cookbook-r.com/)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)
- [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Build stacked barplots of categorical variables.  
2. Build side-by-side barplots using `position= "dodge"`.  
3. Customize colors in plots using `RColorBrewer` and `paletteer`.  
4. Build histograms and density plots.  

## Install `RColorBrewer` and `paletteer`
`RColorBrewer` and `paletteer` are helpful in selecting palettes of color that work with your data. Please install these now, we will work with them later in the session.
```{r, eval=FALSE, echo=TRUE}
install.packages("RColorBrewer")
install.packages("paletteer")
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(RColorBrewer)
library(paletteer)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

## Data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

```{r message=FALSE, warning=FALSE}
homerange <- readr::read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```

A little bit of cleaning to focus on the variables of interest.
```{r}
homerange %<>%  
  select(common.name, taxon, class, family, genus, species, mean.mass.g, log10.mass, mean.hra.m2, log10.hra, preymass, log10.preymass, trophic.guild)
```

## Barplots revisited
At this point you should be able to build a barplot that shows counts of observations in a variable using `geom_bar()`. But, you should also be able to use `stat="identity"` to specify both x and y axes.  

Although we did not use it last time, `geom_col()` is the same as specifying `stat="identity"` using `geom_bar()`.     

####Here is the plot using `geom_bar(stat="identity")`
```{r}
homerange %>% 
  filter(family=="salmonidae") %>% #filter for salmonid fish only
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_bar(stat="identity")
```

####Here is the plot using `geom_col`
```{r}
homerange %>% 
  filter(family=="salmonidae") %>% #filter for salmonid fish only
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_col()
```

## Barplots and multiple variables
Last time we explored the `fill` option in barplots as a way to shade by another variable. In this cases the variable was  continuous, but what happens when a fill variable is categorical?

Let's start by counting how many obervations we have in each taxonomic group.
```{r}
homerange %>% 
  count(taxon)
```

Now let's make a barplot of these data.
```{r}
homerange %>% 
  ggplot(aes(x=taxon))+
  geom_bar(alpha=0.9, na.rm=T)+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group")
```

By specifying `fill=trophic.guild` we build a stacked bar plot that shows the proportion of a given taxonomic group that is an herbivore or carnivore.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(alpha=0.9, na.rm=T)+
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")
```

We can also have the proportions of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(alpha=0.9, na.rm=T, position="dodge")+
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")
```

## RColorBrewer
The default color palettes used by ggplot are often uninspiring. They don't make plots pop out in presentations or publications, and you may want to use a customized palette to make things visually consistent.  

Access the help for RColor Brewer.
```{r}
?RColorBrewer
```

The key thing to notice is that there are three different palettes: 1) sequential, 2) diverging, and 3) qualitative. Within each of these there are several selections. You can bring up the colors by using `display.brewer.pal()`. Specify the number of colors that you want and the palette name.
```{r}
display.brewer.pal(8,"BrBG")
```

The [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) website is very helpful for getting an idea of the color palettes. To make things easy, use these two guidelines:

+`scale_colour_brewer()` is for points  
+`scale_fill_brewer()` is for fills  

Here I chose the `Paired` palette. Take a moment and experiment with other options.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(alpha=0.9, na.rm=T)+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")+
  theme_linedraw()+
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 60, hjust=1))+
  scale_fill_brewer(palette = "Paired")
```

```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(alpha=0.9, na.rm=T, position="dodge")+
 labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")+
  theme_linedraw()+
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 60, hjust=1))+
  scale_fill_brewer(palette = "Paired")
```

## Manually Setting Colors
You can also use `RColorBrewer` or `paleteer` to build a custom palette for consistency. To access the `paleteer` collection, I add it to a new object.
```{r}
colors <- paletteer::palettes_d_names
```

Now we can display the palettes. Assign the palette to `my_palette` and then build this base R barplot. There are a lot of options; `paleteer` is a collection of popular palettes. I really like the [`ggsci` package] (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)
```{r}
my_palette <- paletteer_d(package = "ggsci", palette = "springfield_simpsons")
barplot(rep(1,14), axes=FALSE, col=my_palette)
```

Now we just identify `my_palette` as part of `scale_fill_manual()`
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(alpha=0.9, na.rm=T, position="dodge")+
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")+
 scale_fill_manual(values=my_palette)
```

## Histograms and Density Plots
Histograms are frequently used by biologists; they show the distribution of a `continuous` variable. As students, you almost certainly have seen histograms of grade distributions. Without getting into the statistics, a histogram `bins` the data and you specify the number of bins that encompass a range of observations. For something like grades, this is easy because the number of bins corresponds to the grades A-F. By default, R uses a formula to calculate the number of bins but some adjustment is often required.  

What is the distribution of body mass look like in the homerange data? When adjusting binwidth note that we are using `log10.mass`.
```{r}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(binwidth = .4, fill="#709AE1", alpha=0.8, color="black")
```

`Density plots` are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_density(fill="#709AE1", alpha=0.4)
```

I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r message=FALSE}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(aes(y = ..density..), binwidth = .4, fill="#709AE1", alpha=0.8, color="black")+
  geom_density(color="red")
```

## Faceting- `facet_wrap()`
Faceting is one of the amazing features of ggplot. It allows you to make multi-panel plots for easy comparison. Here, I am making histograms of log10.mass for every taxon. We read the `~` in the `facet_wrap()` layer as `by`. The bin number doesn't work for every taxon, but it is enough to get the idea.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram(binwidth = .5, fill="#709AE1", alpha=0.8, color="black") +
  facet_wrap(~taxon)+
    theme(strip.text.x = element_text(size=12))
```

## Practice
1. Revisit the relationship bewtween mass and homerange using a scatterplot, but color the points by taxonomic class using a custom color theme.
```{r}
homerange %>% 
  ggplot(aes(x=log10.mass, y=log10.hra, color=class)) +
  geom_jitter(alpha=0.6)+
  scale_color_brewer(palette = "Dark2")
```

2. Build stacked and dodged barplots that show the representation by taxonomic class with carnivores and herbivores. Try using a color theme that matches your scatterplot to keep everything consistent.
```{r}
homerange %>% 
  ggplot(aes(x=trophic.guild, fill=class)) +
  geom_bar(alpha=0.9)
```

```{r}
homerange %>% 
  ggplot(aes(x=trophic.guild, fill=class)) +
  geom_bar(alpha=0.9, position="dodge")
```

3. 

```{r}
homerange %>%
  ggplot(aes(x = log10.mass)) + 
  geom_density(aes(fill=class), alpha=0.25)
```