---
title: "Midterm 2 W25"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Instructions
Before starting the exam, you need to follow the instructions in `02_midterm2_cleaning.Rmd` to clean the data. Once you have cleaned the data and produced the `heart.csv` file, you can start the exam.  

Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code must be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above.  

Your code must knit in order to be considered. If you are stuck and cannot answer a question, then comment out your code and knit the document. You may use your notes, labs, and homework to help you complete this exam. Do not use any other resources- including AI assistance or other students' work.  

Don't forget to answer any questions that are asked in the prompt! Each question must be coded; it cannot be answered by a sort in a spreadsheet or a written response.  

All plots should be clean, with appropriate labels, and consistent aesthetics. Poorly labeled or messy plots will receive a penalty. Your plots should be in color and look professional!   

Be sure to push your completed midterm to your repository and upload the document to Gradescope. This exam is worth 30 points.  

## Load the libraries
You may not use all of these, but they are here for convenience.
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("shiny")
library("ggthemes")
library("RColorBrewer")
library("paletteer")
```

## Load the data
These data are a modified version of the Statlog (Heart) database on heart disease from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/145/statlog+heart). The data are also available on [Kaggle](https://www.kaggle.com/datasets/ritwikb3/heart-disease-statlog/data).  

You will need the descriptions of the variables to answer the questions. Please reference `03_midterm2_descriptions.Rmd` for details.  

Run the following to load the data.  
```{r message=FALSE, warning=FALSE}
heart <- read_csv("data/heart.csv")
```

## Questions
Problem 1. (1 point) Use the function of your choice to provide a data summary.
```{r}
glimpse(heart)
head(heart)
```

Problem 2. (1 point) Let's explore the demographics of participants included in the study. What is the number of males and females? Show this as a table.  

_There are 87 females and 183 males in the study._

```{r}
heart %>% count(gender)
```

Problem 3. (2 points) What is the average age of participants by gender? Show this as a table.
```{r}
heart %>% 
  group_by(gender) %>% 
  summarize(mean_age=mean(age))
```
_The average age of male and female participants are ~56 and ~54, respectively._

Problem 4. (1 point) Among males and females, how many have/do not have heart disease? Show this as a table, grouped by gender.  

_For the females, 20 have heart disease and 67 do not. For the males, 100 have heart disease and 83 do not._

```{r}
heart %>%
  tabyl(gender, target)
```

Problem 5. (4 points) What is the percentage of males and females with heart disease? Show this as a table, grouped by gender.  

_23% of females and 55% of males have heart disease._

```{r}
disease_percent <- heart %>% 
  group_by(gender) %>% 
  count(target) %>% 
  mutate(percent_disease = (n/sum(n)*100))
```

```{r}
heart %>% 
  group_by(gender, target) %>% 
  summarize(n=n(), .groups = 'keep') %>% 
  pivot_wider(names_from=target, 
              values_from=n) %>% 
  mutate(total=sum(disease, no_disease),
         percent_heart_disease=(disease/total)*100)
disease_percent
```

```{r}
heart %>%
  count(gender, target) %>%
  pivot_wider(names_from = target, 
              values_from = n, 
              values_fill = 0) %>%
  mutate(total = disease + no_disease,
         percent_heart_disease = (disease / total) * 100)
```

Problem 6. (3 points) Make a plot that shows the results of your analysis from problem 5. If you couldn't get the percentages to work, then make a plot that shows the number of participants with and without heart disease by gender.
```{r}
disease_percent %>% 
  filter(target=="disease") %>% 
  ggplot(aes(x=gender, y=percent_disease, fill=gender))+
  geom_col()+
  labs(title="% Heart Disease by Gender", 
       x= "Gender", 
       y="%")+
  scale_fill_brewer(palette = "Paired")+
  theme_clean()+ 
  theme(legend.position='none')
```

Problem 7. (3 points) Is there a relationship between age and cholesterol levels? Make a plot that shows this relationship separated by gender (hint: use faceting or make two plots). Be sure to add a line of best fit (linear regression line).

_There positive relationship among females and a slightly positive relationship among males._

```{r message=FALSE}
heart %>% 
  ggplot(aes(x=age, y=chol))+
  geom_jitter()+
  facet_wrap(~gender)+
  geom_smooth(method="lm")+
  labs(title="Age vs. Cholesterol Levels", 
       x= "Age", 
       y="Serum cholesterol level in mg/dL")+
  theme_clean()
```

Problem 8. (3 points) What is the range of resting blood pressure for participants by type of chest pain? Make a plot that shows this information.
```{r}
heart %>% 
  ggplot(aes(x=cp, y=trestbps, fill=cp))+
  geom_boxplot()+
  labs(title="Resting Blood Pressure by Chest Pain", 
       x= NULL, 
       y="Resting Blood Pressure (mmHg)")+
  scale_fill_brewer(palette = "Paired")+
  theme_clean()+ 
  theme(legend.position='none')
```

Problem 9. (4 points) What is the distribution of maximum heart rate achieved, separated by gender and whether or not the patient has heart disease? Make a plot that shows this information- you must use faceting.
```{r}
heart %>% 
  ggplot(aes(x=thalach, fill=target))+
  geom_density(alpha=0.6)+
  facet_grid(gender~target, labeller=label_both)+
  labs(title="Maximum Heart Rate")+
  scale_fill_brewer(palette = "Paired")+
  theme_clean()+ 
  theme(legend.position='none')
```

Problem 10. (4 points) What is the range of ST depression (oldpeak) by the number of major vessels colored by fluoroscopy (ca)? Make a plot that shows this relationship. (hint: should ca be a factor or numeric variable?)
```{r}
heart %>% 
  mutate(ca=as.factor(ca)) %>%
  select(ca, oldpeak) %>% 
  ggplot(aes(x=ca, y=oldpeak, fill=ca))+
  geom_boxplot()+
  labs(title="Range of ST Depression", 
       x= "# Major Blood Vessels", 
       y="ST Depression")+
  scale_fill_brewer(palette = "Paired")+
  theme_clean()+ 
  theme(legend.position='none')
```

Problem 11. (4 points) Is there an age group where we see increased prevalence of heart disease? Make a plot that shows the number of participants with and without heart disease by age group.

_For the participants in the study, there is an increase in prevalence of heart disease at age 58._

```{r}
heart %>% 
  ggplot(aes(x=as.factor(age), fill=target))+
  geom_bar(position="dodge")+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title="Heart Disease by Age Group", 
       x= "Age", 
       y="n")+
  scale_fill_brewer(palette = "Paired")+
  theme_clean()+ 
  theme(legend.position="bottom")
```

```{r}
library(tidyverse)

# Read the CSV file (adjust the file name/path if needed)
df <- heart

# Filter for patients with heart disease, then count by exact age
heart_disease_counts <- df %>%
  filter(target == "disease") %>%
  group_by(age) %>%
  summarise(count = n())

# Plot the counts by exact age
ggplot(heart_disease_counts, aes(x = age, y = count)) +
  geom_line(color = "darkred") +
  geom_point(color = "darkred") +
  labs(
    title = "Count of Heart Disease Cases by Age",
    x = "Age",
    y = "Count of Heart Disease Cases"
  ) +
  theme_minimal()

```

```{r}
library(tidyverse)

df <- heart

# Create age bins (adjust the breaks as appropriate for your data)
df <- df %>%
  mutate(age_bin = cut(age, breaks = seq(30, 80, by = 5), include.lowest = TRUE))

# Filter for heart disease cases and count by age bin
heart_disease_counts_bin <- df %>%
  filter(target == "disease") %>%
  group_by(age_bin) %>%
  summarise(count = n())

# Plot the counts by age bins as a bar chart
ggplot(heart_disease_counts_bin, aes(x = age_bin, y = count)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Count of Heart Disease Cases by Age Group",
    x = "Age Group",
    y = "Count of Heart Disease Cases"
  ) +
  theme_minimal()

```

```{r}
library(tidyverse)

# Read in your data (adjust the file name/path if needed)
df <- heart

# Create a numeric indicator for heart disease (1 = disease, 0 = no_disease)
df <- df %>%
  mutate(target_num = if_else(target == "disease", 1, 0))

# Fit a logistic regression model with age as the predictor
model <- glm(target_num ~ age, data = df, family = "binomial")

# Display the model summary
summary(model)

```

The model indicates that age is positively associated with the likelihood of heart disease. Specifically, each additional year increases the odds by approximately 5%. The association is statistically significant.

```{r}
# Create age bins (adjust breaks as needed)
df <- heart %>%
  mutate(age_bin = cut(age, breaks = seq(30, 80, by = 5), include.lowest = TRUE))

# Count heart disease cases per age bin
age_bin_counts <- df %>%
  filter(target == "disease") %>%
  group_by(age_bin) %>%
  summarise(count = n())

# View the binned counts
print(age_bin_counts)

```

## Submit the Midterm
1. Save your work and knit the .rmd file.  
2. Open the .html file and "print" it to a .pdf file in Google Chrome (not Safari).    
3. Go to the class Canvas page and open Gradescope.  
4. Submit your .pdf file to the midterm assignment- be sure to assign the pages to the correct questions.  
5. Commit and push your work to your repository. 