---
title: "Midterm 2 W25 Cleaning"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Cleaning
In order to make the data easier to work with, I have cleaned the data and adjusted some of the variables. For the exam, we will work with the cleaned data `heart.csv`. All you need to do is run the code below!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
These data are a modified version of the Statlog (Heart) database on heart disease from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/145/statlog+heart). The data are also available on [Kaggle](https://www.kaggle.com/datasets/ritwikb3/heart-disease-statlog/data).  
```{r message=FALSE, warning=FALSE}
heart <- read_csv("data/Heart_disease_statlog.csv")
```

## Summary
Before running the next code chunk, you should have a look at the data to understand why I made the changes below.
```{r}
glimpse(heart)
```

## Clean the data
```{r}
heart <- heart %>% 
  clean_names() %>% 
  rename(gender = sex) %>%
  mutate(
    gender = ifelse(gender == 1, "male", "female"),
    slope = case_when(
      slope == 0 ~ "upsloping",
      slope == 1 ~ "flat",
      slope == 2 ~ "downsloping"
    ),
    cp = case_when(
      cp == 0 ~ "typical_angina",
      cp == 1 ~ "atypical_angina",
      cp == 2 ~ "non_anginal_pain",
      cp == 3 ~ "asymptomatic"
    ),
    fbs = ifelse(fbs == 0, "false", "true"),
    restecg = case_when(
      restecg == 0 ~ "normal",
      restecg == 1 ~ "t_wave_abnormality",
      restecg == 2 ~ "left_ventricular_hypertrophy"
    ),
    exang = ifelse(exang == 0, "no", "yes"),
    thal = case_when(
      thal == 0 ~ "unknown",
      thal == 1 ~ "normal",
      thal == 2 ~ "fixed_defect",
      thal == 3 ~ "reversable_defect"
    ),
    target = ifelse(target == 0, "no_disease", "disease")
  )
```

## Export to .csv
```{r}
write.csv(heart, "data/heart.csv", row.names = FALSE)
```