---
title: "Data Visualization: Violin Plots & Heat Maps"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
    keep_md: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Build a violin plot with `geom_violin`.
2. Build heat maps using `geom_tile`.


## Load the Libraries
```{r}
library(tidyverse)
library(janitor)
library("palmerpenguins")

options(scipen=999) #cancels the use of scientific notation for the session


```

## Load The Data

We will be using the `palmer penguins` data for checking out violin plots.
We will use a slightly modified version of the `heartrate` data for trying out heat maps.

We will use the homerange data for practice. I've made some modifications that will be important for part 2 of the lab, so we will load it all in now.
```{r}

penguins <- penguins

heartrate <- readr::read_csv("data/heartrate.csv")
heartrate <- heartrate %>%
  add_column(no_drug = c(71, 79, 66, 62, 73, 81) )


homerange <- read_csv("data/modified_Tamburelloetal_HomeRangeDatabase.csv")
```

## Violin Plots

  Violin plots are relatively common in scientific publications. They show 
the distribution of continuous data across a categorical variable.

  Violin plots are sometimes chosen as an alternative to box plots, especially if you expect the data to have a [non-normal distribution](https://en.wikipedia.org/wiki/Normal_distribution) (i.e. not a bell curve). They are very similar to density plots 
  
  They are great for looking at the range & spread of data, while roughly comparing averages across categorical variables.

```{r}
head(penguins)
```


**Let's say we want to travel to an island and take pictures of the cutest little penguines we can find. We need to determine which island we are most likely to find the smallest penguins on.**

We can use a boxplot look at how penguin mass changes by island:
```{r}
penguins %>% 
  ggplot(aes(x = island, y = body_mass_g, fill = island)) +
  geom_boxplot(na.rm = TRUE)+
  coord_flip()
```

It looks like pretty much a tie between Torgersen and Dream islands, but sometimes we miss information about the spread of data with box plots. 

Lets try a Violin plot! We can even layer it over the box plot to compare.
```{r}
penguins %>% 
  ggplot(aes(x = island, y = body_mass_g, fill = island)) +
  geom_boxplot(color = "grey", alpha = .3, na.rm = TRUE) +
  geom_violin(alpha = .5, na.rm = TRUE) +
  coord_flip()
```

It looks like there is kind of a non-normal shape in at least 2 islands- 
this could be a difference in sex, so lets examine only females:
```{r}
penguins %>% 
  filter(sex == "female") %>% 
  ggplot(aes(x = island, y = body_mass_g, fill = island)) +
  geom_boxplot(color = "grey", alpha = .3, na.rm = TRUE) +
  geom_violin(alpha = .5, na.rm = TRUE) +
  coord_flip()
```
Well that didnt really reveal the reason for the bimodal shape on Biscoe. 
Let's try filling by species.
```{r}
penguins %>% 
  ggplot(aes(x = island, y = body_mass_g, fill = species)) +
  geom_violin(alpha = .3, na.rm = TRUE) +
  geom_boxplot(color = "grey", alpha = .3, na.rm = TRUE) +
  coord_flip()
```


Cool! Now we see that there are actually some of the smallest penguins on Biscoe island. 

## Practice

Try making a violin plot of taxon from the homerange data:
 1. Remind yourself what the data look like:

```{r}

```

  2. Make a violin plot of `log10.mass` by `taxon`, and a box plot of `log10.mass` by `taxon`. You can use 1 or 2 plots.
```{r}

```

3. What information can you gather from the violin plot that might have been missed with the box plot?


## Heat maps

Heat maps are cool for a lot of things, but they are particularly good for showing effect size, or magnitude of change.

Let's check out the `heartrate` data, which show results from a drug trial with four treatments on six patients. The `none` column represents heart rate with no drug, and it will act as a control.  
```{r}
heartrate
```

Let's pivot this data, so there is no data stored in variables, then we can use `geom_tile()` to represent the values (heartrate) by patient and treatment. 
```{r}
heartrate_long <- heartrate %>% 
  pivot_longer(-patient, #patient does not pivot
               names_to = "drug", 
               values_to = "heartrate"
               )
heartrate_long
```

```{r}
heartrate_long %>%
  ggplot(aes(patient, drug, fill= heartrate)) + 
  geom_tile() +
  scale_fill_gradient(low="yellow", high="red")
```

Now, let's instead look at how each drug changed the heartrate from the normal, using mutate
```{r}
heartrate_diff <- heartrate%>%
  mutate(across(c(a,b,c,d), ~ . - no_drug)) %>% 
  select(-no_drug) %>% 
    pivot_longer(-patient, #patient does not pivot
               names_to = "drug", 
               values_to = "heartrate"
               ) 
heartrate_diff
```

```{r}
heartrate_diff %>% 
  ggplot(aes(patient, drug, fill= heartrate)) + 
  geom_tile()+
  scale_fill_gradient2(low="green", mid = "white", high="red")
# scale_fill_graient2 lets you set a color direction for zero (mid), positive (high), and negative (low)
```


## Practice
  1. Which drug lowers heart rate? raises it?

  2. Make a heat map for body mass by species & sex for penguins

```{r}

```







## That's it! Take a break and I will see you for part 2!  

-->[Home](https://jmledford3115.github.io/datascibiol/)





