---
title: "Data Visualization: Venn Diagrams & Word Clouds"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Build a venn diagram  
2. Build a word cloud  

## Load Libraries  
```{r}
library(tidyverse)
```

```{r}
#install.packages("ggVennDiagram")
#install.packages("ggwordcloud")
library(ggVennDiagram)
library(RColorBrewer)
library(ggwordcloud)
```

## Data
Let's use the homerange data-- I've slightly modified it with a new observation.  
```{r}
homerange <- read_csv("data/modified_Tamburelloetal_HomeRangeDatabase.csv")
```

## Venn Diagrams  
[great info here](https://cran.r-project.org/web/packages/ggVennDiagram/vignettes/using-ggVennDiagram.html) on how these are coded in ggplot.  

Sometimes its tricky to figure out what kind of data can be represented with a Venn diagram. I like to think of them as asking a series of questions about each row in you dataset, that are not mutually exclusive/inclusive.  

For the animals in the homerange database, we will ask a series of yes/no questions.  

Let's start with: Do they fly? Do they swim?  
We will use `pull` to create a vector of the `common.name`s of animals that would answer this question with "yes".  

Do they fly?  
```{r}
flying_vec <- 
  homerange %>%
  filter(locomotion == "flying")%>%
  pull(common.name) #make a vector of the common names resulting from the filter
flying_vec
```

Do they swim?
```{r}
swimming_vec <- 
  homerange %>%
  filter(locomotion == "swimming")%>%
  pull(common.name)
swimming_vec
```

To build a Venn diagram of these, we make a list or the vectors that represent our questions.  

The list represents the number of circles going into the venn diagram" and the vectors represent "what are the numbers going into the venn diagram"  
```{r}
fly_or_swim_list <- list(flying_vec, swimming_vec)
fly_or_swim_list
```

```{r}
ggVennDiagram(fly_or_swim_list, category.names = c("Fliers", "Swimmers"))
```

Woah! an image! But its not that useful. Since we are pretty familiar with this data, we know that `locomotion` is one column, and each animal only has one entry for locomotion, which is probably their primary for of movement. This results in `Swim` and `Fly` being mutually exclusive.  

So venn diagrams work best when asking questions **about data from 2 different columns**. Let's retry this with a more informative question:  

Do they fly? Are they a bird?  
We can recycle the `flying_vec`  
Do they fly?  
```{r}
# flying_vec
```

Are they a bird?
```{r}
bird_vec <- 
  homerange %>%
  filter(taxon == "birds")%>%
  pull(common.name)
```


```{r}
fly_or_bird_list <- list(flying_vec, bird_vec)

ggVennDiagram(fly_or_bird_list, category.names = c("Fly", "Non-Flying Birds"))
```

What are the 26 non-flying birds?
```{r}
homerange%>%
  filter(taxon == "birds")%>%
  filter(locomotion != "flying")%>%
  select(common.name)
```

what is the flying non-bird?  
```{r}
homerange %>% filter(locomotion=="flying" & taxon!="birds")
```

### Adding Circles  
Are they an herbivore?  
```{r}
herbivore_vec <- 
  homerange %>%
  filter(trophic.guild == "herbivore")%>%
  pull(common.name)
herbivore_vec
```

```{r}
fly_bird_herbi_list <- list(flying_vec, bird_vec, herbivore_vec)
```

```{r}
ggVennDiagram(fly_bird_herbi_list, category.names = c("Fly", "Bird", "Herbivore"))
```

### Make It Pretty!  
Let's work with the 2 question plot for now:  
```{r}
  # you can toggle edge size
venn_plot <- ggVennDiagram(fly_or_bird_list, category.names = c("Fly", "Bird"), edge_size = 2)
```

```{r}
venn_plot +
  scale_fill_distiller(palette = "Greys")+ #distiller is for discrete data
  labs(title = "Do all birds fly? Do only birds fly?",
         caption = ("you can add a caption here, e.g.: Studys show not all birds fly, and not only birds fly"))
```

```{r}
  # you can toggle edge size and label (Fly, Bird) colors here
ggVennDiagram(fly_or_bird_list,
              set_color = c("black", "blue"),             # sets label colors
              category.names = c("Fly", "Bird"),          # sets category label names
              label_color = "firebrick",                  # sets overlap-quantity label color
              label_size = 4                              # sets overlap-label size
              )+
  scale_color_brewer(palette = "Set3")+                   # sets palette for circle colors
  scale_fill_distiller(palette = "Greys")+                # distiller is for discrete data
  labs(title = "Do all birds fly? Do only birds fly?",
         caption = ("Figure 1: you can add a caption here, e.g.: Studys show not all birds fly, and not only birds fly") )
```
## Practice  
Try making a 4 circle Venn diagram to show the intersection of:  
- swimming  
- carnivores  
- aquatic  
- ectotherms  

You can use `locomotion`, `trophic.guild`, `realm`, and `thermoregulation`  
```{r}
#Set Up the vectors:

   #you can also recycle the one we made earlier
swimming_vec<- 
  homerange %>%
  filter(locomotion=="swimming")%>%
  pull(common.name)

carni_vec<- 
  homerange %>%
  filter(trophic.guild=="carnivore")%>%
  pull(common.name)

aqua_vec <- 
  homerange %>% 
  filter(realm == "aquatic")%>%
  pull(common.name)

ecto_vec<- 
  homerange %>%
  filter(thermoregulation =="ectotherm")%>%
  pull(common.name)
```

```{r}
homerange_list <- list(swimming_vec, carni_vec, aqua_vec, ecto_vec)
```

```{r}
venn_plot <- ggVennDiagram(homerange_list, category.names = c("Swimming", "Carni", "Aqua", "Ecto"), edge_size = 2, label="count")
venn_plot
```

Use dplyr to determine which 20 animals from this plot are:  
- Ectotherms  
- Aquatic  
- Swimmers  
- NOT Carnivores?  
```{r}

```

## Word Clouds  
Check out [this reference](https://lepennec.github.io/ggwordcloud/) on how word clouds are built in R.  

Word clouds allow you to make a image of multiple words, where their size is scaled to the frequency that they occur in the dataset.  

First, you have to make a frequency list with count.  
```{r}
hr_taxon_count <- 
  homerange %>% 
  count(taxon)
hr_taxon_count
```

then you can plot like this!!
```{r}
hr_taxon_count %>% 
  ggplot(aes(label = taxon, size = n, color = taxon)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 20) +
  theme_minimal()
```

## Practice  
in the homework, we will make word clouds with the super hero data.  

For now, practice word clouds by making a could of all the different families in the class `reptilia`.
If they don't fit, try tinkering the `scale_size_area(max_size =   )`.  

make the count table:  
```{r}
hr_taxon_count <- homerange %>%
  filter(class=="reptilia") %>% 
  count(family)
hr_taxon_count
```

make the plot:  
```{r}
hr_taxon_count %>% 
  ggplot(aes(label = family, size = n, color = family)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 10) +
  theme_minimal()
```