---
title: "Midterm Review and `across()` "
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Work through midterm 1 and practice some new approaches to solving common problems in R.    
2. Use the `across()` operator to produce summaries across multiple variables. 

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("skimr")
library("janitor")
library("palmerpenguins")
```

## Midterm 1 Review
Let's briefly review the questions from midterm 1 so you can get an idea of how I was thinking about the problems. Remember, there is more than one way to get at these answers, so don't worry if yours looks different than mine!  

## Penguins
We rushed through the second part of lab 7 last week, so let's do a quick review of some key functions. As you know, summarize() and group_by() are powerful tools that we can use to produce clean summaries of data. Especially when used together, we can quickly group variables of interest and save time. Let's do some practice with the [palmerpenguins(https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data to refresh our memory.

```{r}
?penguins
```

Recall that group_by() and summarize() work great together. Let's say we were interested in how body mass varied by island. It is reasonable to assume that the islands are different, so maybe the penguins are as well.
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(mean_body_mass_g=mean(body_mass_g, na.rm=T)) # remember to remove those NA's!
```

What if we are interested in the number of observations (penguins) by species and island?
```{r}
penguins %>% 
  group_by(island, species) %>% 
  summarize(n_penguins=n(), .groups = 'keep')
```

Recall that that `count()` works like a combination of `group_by()` and `summarize()` but just shows the number of observations.
```{r}
penguins %>% 
  count(island, species)
```

`tabyl()` will also produce counts, but the output is different. It is just a matter of personal preference.
```{r}
penguins %>% 
  tabyl(island, species)
```

## Practice
1. Produce a summary of the mean for bill_length_mm, bill_depth_mm, flipper_length_mm, and body_mass_g within Adelie penguins only. Be sure to provide the number of samples.
```{r}

```

2. How does the mean of `bill_length_mm` compare between penguin species?
```{r}

```


3. For some penguins, their sex is listed as NA. Where do these penguins occur?
```{r}

```

## `across()`
How do we use `filter()` and `select()` across multiple variables? There is a function in dplyr called `across()` which is designed to help make this efficient. 

What if we wanted to use `summarize()` to produce distinct counts over multiple variables; i.e. species, island, and sex? Although this isn't a lot of coding you can image that with a lot of variables it would be cumbersome.
```{r}
penguins %>%
  summarize(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_sex = n_distinct(sex))
```

By using `across()` we can reduce the clutter and make things cleaner. 
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```

This is very helpful for continuous variables.
```{r}
penguins %>%
  summarize(across(contains("mm"), mean, na.rm=T))
```

`group_by` also works.
```{r}
penguins %>%
  group_by(sex) %>% 
  summarize(across(contains("mm"), mean, na.rm=T))
```

Here we summarize across all variables.
```{r}
penguins %>%
  summarise_all(n_distinct)
```

Operators can also work, here I am summarizing `n_distinct()` across all variables except `species`, `island`, and `sex`.
```{r}
penguins %>%
  summarize(across(!c(species, island, sex, year), 
                   mean, na.rm=T))
```

All variables that include "bill"...all of the other dplyr operators also work.
```{r}
penguins %>%
  summarise(across(starts_with("bill"), n_distinct))
```

```{r}
names(penguins)
```

## Practice
1. Produce separate summaries of the mean and standard deviation for bill_length_mm, bill_depth_mm, and flipper_length_mm for each penguin species. Be sure to provide the number of samples.  
```{r}

```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  