---
title: "Tidyr 2: `pivot_wider()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Explain the difference between tidy and messy data.  
2. Evaluate a data set as tidy or untidy.  
3. Use the `pivot_wider()` function of `tidyr` to transform data from long to wide format.  

## Breakout Rooms  
Please take 5-8 minutes to check over your answers to HW 7 in your group. If you are stuck, please remember that you can check the key in [Joel's repository](https://github.com/jmledford3115/BIS15LW2022_jledford).  

## Midterm 1 Review
Let's briefly review the questions from midterm 1 so you can get an idea of how I was thinking about the problems. Remember, there is more than one way to get at these answers, so don't worry if yours looks different than mine!  

## Group Project
Let's take a few minutes to discuss group projects. What data interest you? Make a plan on doing some internet searches and thinking about data that might work. Think about the kinds of questions you might ask and how they might be addressed. Please complete the [class project survey](https://forms.gle/EJkbGyHN5MycyQpSA) as part of your homework.  

## Resources  
- [tidyr `pivot_wider()`](https://tidyr.tidyverse.org/reference/pivot_wider.html)  
- [pivoting](https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html)  
- [R Ladies Sydney](https://rladiessydney.org/courses/ryouwithme/02-cleanitup-5/)  

## Review
Last time we learned the fundamentals of tidy data and used the `pivot_longer()` function to wrangle a few examples of frequently encountered untidy data. In the second part of today's lab we will look at a few more examples of `pivot_longer()` but also use the `pivot_wider()` function to deal with another type of untidy data.  

## Load the tidyverse
```{r}
library("tidyverse")
```

## `pivot_longer()`
Recall that we use `pivot_longer()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
datasets::USPersonalExpenditure
?USPersonalExpenditure
```

Here we add a new column of expenditure types, which are stored as rownames above, with `mutate()`. The `USPersonalExpenditures` data also needs to be converted to a data frame before we can use the tidyverse functions, because it comes as a matrix.
```{r}
expenditures <- 
  USPersonalExpenditure %>% 
  as_tibble() %>% 
  mutate(expenditure = rownames(USPersonalExpenditure))
expenditures
```

## Practice
1. Are these data tidy? Please use `pivot_longer()` to tidy the data.
```{r}
expenditures_long <- expenditures %>% 
  pivot_longer(-expenditure,
               names_to="year",
               values_to="bn_dollars")
expenditures_long
```

2. Restrict the data to medical and health expenditures only. Sort in ascending order.
```{r}
expenditures_long %>% 
  filter(expenditure=="Medical and Health") %>% 
  arrange(bn_dollars)
```

## `separate()`
In this new heart rate example, we have the sex of each patient included with their name. Are these data tidy? No, there is more than one value per cell in the patient column and the columns a, b, c, d once again represent values.
```{r}
heartrate2 <- readr::read_csv("data/heartrate2.csv")
heartrate2
```

We need to start by separating the patient names from their sexes. `separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_")
```

## Practice
1. Re-examine `heartrate2`. Use `separate()` for the sexes, `pivot_longer()` to tidy, and `arrange()` to organize by patient and drug. Store this as a new object `heartrate3`.  
```{r}
heartrate3 <- 
  heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_") %>% 
  pivot_longer(-c(patient, sex),
               names_to="drug",
               values_to="heartrate"
               ) %>% 
  arrange(sex, drug, heartrate)
heartrate3
```

2. `unite()` is the opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.  Give it a try below by recombining patient and sex from `heartrate3`.  
```{r}
heartrate3 %>% 
  unite(patient_sex, "patient", "sex", sep="_")
```

## `pivot_wider()`
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  

Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  

```{r}
tb_data <- read_csv("data/tb_data.csv")
tb_data
```

When using `pivot_wider()` we use `names_from` to identify the variables (new column names) and `values_from` to identify the values associated with the new columns.
```{r}
tb_data %>% 
  pivot_wider(names_from = "key", #the observations under key will become new columns
              values_from = "value")
```

## Practice
1. Load the `gene_exp.csv` data as a new object `gene_exp`. Are these data tidy? Use `pivot_wider()` to tidy the data.
```{r}
gene_exp <- readr::read_csv("data/gene_exp.csv")
gene_exp
```

```{r}
gene_exp %>% 
  pivot_wider(names_from = "type",
              values_from = "L4_values")
```

## `pivot_wider()` with multiple variables  
What happens when we want to use `pivot_wider()` with multiple columns?
```{r}
edu_level <- readr::read_csv("data/education_level.csv")
edu_level
```

We want to pivot the `edu_level` data such that there is only a single row per demographic. 
```{r}
edu_level %>% 
pivot_wider(names_from = (education_level), #new column names come from the education_level column
                values_from = c(mean_income, n)) #values come from two separate columns
```

## Practice
For the last practice example, I will use data from the awesome [R Ladies Sydney](https://rladiessydney.org/courses/ryouwithme/02-cleanitup-5/) blog. This data set is compiled by the NSW Office of Environment and Heritage contains the enterococci counts in water samples obtained from Sydney beaches as part of the Beachwatch Water Quality Program! The data set we’ll be working with is current as of October 13th 2018.  

1. Load the beachbugs data and have a look.
```{r}

```

2. Use `pivot_wider` to transform the data into wide format.
```{r}

```

3. Now, use `pivot_longer` to transform them back to long!
```{r}

```

## Wrap-up  
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)