---
title: "Dealing with NA's"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Define `NA` and describe how they are treated in R.  
2. Produce summaries of the number of NA's in a data set.   
3. Replace values with `NA` in a data set as appropriate.  

## Install the package `naniar`
```{r}
#install.packages("naniar")
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("naniar")
library("janitor")
```

## Review
When working with "wild" data, dealing with NA's is a fundamental part of the data cleaning process. Data scientists spend most of their time cleaning and transforming data- including managing NA's. There isn't a single approach that will always work so you need to be careful about using replacement strategies across an entire data set. And, as the data sets become larger NA's can become trickier to deal with.  

For the following, we will use life history data for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from:  
**S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.**  

## Load the mammals life history data and clean the names  
```{r message=FALSE, warning=FALSE}
life_history <- read_csv("data/mammal_lifehistories_v3.csv") %>% clean_names()
```

## Are there any NA's?
Sometimes using one or more of the summary functions can give us clues to how the authors have represented missing data. This doesn't always work, but it is a good place to start.
```{r}
summary(life_history)
```

This will give you a quick summary of the number of NA's in each variable. Notice that, at least for now, it doesn't look like there are NA's in most variables. But, given the size of the data this is unlikely.
```{r}
life_history %>% 
  summarize(across(everything(), ~ sum(is.na(.))))
```

## Practice
1. Can we use a single approach to deal with NA's in this data set? Given what you learned in the previous lab, how would you manage the NA values?
```{r}
life_history_no_nas <- 
  read_csv("data/mammal_lifehistories_v3.csv", na = c("NA", " ", ".", "-999")) %>% 
  clean_names() # notice that I am creating a new dataframe that doesn't have any NA's
```

Recheck
```{r}
life_history_no_nas %>% 
  summarize(across(everything(), ~ sum(is.na(.))))
```

## `naniar`
[naniar](https://naniar.njtierney.com/articles/replace-with-na.html) is a package that is built to manage NA's. Many of the functions it performs can also be performed using tidyverse functions, but it does provide some interesting alternatives.  

`miss_var_summary` provides a summary of NA's across the data frame.
```{r}
miss_var_summary(life_history_no_nas)
```

Notice that `max_life` has no NA's. Does that make sense in the biological context of this data?
```{r}
hist(life_history_no_nas$max_life) #we found another way that NA's are represented
```

```{r}
life_history_no_nas %>%
  filter(max_life==0) %>% 
  select(order, family, genus, species, max_life)
```

Let's use `mutate()` and use `na_if()` to replace 0's with NA's in `max_life`. This chunk allows us to address problems in a single variable.
```{r}
life_history_no_nas <- life_history_no_nas %>% 
  mutate(max_life=na_if(max_life, 0))
```

We can also use `miss_var_summary` with `group_by()`. This helps us better evaluate where NA's are in the data.
```{r}
life_history_no_nas %>%
  group_by(order) %>%
  miss_var_summary(order=T)
```

`naniar` also has a nice replace functions which will allow you to precisely control which values you want replaced with NA's in each variable. This is a nice alternative to `mutate()` and `na_if()`.  
```{r}
life_history %>% #going back to the original data
  replace_with_na(replace = list(newborn = "not measured", 
                                 weaning= -999, 
                                 wean_mass= -999, 
                                 afr= -999, 
                                 max_life= 0, 
                                 litter_size= -999, 
                                 gestation= -999, 
                                 mass= -999)) %>% 
miss_var_summary()
```

You can also use naniar to replace a specific value (like -999) with NA across the entire data set.  
```{r}
life_history %>% #going back to the original data
  replace_with_na_all(condition = ~.x == -999)%>% 
miss_var_summary()
```

Finally, naniar has some built-in examples of common values or character strings used to represent NA's. The chunk below will use these built-in parameters to replace NA's across the entire data set.
```{r}
common_na_strings
```

```{r}
common_na_numbers
```

```{r}
life_history %>% #going back to the original data
  replace_with_na_all(condition = ~.x %in% c(common_na_strings, common_na_numbers)) %>% 
  mutate(newborn=na_if(newborn, "not measured"))
```

## Practice
Let's practice evaluating NA's in a large data set. The data are compiled from [CITES](https://cites.org/eng). This is the international organization that tracks trade in endangered wildlife. You can find information about the data [here](https://www.kaggle.com/cites/cites-wildlife-trade-database).  

Some key information:  
[country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2)  

1. Import the data and do a little exploration. Be sure to clean the names if necessary.
```{r}
cites <- read_csv("data/cites.csv") %>% clean_names()
```

2. Use `naniar` to summarize the NA's in each variable.
```{r}
miss_var_summary(cites)
```

3. Try using `group_by()` with `naniar`. Look specifically at `class` and `exporter_reported_quantity`. For which taxonomic classes do we have the highest number of missing export data?
```{r}
cites %>%
  select(class, exporter_reported_quantity) %>% 
  group_by(class) %>%
  miss_var_summary() %>% 
  arrange(desc(n_miss))
```

## Dealing with NA's in advance- nuclear option
Once you are sure that you know how NA's are treated in the data, then you can deal with them in advance using `na()` as part of the `readr` package. Notice that you still need to deal with some variables seperately. For example, you don't want to replace all 0's with NA, just the `max_life` variable.  
```{r}
read_csv(file = "data/mammal_lifehistories_v3.csv", na = c("NA", " ", ".", "-999", "not measured")) %>% clean_names()
```

## That's it! Let's take a break and then move on to part 2!  

-->[Home](https://jmledford3115.github.io/datascibiol/)