---
title: "Fun with NA's"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Define `NA` and describe how they are treated in R.  
2. Produce summaries of the number of NA's in a data set.   
3. Replace values with `NA` in a data set as appropriate.  

```{r}
#install.packages("naniar")
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("naniar")
library("skimr")
library("janitor")
```

## Review
When working with "wild" data, dealing with NA's is a fundamental part of the data cleaning process. Data scientists spend most of their time cleaning and transforming data- including managing NA's. There isn't a single approach that will always work so you need to be careful about using replacement strategies across an entire data set. And, as the data sets become larger NA's can become trickier to deal with.  

For the following, we will use life history data for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from:  
**S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.**  

## Load the mammals life history data and clean the names  
```{r}
life_history <- read_csv("data/mammal_lifehistories_v3.csv") %>% clean_names()
```

## Are there any NA's?
Sometimes using one or more of the summary functions can give us clues to how the authors have represented missing data. This doesn't always work, but it is a good place to start.
```{r}
glimpse(life_history)
```

```{r}
summary(life_history)
```

Here is a new one for you using the `purrr` package. This will give you a quick summary of the number of NA's in each variable.
```{r}
life_history %>% 
  map_df(~ sum(is.na(.)))
```

## Practice
1. Can we use a single approach to deal with NA's in this data set? Given what you learned in the previous lab, how would you manage the NA values?
```{r}

```

## `naniar`
`naniar` is a package that is built to manage NA's. Many of the functions it performs can also be performed using tidyverse functions, but it does provide some interesting alternatives.  

`miss_var_summary` provides a clean summary of NA's across the data frame.
```{r}
naniar::miss_var_summary(life_history)
```

Notice that `max_life` has no NA's. Does that make sense in the context of this data?
```{r}
hist(life_history$max_life)
```

Let's use `mutate()` and `na_if()` to replace 0's with NA's in `max_life`.
```{r}
life_history <- 
  life_history %>% 
  mutate(max_life=na_if(max_life, 0))
```

```{r}
miss_var_summary(life_history)
```

We can also use `miss_var_summary` with `group_by()`. This helps us better evaluate where NA's are in the data.
```{r}
life_history %>%
  group_by(order) %>%
  select(order, wean_mass) %>% 
  miss_var_summary(order=T)
```

`naniar` also has a nice replace function which will allow you to precisely control which values you want replaced with NA's in each variable.
```{r}
life_history %>% 
  replace_with_na(replace = list(newborn = "not measured", 
                                 weaning= -999, 
                                 wean_mass= -999, 
                                 afr= -999, 
                                 max_life= 0, 
                                 litter_size= -999, 
                                 gestation= -999, 
                                 mass= -999)) %>% 
miss_var_summary()
```

## Practice
Let's practice evaluating NA's in a large data set. The data are compiled from [CITES](https://cites.org/eng). This is the international organization that tracks trade in endangered wildlife. You can find information about the data [here](https://www.kaggle.com/cites/cites-wildlife-trade-database).  

Some key information:  
[country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2)  

1. Import the data and do a little exploration. Be sure to clean the names if necessary.
```{r}

```

2. Use `naniar` to summarize the NA's in each variable.
```{r}

```

3. Try using `group_by()` with `naniar`. Look specifically at class and `exporter_reported_quantity`. For which taxonomic classes do we have a high proportion of missing export data?
```{r}

```

## Visualizing NAs
There is another package `visdat` that can be used to visualize the proportion of different classes of data, including missing data. But, it is limited by size.
```{r}
library(visdat)
```

```{r}
vis_dat(life_history) #classes of data
```

```{r}
vis_miss(life_history)
```

## Dealing with NA's in advance
If you are sure that you know how NA's are treated in the data, then you can deal with them in advance using `na()` as part of the `readr` package.
```{r}
life_history_advance <- 
  readr::read_csv(file = "data/mammal_lifehistories_v3.csv", 
                  na = c("NA", " ", ".", "-999")) #all NA, blank spaces, .,and -999 are treated as NA
```

```{r}
miss_var_summary(life_history_advance)
```

## Wrap-up  
Please review the learning goals and be sure to use the code here as a reference when completing the homework.

-->[Home](https://jmledford3115.github.io/datascibiol/)