---
title: "BIS 15L Packages"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is a package in R?
Packages are free compilations of code that are written to perform specific functions. They are often developed to perform very specific functions or to make some things in R easier to do. We will use many packages in R over the quarter and I will add these to this markdown script. Each time you update R or RStudio, you need to re-install these packages so keeping this script handy will be helpful. Better yet, add packages that you like to use!  

Once you install a package, you don't need to re-install until you update R and RStudio.  In general, you should not re-install packages unless you do an update or there is an error. Continually re-installing packages can lead to problems.  To prevent this problem here, I have commented out `install.packages()`

The package we use the most is the [tidyverse](https://www.tidyverse.org/).  
```{r}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
```

Other useful packages.  
```{r}
install.packages("janitor", repos = "http://cran.us.r-project.org")
install.packages("naniar", repos = "http://cran.us.r-project.org")
install.packages("shiny")
install.packages("here")
install.packages("skimr")
```

Here is a useful alternative to add to scripts that require certain packages.
```{r}
#if (!require(pkg)) { install.packages(pkg); library(pkg) } (you might need character.only=TRUE
```