---
title: "Intro to Spatial Data"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*  
1. Produce distribution maps in R.  
 
## Resources
[Overview of Cooordinate Reference Systems in R](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf)

## Spatial Data in R
There are many packages and techniques for working with spatial data in R. We will cover just some of the basics. One nice package is `ggmap`, which allows us to get base maps from Google Maps, OpenStreetMap, and Stamen Maps. It also works well with `ggplot2`.  

If you completed part 1 of today's lab then you should have the following packages installed. If not, then you should get them going now. You need to do these in order!  

`devtools` and `ggmap`
```{r}
#install.packages("devtools")
#library(devtools)
#devtools::install_github("dkahle/ggmap")
```

`albersusa`
```{r, message = FALSE}
#install.packages("remotes")
#remotes::install_github("hrbrmstr/albersusa")
```

## Load the libraries
```{r, message = FALSE}
library(tidyverse)
library(here)
library(janitor)
library(ggmap)
```

Let's load our processed data from the first part of the lab.
```{r, message = FALSE}
spiders <- read_csv(here("lab12", "data", "spiders_with_locs.csv")) %>% clean_names()
```

There is an error in one of the coordinates that we will fix here.
```{r}
spiders <- spiders %>% filter(latitude<=42)
```

## Create Base Map
Our goal here is to plot the spiders locations from the columns which contain the latitude and longitude. First, we need to get a base map for plotting our points on. We could plot them without a base map, but that wouldn't give us any context as to where they are in space. To get a base map we specify a min and max of each x and y coordinate, and create a bounding box.  

We set the bounding box to a little outside our min and max locations with `f = 0.05`.    

`summary()` gives us our min and max.
```{r}
spiders %>% 
  select(latitude, longitude) %>% 
  summary()
```

Now we set the bounding box.
```{r}
lat <- c(34.67, 41.80)
long <- c(-124.1, -115.5)
bbox <- make_bbox(long, lat, f = 0.05)
```

Let's get a base map for our bounding box area. We will use the stamen maps because they are free. There are several different map types, including: `terrain-labels`, `terrain-lines`, `toner`, `toner-2011`, `toner-background`, `toner-hybrid`, `toner-lines`, `toner-lite`, and `watercolor.`
```{r}
map1 <- get_map(bbox, maptype = "terrain", source = "stamen")
```

```{r}
ggmap(map1)
```

## Adding Points to Base Map
`ggmap` works well with `ggplot2`. To add our points we only need to specify the x and y location similar to how we made charts in previous labs. 
```{r}
ggmap(map1) + 
  geom_point(data = spiders, aes(longitude, latitude)) +
  labs(x = "Longitude", y = "Latitude", title = "Spider Locations")
```

## Practice
1. Map the spider locations with a different type of base map. Try to adjust the map using different aesthetics including `size`, `color`, and `alpha`.

## Albers USA
The Albers USA [package](https://github.com/hrbrmstr/albersusa) is a very fast way to produce clean maps that include nice overlays of counties across the US.
```{r}
library(albersusa)
```

State Boundaries.
```{r}
us_comp <- usa_sf() # get a composite map of the USA
```

County boundaries in each state.
```{r}
cnty_comp <- counties_sf() # get a composite map of USA counties
```

Map of the USA- notice that you can still use themes. `geom_sf` stands for simple feature.
```{r}
ggplot() + 
  geom_sf(data = us_comp, size = 0.125) + 
  theme_linedraw()+
  labs(title = "USA State Boundaries")
```

Since we want to produce a map of our spiders we need to restrict the map to California.
```{r}
ca_comp <- us_comp %>% 
  filter(name=="California")
```

```{r}
ca_cnty_comp <- cnty_comp %>% 
  filter(state=="California")
```

Our base map
```{r}
ggplot() +
  geom_sf(data = ca_comp, size = 0.125)+
  geom_sf(data = ca_cnty_comp, size = 0.125)
```

All we do now is add the `geom_point()` layer.
```{r}
ggplot() +
  geom_sf(data = ca_comp, size = 0.125)+
  geom_sf(data = ca_cnty_comp, size = 0.125)+
  geom_point(data = spiders, aes(longitude, latitude))
```

We can spice things up a bit by adding aesthetics.
```{r}
ggplot() +
  geom_sf(data = ca_comp, size = 0.125)+
  geom_sf(data = ca_cnty_comp, size = 0.125)+
  geom_point(data = spiders, aes(longitude, latitude), size=0.8)+
  labs(x = "Figure 1: Distribution of Usofila spiders")+
  theme_minimal() +
  theme(axis.text = element_blank()) +
  theme(axis.title.x = element_text(hjust=0.5, size = 10)) +
  theme(axis.title.y = element_blank()) +
  theme(panel.grid = element_blank()) +
  theme(legend.position = c(0.9, 0.3)) -> uso_map
uso_map
```

## Practice
1. Make a new column that labels each specimen as occurring in a cave or on the surface. The word "Cave" is part of the locality information.


2. Now adjust the plot above to give a different color for specimens that occur in caves.


## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)