---
title: "Data Visualization: `ggplot` part 5"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Build histograms and density plots.
2. Adjust colors using R's built-in color options.  
3. Create new categories with `case_when()` and use those categories to build plots.  
4. Build line plots.  

## Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [R for Data Science](https://r4ds.had.co.nz/)
- [R Cookbook](http://www.cookbook-r.com/)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)
- [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

## Data
For this tutorial, we will use two data sets.  

[Desert ecology](http://esapubs.org/archive/ecol/E090/118/). The data are from: S. K. Morgan Ernest, Thomas J. Valone, and James H. Brown. 2009. Long-term monitoring and experimental manipulation of a Chihuahuan Desert ecosystem near Portal, Arizona, USA. Ecology 90:1708.  
```{r message=FALSE, warning=FALSE}
deserts <- read_csv("data/surveys_complete.csv")
```

[Homerange](http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1). The data are from: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
```{r message=FALSE, warning=FALSE}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", na = c("", "NA", "\\", "-999"))
```

## Review
1. Use the `homerange` data to make a plot that shows the range of `log10.mass` by taxonomic class.
```{r}

```

2. Now, add a layer to this plot that fills color for each box by taxonomic class.
```{r}

```

3. Use `fill` to show the range of `log10.mass` by taxonomic class with color codes.
```{r}

```

## Line plots
Line plots are great when you need to show changes over time. Here we look at the number of samples for species DM and DS over the years represented in the `deserts` data. This takes some careful thought- we want to know how sampling has changed over time for these two species.    

Start by making a clear x and y so we know what we are going to plot.
```{r}

```

Now let's make the plot.  
```{r}

```

## Histograms
Histograms show the distribution of continuous variables. As students, you have seen histograms of grade distributions. A histogram `bins` the data and you specify the number of bins that encompass a range of observations. For something like grades, this is easy because the number of bins corresponds to the grades A-F. By default, R uses a formula to calculate the number of bins but some adjustment may be required.  

What does the distribution of body mass look like in the `homerange` data?
```{r}

```

Let's play with the colors. This shows all 657 of R's built-in colors. Notice that `color` and `fill` do different things!
```{r}
#colors()
```

Let's rebuild the histogram, but this time we will specify the color and fill. Do a little experimentation on your own with the different colors.
```{r}

```

## Density plots  
Density plots are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}

```

I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r message=FALSE}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(aes(y = after_stat(density)), fill = "deepskyblue4", alpha = 0.4, color = "black")+
  geom_density(color = "red")+
  labs(title = "Distribution of Body Mass")
```

## Practice
1. Make a histogram of `log10.hra`. Make sure to add a title.
```{r}

```

2. Now plot the same variable using `geom_density()`.
```{r}

```

3. Combine them both!
```{r}

```

## Create Categories with mutate and case_when()
`case_when()` is a very handy function from `dplyr` which allows us to calculate a new variable from other variables. We use `case_when()` within `mutate()` to do this.`case_when()` allows us to specify multiple conditions. Let's reclassify the body mass variable into a new factor variable with small, medium, and large animals. In this case, we are making a continuous variable into a categorical variable.  
```{r}

```

```{r}

```

Use gtools to find the quartiles.
```{r}

```

```{r}

```

Here we check how the newly created body mass categories compare across `trophic.guild`. 
```{r}

```

## Practice
1. Use case_when() to make a new column `range_category` that breaks down `log10.hra` into very small, small, medium, and large classes based on quartile.
```{r}

```

```{r}

```

2. Make a plot that shows how many and which taxonomic classes are represented in each `range_category`.   
```{r}

```

3. Isolate the small `range_category` and plot the range of `log10.mass` by taxonomic class.
```{r}

```

## That's it! Let's take a break!  

-->[Home](https://jmledford3115.github.io/datascibiol/)