---
title: "Transforming data 2: `filter()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Use the functions of dplyr (filter, select, arrange) to organize and sort data frames.  
2. Use `mutate()` to calculate a new column from existing columns.  

## Review  
In the previous lab, we used `select()` to extract columns of interest from a data frame. This helps us focus our attention on the variables relevant to our question. However, it doesn't allow us to extract observations from within the data frame. The `filter()` function allows us to extract data that meet specific criteria. When combined with `select()`, we have the power to transform, shape, and explore data with the potential to make new discoveries.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Load the data
For these exercises we will use the data from lab 4_1.  
```{r}
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")
```

```{r}
mammals <- read_csv("data/mammal_lifehistories_v2.csv")
```

To keep things tidy, I am going to rename the variables in the mammals data.  
```{r}
mammals <- rename(mammals, genus=Genus, wean_mass="wean mass", max_life="max. life", litter_size="litter size", litters_per_year="litters/year")
```

## `filter()`
Unlike `select()`, `filter()` allows us to extract data that meet specific criteria within a variable. Let's say that we are interested only in the fish that occur in lake "AL". We can use `filter()` to extract these observations.  
```{r}
filter(fish, lakeid == "AL")
```

Similarly, if we are only interested in fish with a length greater than or equal to 350 we can use `filter()` to extract these observations.  
```{r}
filter(fish, length >= 350)
```

+ `filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).  

Using the `!` operator allows for the exclusion of specific observations.
```{r}
filter(fish, lakeid != "AL")
```

## Using `filter()` with multiple observations  
Filtering multiple values within the same variable requires the `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm).    
```{r}
filter(fish, length %in% c(167, 175))
```

Alternatively, you can use `between` if you are looking for a range of specific values.  
```{r}
filter(fish, between(scalelength, 2.5, 2.55))
```

You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
filter(fish, near(radii_length_mm, 2, tol = 0.2))
```

## Practice
1. Filter the `fish` data to include the samples from lake "BO".
```{r}

```

2. Filter the data to include all lakes except "AR".
```{r}

```

3. Filter the fish data to include all fish with a scalelength within 0.25 of 8.
```{r}

```

4. Filter the fish data to include fish with a scalelength between 2 and 4.
```{r}

```

```{r}

```

## Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract only the fish that have lakeid "AL" and length >350.
```{r}
filter(fish, lakeid == "AL" & length > 350)
```

Notice that the `|` operator generates a different result.
```{r}
filter(fish, lakeid == "AL" | length > 350)
```

Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met.  
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 or condition 2 is met.  
+ `filter(xor(condition1, condition2)` will return all rows where only one of the conditions is met, and not when both conditions are met.  

In this case, we filter out the fish with a length over 400 and a scale length over 11 or a radii length over 8.
```{r}
filter(fish, length > 400, (scalelength > 11 | radii_length_mm > 8))
```

## Practice  
1. Have a look at the mammals data using the summary `functions` of your choosing.    
```{r}

```

2. What are the names of the variables in the mammals data?  
```{r}

```

3.  From the `mammals` data, filter all members of the family Bovidae with a mass greater than 450000.
```{r}

```

4. Let's say we are only interested in family, genus, species, mass, and gestation. Relimit the mammals data frame to these variables.  
```{r}

```

5. Use the output from #4 to focus on the family Bovidae with a mass greater than 450000.
```{r}

```

```{r}

```

```{r}

```

6. From the `mammals` data, build a data frame that compares `mass`, `gestation`, and `newborn` among the primate genera `Lophocebus`, `Erythrocebus`, and `Macaca`. Among these genera, which species has the smallest `newborn` mass?
```{r}

```

```{r}

```

## Wrap-up  

Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)