---
title: "Tidyr 2: `pivot_wider()`, `summarize()`, and `group_by()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Explain the difference between tidy and messy data.  
2. Evaluate a dataset as tidy or untidy.  
3. Use the `pivot_wider()` function of tidyr to transform data from long to wide format.  
4. Use `summarize()` and `group_by()` to produce statistical summaries of data.

## Group Project
Let's take 10 minutes to decide on groups and think about a theme for our projects. What data interest you? Make a plan on doing some internet searches and selecting data that will work for your group. Think about the kinds of questions you might ask.  

## Resources
- [tidyr `pivot_wider()`](https://tidyr.tidyverse.org/reference/pivot_wider.html)  
- [pivoting](https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html)  

## Review
Last time we learned the fundamentals of tidy data and used the `pivot_longer()` function to wrangle a few types of frequently encountered untidy data. In the second part of today's lab we will use the `pivot_wider()` function to deal with a second type of untidy data.  

## Load the tidyverse
```{r}
library("tidyverse")
```

## `pivot_longer()`
Recall that we use `pivot_longer()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
datasets::USPersonalExpenditure
?USPersonalExpenditure
```

Here we add a new column of expenditure types, which are stored as rownames above, with `mutate()`. The USPersonalExpenditures data also needs to be converted to a data frame before we can use the tidyverse functions, because it comes as a matrix.
```{r}
expenditures <- 
  USPersonalExpenditure %>% 
  as.data.frame() %>% 
  mutate(expenditure = rownames(USPersonalExpenditure))
expenditures
```

## Practice
1. Are these data tidy? Please use `pivot_longer()` to tidy the data. Here we use back ticks on the column names so R does not confuse them with numerical indices.

2. Restrict the data to medical and health expenditures only. Sort in ascending order.

## `pivot_wider()`
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  

Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  

```{r}
tb_data <- read_csv("data/tb_data.csv")
tb_data
```

When using `pivot_wider()` we use `names_from` to identify the variables (new column names) and `values_from` to identify the values associated with the new columns.
```{r}
tb_data %>% 
  pivot_wider(names_from = "key", #the observations under key will become new columns
              values_from = "value")
```

## Practice
1. Load the `gene_exp.csv` data as a new object `gene_exp`. Are these data tidy? Use `pivot_wider()` to tidy the data.

## `rename()`
The `rename()` function is actually part of *dplyr*, but I put it here because I think of it as part of transforming untidy data. Remember, youcan still rename columns as part of `select()`.
```{r}
tb_data %>% 
  pivot_wider(names_from = "key",
              values_from = "value") %>% 
  dplyr::rename(
    Country=country,
    Year=year,
    New_Cases=cases,
    Population=population
  )
```

## `summarize()`
`summarize()` will produce summary statistics for a given variable in a data frame. For example, in homework 2 you were asked to calculate the mean of the sleep total column for large and small mammals. We did this using a combination of tidyverse and base R commands, which isn't very efficient or clean. We can do better!  
```{r}
msleep
```

From homework 2.
```{r}
large <- 
  msleep %>% 
  select(name, genus, bodywt, sleep_total) %>% 
  filter(bodywt >= 200) %>% 
  arrange(desc(bodywt))
large
```

```{r}
mean(large$sleep_total)
```

We can accomplish the same task using the `summarize()` function in the tidyverse and make things cleaner.
```{r}
msleep %>% 
  filter(bodywt >= 200) %>%
  summarize(mean_sleep_lg = mean(sleep_total))
```

You can also combine functions to make useful summaries for multiple variables.
```{r}
msleep %>% 
    filter(bodywt >= 200) %>% 
    summarize(mean_sleep_lg = mean(sleep_total), 
              min_sleep_lg = min(sleep_total),
              max_sleep_lg = max(sleep_total),
              total = n())
```

`n_distinct()` is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 200 in body weight.
```{r}
msleep %>% 
  filter(bodywt >= 200) %>% 
  summarise(n_genera=n_distinct(genus))
```

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of vector), first() (returns first value in vector), last() (returns last value in vector) and n_distinct() (number of distinct values in vector).

## Practice
1. How many genera are represented in the msleep data frame?

2. What are the min, max, and mean body weight for all of the mammals? Be sure to include the total n.

## `group_by()`
The `summarize()` function is most useful when used in conjunction with `group_by()`. Although producing a summary of body weight for all of the mammals in the dataset is helpful, what if we were interested in body weight by feeding ecology?
```{r}
names(msleep)
```

```{r}
msleep %>%
  group_by(vore) %>% #we are grouping by feeding ecology
  summarize(min_bodywt = min(bodywt),
            max_bodywt = max(bodywt),
            mean_bodywt = mean(bodywt),
            total=n())
```

`count()` is an easy way of determining how many observations you have within a column. It effectively acts like a combination of `group_by()` and `n()`.
```{r}
msleep %>% 
  count(order, sort = T)
```

You can also use `count()` across multiple variables.
```{r}
msleep %>% 
  count(order, vore, sort = TRUE)
```

`add_count()` adds a new column to the dataframe.
```{r}
msleep %>% 
  select(name:order) %>% 
  add_count(order) %>% 
  top_n(-5) #the bottom 5 
```

## Practice
1. Calculate mean brain weight by taxonomic order in the msleep data.

2. What does `NA` mean?

3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea?

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)