---
title: "Pipes, `arrange()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use pipes to connect functions in dplyr.  
2. Use `arrange()` to order dplyr outputs.  
3. Use `mutate()` to add columns in a dataframe.  

## Review
At this point, you should be comfortable using the `select()` and `filter()` functions of `dplyr`. If you need help, please [email me](mailto: jmledford@ucdavis.edu).  

## Load the tidyverse and janitor
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
For this lab, we will use the following two datasets:  

1. Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.   [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
```{r}
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")
```

2. S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.   [link](http://esapubs.org/archive/ecol/E084/093/)  
```{r}
mammals <- read_csv("data/mammal_lifehistories_v2.csv")
```

Rename the `mammals` variables using janitor.  
```{r}
mammals <- clean_names(mammals)
```

## Pipes `%>%` 
In order to start combining `select()`, `filter()`, and other functions efficiently, we need to learn pipes. Pipes feed the output from one function into the input of another function. This helps us keep our code sequential and clean.  

In this example, we use a pipe to select only `lakeid` and `scalelength` then filter that output only for lakes "AL". Notice that we only need to call the data one time.  
```{r}
select(fish, lakeid, scalelength)
```

```{r}
filter(fish, lakeid == "AL")
```

```{r}

```

Here we select family, genus, and species then filter for gestation greater than 15 months.
```{r}

```

## Practice
1. We are interested in the fish from the lakes "AL" and "AR" with a radii length between 2 and 4. Extract this information from the `fish` data. Use pipes!  
```{r}

```

## `arrange()`
The `arrange()` command is a bit like a sort command in excel. Note that the default is ascending order.  
```{r}
fish %>% 
  select(lakeid, scalelength) %>% 
  arrange(scalelength)
```

To sort in decreasing order, wrap the variable name in `desc()`.
```{r}
fish %>% 
  select(lakeid, scalelength) %>% 
  arrange(desc(scalelength))
```

`arrange()` can be very helpful in combination with the other functions.
```{r}

```

## Practice
1. We are interested in the radii length and scalelength of fish with the id `300`. Extract this information from the `fish` data. Please sort the data by radii length.
```{r}

```

2. Let's re-do question #3 from part 1, but this time using pipes.  

From the `mammals` data, filter all members of the family `Felidae` with a gestation period less than 3 months. As part of this, remove all variables except `genus`, `family`, and `gestation`. Lastly, remove all NA's in the variable `gestation`- in this data they are represented by -999. Sort by gestation.
```{r}

```

## `mutate()`  
Mutate allows us to create a new column from existing columns in a data frame. We are doing a small introduction here and will add some additional functions later.  

Let's convert the `scalelength` variable from millimeters to centimeters.  
```{r}

```

## Practice
1. In the mammals data, let's convert the mass variable from grams to kilograms.  Don't forget to remove those pesky -999's and arrange in descending order.
```{r}

```

2. Using the mammals data, compare wean mass to gestation. Create a new column that calculates the ratio of wean mass to gestation. Arrange the data by this ratio in descending order. Don't forget to remove the -999's from wean mass and gestation.
```{r}

```

## Wrap-up  

Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)