---
title: "`filter()` on multiple conditions"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use `filter()` to extract variables of interest.  
2. Use `filter()` to extract variables of interest under multiple conditions.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
For this lab, we will use the following two datasets:  

1. Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.   [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
```{r}
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")
```

2. S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.   [link](http://esapubs.org/archive/ecol/E084/093/)  
```{r}
mammals <- read_csv("data/mammal_lifehistories_v2.csv")
```

Let's rename some of the `mammals` variables.  
```{r}
mammals <- rename(mammals, genus="Genus", wean_mass="wean mass", max_life= "max. life", litter_size="litter size", litters_per_year="litters/year")
```

Do you remember the easier way?
```{r}

```

## Review
In lab 4, we practiced extracting observations of interest. For example, we can pull out all of the fish from a specific lake.  
```{r}
glimpse(fish)
```

Let's convert `lakeid` to a factor and have a look at which states are represented in the data.  
```{r}
fish$lakeid <- as.factor(fish$lakeid)
```

```{r}

```

Now we can pull out any state of interest.  
```{r}

```

## Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract only the fish that have lakeid "AL" and length >350.
```{r}

```

Notice that the `|` operator generates a different result. Why?
```{r}

```

Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met. By default the , means &.    
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 or condition 2 is met.  
+ `filter(xor(condition1, condition2)` will return all rows where only one of the conditions is met, and not when both conditions are met.  

In this case, we filter out the fish with a length over 400 and a scale length over 11 or a radii length over 8.
```{r}

```

## Practice  
1.  From the `mammals` data, filter all members of the family Bovidae with a mass greater than 450000.
```{r}

```

2. From the `mammals` data, build a data frame that compares `mass`, `gestation`, and `newborn` among the primate genera `Lophocebus`, `Erythrocebus`, and `Macaca`. Among these genera, which species has the smallest `newborn` mass?
```{r}

```

```{r}

```

3. From the `mammals` data, filter all members of the family `Felidae` with a gestation period less than 3 months. As part of this, remove all variables except `genus`, `family`, and `gestation`. Lastly, remove all NA's in the variable `gestation`- in this data they are represented by -999.
```{r}

```

## That's it! Let's take a break and then move on to part 2!  

-->[Home](https://jmledford3115.github.io/datascibiol/)  