---
title: "Pipes, `arrange()`, `mutate()`, and `if_else()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use pipes to connect functions in dplyr.  
2. Use `arrange()` to order dplyr outputs.  
3. Use `mutate()` to add columns in a dataframe.  
4. Use `mutate()` and `if_else()` to replace values in a dataframe.  

## Review
At this point, you should be comfortable using the `select()` and `filter()` functions of `dplyr`. If you need extra help, please [email me](mailto: jmledford@ucdavis.edu).  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Load the data
For this lab, we will use the following two datasets:  

1. Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.   [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
```{r}
fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")
```

2. S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.   [link](http://esapubs.org/archive/ecol/E084/093/)  
```{r}
mammals <- readr::read_csv("data/mammal_lifehistories_v2.csv")
```

Let's rename some of the `mammals` variables.  
```{r}
mammals <- rename(mammals, genus="Genus", wean_mass="wean mass", max_life= "max. life", litter_size="litter size", litters_per_year="litters/year")
```

## Pipes `%>%` 
In order to start combining `select()`, `filter()`, and other functions efficiently, we need to learn pipes. Pipes feed the output from one function into the input of another function. This helps us keep our code sequential and clean.  

In this example, we use a pipe to select only `lakeid` and `scalelength` then filter that output only for lakes "AL". Notice that we only need to call the data one time.  
```{r}
select(fish, lakeid, scalelength)
```

```{r}
filter(fish, lakeid == "AL")
```

```{r}
fish %>%
  select(lakeid, scalelength) %>%
  filter(lakeid == "AL")
```

Here we select family, genus, and species then filter for gestation greater than 15 months.
```{r}
mammals %>% 
  select(family, genus, species, gestation) %>% 
  filter(gestation>=15)
```

## Practice
1. We are interested in the fish from the lakes "AL" and "AR" with a radii length between 2 and 4. Extract this information from the `fish` data. Use pipes!  
```{r}

```

## `arrange()`
The `arrange()` command is a bit like a sort command in excel. Note that the default is ascending order.  
```{r}
fish %>% 
  select(lakeid, scalelength) %>% 
  arrange(scalelength)
```

To sort in decreasing order, wrap the variable name in `desc()`.
```{r}
fish %>% 
  select(lakeid, scalelength) %>% 
  arrange(desc(scalelength))
```

`arrange()` can be very helpful in combination with the other functions.
```{r}
fish %>% 
  select(lakeid, length, fish_id, scalelength) %>% 
  filter(lakeid=="AL") %>% 
  arrange(fish_id)
```

## Practice
1. We are interested in the radii length and scalelength of fish with the id `300`. Extract this information from the `fish` data. Please sort the data by radii length.
```{r}

```

## `mutate()`  
Mutate allows us to create a new column from existing columns in a data frame. We are doing a small introduction here and will add some additional functions later. Let's convert the length variable from cm to millimeters and create a new variable called length_mm.  
```{r}
fish %>% 
  mutate(length_mm = length*10) %>% 
  select(fish_id, length, length_mm)
```

## `mutate_all()`
This last function is super helpful when cleaning data. With "wild" data, there are often mixed entries (upper and lowercase), blank spaces, odd characters, etc. These all need to be dealt with before analysis.  

Here is an example that changes all entries to lowercase (if present).  
```{r}
mammals %>%
  mutate_all(tolower)
```

Using the across function we can specify individual columns.
```{r}
mammals %>% 
  mutate(across(c("order", "family"), tolower))
```

## `if_else()`
We will briefly introduce `if_else()` here because it allows us to use `mutate()` but not have the entire column affected in the same way. In a sense, this can function like find and replace in a spreadsheet program. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  

Have a look at the data from mammals below. Notice that the values for newborn include `-999.00`. This is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.  
```{r}
mammals %>% 
  select(genus, species, newborn) %>% 
  arrange(newborn)
```

```{r}
mammals %>% 
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>% 
  arrange(newborn)
```

## Practice
1. Use `mutate()` to make a new column that is the half length of each fish: length_half = length/2. Select only fish_id, length, and length_half.
```{r}

```

2. We are interested in the family, genus, species and max life variables. Because the max life span for several mammals is unknown, the authors have use -999 in place of NA. Replace all of these values with NA in a new column titled `max_life_new`. Finally, sort the date in descending order by max_life_new. Which mammal has the oldest known life span?
```{r}

```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  