---
title: "Dealing with NA's"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce summaries of the number of NA's in a data set.  
2. Replace values with `NA` in a data set as appropriate.  

## Review
In the last section we practiced wrangling untidy data using `tidyr`. We also learned the `summarize()` function and `group_by()` to produce useful summaries of our data. But, we ended the last session with the discovery of NA's and how they can impact analyses. This is a big issue in data science and we will spend the remainder of this lab learning how to manage data with missing values.  

## Install `naniar` and `skimr`
```{r}
#install.packages("naniar")
#install.packages("skimr")
```

## Load the libraries
```{r}
library("tidyverse")
library("naniar")
library("skimr")
```

## Dealing with NA's
In almost all scientific data, there are missing observations. These can be tricky to deal with, partly because you first need to determine how missing values were treated in the original study. Scientists use different conventions in showing missing data; some use blank spaces, others use "-", etc. Unfourtunately, sometimes **missing data is indicated with numerics like -999.0 or zero!**, though this can be required in some situations (for example raster data). Often, a combination of methods are used. It is up to the data analyst to find out how missing values are represented in the data set and to deal with them appropriately. 

## Practice
For the following, we will use life history data for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from:  
**S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.**  

1. Load the mammals life history data. Do you see any NA's?
```{r}
life_history <- 
  readr::read_csv("data/mammal_lifehistories_v2.csv")
life_history
```

2. What are the possible problems if missing data are indicated by "-999.0" or "0"?


## For the following examples, we will use the built-in `msleep` data
```{r}
?msleep
glimpse(msleep)
```

## Are there any NA's?
Let's first check to see if our data has any NA's. `is.na()` is a function that determines whether a value in a data frame is or is not an NA. This is evaluated logically as TRUE or FALSE.  

Keep in mind, these are the values that R is interpreting as NA. If the authors of the data represented NA's in another way, they will not be detected by R.
```{r}
is.na(msleep)
```

OK, what are we supposed to do with that? Unless you have a small data frame, applying the `is.na()` function is not helpful but we can use the code in another way. Let's incorporate it into the `summarize()` function.
```{r}
msleep %>% 
  summarize(number_nas = sum(is.na(msleep)))
```

This is better, but we still don't have any idea of where those NA's are in our data frame. If there were a systemic problem in the data it would be hard to determine. In order to do this, we need to apply `is.na()` to each variable of interest.
```{r}
msleep %>% 
  summarize(number_nas = sum(is.na(conservation)))
```

The `summary()` function in base R will tally NAs by variable, but only for categorical/factor, numeric, integer, or logical columns. What about any NA values in the character columns?
```{r}
summary(msleep)
```

## Practice
1. Load the data `import_export.csv` and store it as a new object `import_export`.


2. Are there NAs in the data? Evaluate this logically (true/false).


3. How many total NAs are present?


## NAs in large datasets using `purr`
What if we are working with hundreds or thousands (or more!) variables?! In order to deal with this problem efficiently we can use another package in the tidyverse called `purrr`.  

This handy bit of code generates a new data frame that sums all NA's accross variables.
```{r}
msleep %>%
  purrr::map_df(~ sum(is.na(.))) #map to a new data frame the sum results of the is.na function for all columns
```

Don't forget that we can use `gather()` to make viewing this output easier.
```{r}
msleep %>% 
  purrr::map_df(~ sum(is.na(.))) %>% 
  pivot_longer(everything(),
    names_to= "variables",
    values_to = "num_nas") %>% 
  arrange(desc(num_nas))
```

This is much better, but we need to be careful. R can have difficulty interpreting missing data. This is especially true for categorical variables. Always do a reality check if the output doesn't make sense to you. A quick check never hurts and can save you from wasting time in the future or from drawing false conclusions.  

You can explore a specific variable more intently using `count()`. This operates similarly to `group_by()`.
```{r}
msleep %>% 
  count(conservation)
```

Adding the `sort=TRUE` option automatically makes a descending list. We confirm that there are 29 NA's within the conservation variable.
```{r}
msleep %>% 
  count(conservation, sort = TRUE)
```

It is true that all of this is redundant, but you want to be able to run multiple checks on the data. Remember, just because your code runs without errors doesn't mean it is doing what you intended.  

## Replacing NAs with `na_if()` and `mutate()`
The bit of code below is very handy, especially if the data has NAs represented as **actual values that you want replaced** or if you want to make sure any blanks are treated as NA. You can also use it to treat specific values as NA.  

Let's say we know that NA's are represented by "-999". We can replace this value with NA accross the data frame.
```{r}
msleep %>% 
  na_if("-999")
```

Let's say we want to treat "domesticated" as NA in the conservation column. For this, we use `mutate()`.
```{r}
msleep %>%  
  mutate(conservation = na_if(conservation, "domesticated"))%>% 
  count(conservation, sort = TRUE) #previously there were only 29 NAs
```

## Practice
For the following examples, we will go back to the `life_history` data. First, let's rename some of the variables.
```{r}
life_history <- 
  life_history %>% 
  dplyr::rename(
          genus        = Genus,
          wean_mass    = `wean mass`,
          max_life     = `max. life`,
          litter_size  = `litter size`,
          litters_yr   = `litters/year`
          )
```

1. Explore the data using any method that you prefer. Below, I show a package called `skimr`. If you want to try this, make sure that it is installed and loaded.  

`skimr` provides a nice output with histograms for numeric variables.
```{r}
life_history %>% 
  skimr::skim()
```

2. Reconsider the practice question from the beginning of the lab. Are there any NA's in the data? Does this seem likely? Do the mean values from the `skim()` output above make sense?  


3. Replace the entries that represent NA's in the data using `na_if()`. Double check your work and compare to #2 above.
```{r}
life_history2 <- 
  life_history %>% 
  na_if("-999")
```

4. In which variable are the most NAs in the data frame?


## `naniar`
`naniar` is a package that is built to manage NAs. Many of the functions it performs can be performed using tidyverse functions above, but it does provide some interesting options.

`miss_var_summary` provides a clean summary of NAs across the dataframe.
```{r}
naniar::miss_var_summary(life_history2)
```

We can also use `miss_var_summary` with `group_by()`.
```{r}
life_history2 %>%
  group_by(order) %>%
  select(order, wean_mass) %>% 
  naniar::miss_var_summary(order=T)
```

We can also replace -999 with NA as before. This shows replacement within the newborn column.
```{r}
life_history %>% 
  naniar::replace_with_na(replace = list(newborn = -999))
```

This shows replacement within the newborn and weaning columns.
```{r}
life_history %>% 
  naniar::replace_with_na(replace = list(newborn = -999, weaning= -999))
```

## Visualizing NAs
There is another package `visdat` that can be used to visualize the proportion of different classes of data, including missing data.
```{r}
#install.packages("vidat)
library(visdat)
```

```{r}
vis_dat(life_history) #classes of data
```

```{r}
vis_miss(life_history)
```

## Dealing with NA's in advance
If you know how NA's are treated in the data, then you can deal with them in advance using `na()` as part of the `readr` package. But, you need to be sure about the analyses you want to conduct and how NA's might be interpreted.  
```{r}
life_history3 <- 
  readr::read_csv(file = "data/mammal_lifehistories_v2.csv", 
                  na = c("NA", " ", ".", "-999")) #all NA, blank spaces, .,and -999 are treated as NA
```

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)