---
title: "Joins and Lubridate"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*  
1. Join data frames in R.   
2. Format dates in a data frame. 

## Load the libraries
```{r, message = FALSE}
library("tidyverse")
library("janitor")
library("lubridate") #this will help us manage dates
```

## Data
For this tutorial, we will use data from my research on _Usofila_ spiders. These spiders live mostly in caves and the data were collected by building a database of all the known museum specimens in the US.   
```{r message=FALSE, warning=FALSE}
spider_records <- read_csv("data/spider_records.csv")
```

```{r}
glimpse(spider_records) #let's have a quick look
```

## Joining Files
Sometimes data we need is stored in a separate file or becomes available later and we need to join it to our existing data in order to work with it. Let's use an easy example to explore some of the different types of joins. This section was based on the excellent example provided by [Holly Emblem](https://hollyemblem.medium.com/joining-data-with-dplyr-in-r-874698eb8898).  

Let's assume we have a database of customers and their purchases.
```{r}
table_A <- read_csv("data/table_A.csv")
table_B <- read_csv("data/table_B.csv")
```

```{r}
table_A
```

```{r}
table_B
```

The general syntax for joins is...
```{r}
#join_type(firstTable, secondTable, by=columnTojoinOn)
```

`inner_join`  
Where did customers 3 and 4 go? Since they are not on both tables, they were dropped.
```{r}
inner_join(table_A, table_B, by="customer_ID")
```

`left_join`
Where did customer 4 go? Notice the missing data for customer 3.
```{r}
left_join(table_A, table_B, by="customer_ID")
```

`right_join`
Where did customer 3 go? Notice the missing data for customer 4.
```{r}
right_join(table_A, table_B, by="customer_ID")
```

`full_join`
```{r}
full_join(table_A, table_B, by="customer_ID")
```

`anti_join`
Provides the rows in the first table for which there are not matching values in the second table.
```{r}
anti_join(table_A, table_B, by="customer_ID")
```

## Joining the spider data
Here, the latitude and longitude for each spider was recorded from the records at a later date, and now we need to join it to our `spiders_records` data frame. The lat/long were recorded into one single file for each observation. Let's read in the lat/long data.
```{r, message = FALSE}
spiders_locs <- read_csv("data/spiders_locations.csv")
```

We will use a join here to merge lat/long to our data frame. Both files contain a unique identifier called `Accession` which we will use to join.  

## Practice
1. Which type of join is most appropriate in this case?  
```{r}
spiders_with_locs <- 
  left_join(spider_records, spiders_locs, by = "Accession")
```

```{r}
names(spiders_with_locs) #make sure it worked
```

As a side note, joining data can highlight problems or typos with the data when the join does not go as expected.  

## Formatting Dates
Did you notice the date column? It seems to be in the format Day/Month/Year and is of class character. We need to change this to a date format that R will recognize.
```{r}
glimpse(spiders_with_locs)
```

The `lubridate` package was created to deal with dates of all types. There are many useful functions in `lubridate` for working with dates. Please see the excellent tutorial provided by [RichardOnData](https://www.youtube.com/watch?v=VYAo69WdJZg&ab_channel=RichardOnData).
```{r}
day <- today()
day
```

```{r}
str(day)
```

```{r}
datetime <- now()
datetime
```

We want to change our date column to the standard "YEAR-MO-DA" format that R will recognize as a date. Right now it is listed as day-month-year. We can use the lubridate function `dmy` for this. 
```{r}
spiders_with_locs <- spiders_with_locs %>% 
  mutate(Date = dmy(Date))
```

There are many different options for working with dates and datetimes in lubridate including `ymd`, `mdy`, `dmy`, `mdy_hms`, `ymd_hms`.
```{r}
dateformat1 <- "20200922"
dateformat2 <- "09-22-2020"
dateformat3 <- "22/09/2020"
dateformat4 <- "09-22-2020 17:00:00"
dateformat5 <- "20200922 170000"
```

## Practice 
1. Convert each of the examples above using lubridate.
```{r}
ymd(dateformat1)
```

```{r}
mdy(dateformat2)
```

```{r}
dmy(dateformat3)
```

```{r}
mdy_hms(dateformat4)
```

```{r}
ymd_hms(dateformat5)
```

Let's save our final data to a `.csv`.
```{r}
write.csv(spiders_with_locs, file = "spiders_with_locs.csv", row.names = FALSE)
```

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)