---
title: "Intro to Spatial Data"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*  
1. Produce distribution maps in R.  
 
## Resources
[Overview of Cooordinate Reference Systems in R](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf)

## Spatial Data in R
There are many packages and techniques for working with spatial data in R. We will cover just some of the basics. One nice package is `ggmap`, which allows us to get base maps from Google Maps, OpenStreetMap, and Stamen Maps. It also works well with `ggplot2`.  

If you completed part 1 of today's lab then you should have the following packages installed. If not, then you should get them going now. You need to do these in order!  

`devtools` and `ggmap`
```{r}
#install.packages("devtools")
library(devtools)
#devtools::install_github("dkahle/ggmap")
```

## Load the libraries
```{r}
library(tidyverse)
library(here)
library(janitor)
```

## Load the `ggmap` package  
```{r}
library(ggmap)
```

Let's load our processed data from the first part of the lab.
```{r}
spiders <- read_csv("data/spiders_with_locs.csv")%>% clean_names()
```

There is an error in one of the coordinates that we will fix here.
```{r}
spiders <- spiders %>% filter(latitude<=42)
```

## Create Base Map
Our goal here is to plot the spiders locations from the columns which contain the latitude and longitude. First, we need to get a base map for plotting our points on. We could plot them without a base map, but that wouldn't give us any context as to where they are in space. To get a base map we specify a min and max of each x and y coordinate, and create a bounding box.  

We set the bounding box to a little outside our min and max locations with `f = 0.05`.    

`summary()` gives us our min and max.
```{r}
spiders %>% 
  select(latitude, longitude) %>% 
  summary()
```

Now we set the bounding box. We use the min and max values for latitude and longitude to set the range.  
```{r}
lat <- c(34.67, 41.80)
long <- c(-124.1, -115.5)
bbox <- make_bbox(long, lat, f = 0.05)
```

Let's get a base map for our bounding box area. We will use the stamen maps because they are free. There are several different map types, including: `terrain-labels`, `terrain-lines`, `toner`, `toner-2011`, `toner-background`, `toner-hybrid`, `toner-lines`, `toner-lite`, and `watercolor.`
```{r}
map1 <- get_map(bbox, maptype = "terrain", source = "stamen")
```

```{r}
ggmap(map1)
```

## Adding Points to Base Map
`ggmap` works well with `ggplot2`. To add our points we only need to specify the x and y location similar to how we made charts in previous labs. 
```{r}
ggmap(map1) + 
  geom_point(data = spiders, aes(longitude, latitude)) +
  labs(x= "Longitude", y= "Latitude", title="Spider Locations")
```

## Practice
Let's practice with a data set that records sightings of grizzly bears (Ursos arctos) in Alaska [data set](https://rcweb.dartmouth.edu/~f002d69/workshops/index_rspatial.html).  
```{r}
grizzly <- read_csv("data/bear-sightings.csv")
```

1. Use the range of the latitude and longitude to build an appropriate bounding box for your map.
```{r}

```

2. Load a map from `stamen` in a terrain style projection and display the map.
```{r}

```

```{r}

```

3. Build a final map that overlays the recorded observations of grizzly bears in Alaska.
```{r}

```

## Wrap-up  
Please review the learning goals and be sure to use the code here as a reference when completing the homework.

-->[Home](https://jmledford3115.github.io/datascibiol/)