---
title: "Joins and Lubridate"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Learning Goals
*At the end of this exercise, you will be able to:*  
1. List and join multiple files from a directory.  
2. Format dates in a data frame. 

## Listing Files in Folder
Often, the data we need is split up into multiple files, either by some geographical variable such as county, or by a time constituent such as months or weeks. The best method to deal with data in multiple files will be quick and also reproducible. We want to avoid doing anything by hand (besides very minor editing) to the original files someone sends us. We want everything to be reproducible so we can easily track any problems or errors that may crop up and prevent us from making careless errors ourselves.  

So copying/pasting multiple files together isn't an option for us. Let's see how we can do this in R. In the `data/spiders` folder there are 32 files, each named for a county in California. Each file contains a record of an observation of a species of cave spider. Each observation contains a unique ID for each spider as well as some other important information such as date and location name. Let's use R to list all the `.csv` files in the `spiders` folder.  

But first, load the tidyverse and the other packages we have been using.
```{r, message = FALSE}
library("tidyverse")
library("janitor")
library("lubridate") #this will help us manage dates
```

Now we can list all of the files in a directory. This is very helpful when you are working with multiple files at once. We can also get the full path names with `full.names = TRUE`.
```{r}
files <- list.files(path = "data/spiders", pattern = ".csv", full.names = TRUE)
files
```

Now we want to read each of these files into R without having to do them one at a time because there are quite a few. There are several ways to do this, but a quick and straightforward way is to import them as a list. A list in R is an object which can store multiple other objects of the same or differing types. Lists are common in R so it's useful to be comfortable with them.  

Let's import our .csv files into a list. The `lapply()` function is a part of the `apply` family of functions. It will iterate over elements of an object, apply a function we specify, and return a list. We have a character vector of file paths, so we want to iterate over all the path names and apply `read_csv()`.  
```{r, message = FALSE}
spider_list <- lapply(files, read_csv)
```

We can view elements in our list with double brackets. Let's view the data for Butte county.
```{r}
spider_list[[3]]
```

## Practice
1. Getting an idea of the structure of lists is a bit tricky, especially when there are many elements. Try exploring the elements in `spider_list` using names(), str(), or glimpse(). For names(), you will need to specify an element using [[]].  
```{r}

```

## Naming List Elements
We don't need to here, but for demonstration purposes we can name the elements in our list.  We want to keep the first element of each list element.
```{r}
names <- list.files(path = "data/spiders", pattern = ".csv")
names
```

We will first get the names of each file, but we only want the county part. We will use `strsplit()` for that, which creates a nested list of strings.
```{r}
names_list <- strsplit(names, split = " .csv")
names_list
```

We can then use `unlist()` to get a vector of the names.
```{r}
names_vec <- unlist(names_list)
names_vec
```

Now we can name the elements in our list.
```{r}
names(spider_list) <- names_vec
names(spider_list)
```

## Practice
1. Now that our list elements are named, how could we access the Butte County data by name?
```{r}

```

## Merging Files
We are fortunate here in that all of our data frames have the same column names. This makes it easy to merge the data together with `bind_rows()` from `dplyr`. `bind_rows()` matches columns by name.
```{r}
spiders_all <- bind_rows(spider_list)
spiders_all
```

## Joining Files
Sometimes data we need is stored in a separate file or becomes available later and we need to join it to our existing data in order to work with it. Let's use an easy example to explore some of the different types of joins. This section was based on the excellent example provided by [Holly Emblem](https://hollyemblem.medium.com/joining-data-with-dplyr-in-r-874698eb8898).  

Let's assume we have a database of customers and their purchases.
```{r}
table_A <- read_csv("data/table_A.csv")
table_B <- read_csv("data/table_B.csv")
```

```{r}
table_A
```

```{r}
table_B
```

The general syntax for joins is...
```{r}
#join_type(firstTable, secondTable, by=columnTojoinOn)
```

`inner_join`  
Where did customers 3 and 4 go?
```{r}
inner_exampleDF <- inner_join(table_A, table_B, by="customer_ID")
inner_exampleDF
```

`left_join`
Where did customer 4 go? Notice the missing data for customer 3.
```{r}
left_exampleDF <- left_join(table_A, table_B, by="customer_ID")
left_exampleDF
```

`right_join`
Where did customer 3 go? Notice the missing data for customer 4.
```{r}
right_exampleDF <- right_join(table_A, table_B, by="customer_ID")
right_exampleDF
```

`full_join`
```{r}
full_exampleDF <- full_join(table_A, table_B, by="customer_ID")
full_exampleDF
```

`anti_join`
Provides the rows in the first table for which there are not matching values in the second table.
```{r}
anti_exampleDF <- anti_join(table_A, table_B, by="customer_ID")
anti_exampleDF
```

## Joining the spider data
Here, the latitude and longitude for each spider were recorded from the field records at a later date, and now we need to join it to our `spiders_all` data frame. The lat/long were recorded into one single file for each observation. Let's read in the lat/long data.
```{r, message = FALSE}
spiders_locs <- read_csv("data/spiders locations/spiders_locations.csv")
```

We will use a join here to merge lat/long to our data frame. Both files contain a unique identifier called `Accession` which we will use to join.  

## Practice
1. Which type of join is most appropriate in this case?  
```{r}

```

As a side note, joining data can highlight problems or typos with the data when the join does not go as expected.  

## Formatting Dates
We now have a single data frame with all of our spider data including locations. That was a lot of work, even with R. But remember, now we have a reproducible work flow starting from the original files. This work flow serves as a record of what we did, keeps the original files untouched, and can make it easier to track down problems later in our analysis. Most importantly, everyone who uses our R script will know exactly what we did!  

There is one last thing to change. Did you notice the date column? It seems to be in the format Day/Month/Year. 
```{r}
#class(spiders_with_locs$Date)
#glimpse(spiders_with_locs)
```

The `lubridate` package was specifically created to deal with dates of all types. There are many useful functions in `lubridate` for working with dates. Please see the excellent tutorial provided by [RichardOnData](https://www.youtube.com/watch?v=VYAo69WdJZg&ab_channel=RichardOnData).
```{r}
day <- today()
day
```

```{r}
str(day)
```

```{r}
datetime <- now()
datetime
```

We want to change our date column to the standard "YEAR-MO-DA" format that R will recognize as a date. Right now it is listed as day-month-year. We can use the base R function `as.Date()` or the lubridate function `dmy` for this. 
```{r}
#dmy(spiders_with_locs$Date)
```

There are many different options for working with dates and datetimes in lubridate including `ymd`, `mdy`, `dmy`, `mdy_hms`, `ymd_hms`.
```{r}
dateformat1 <- "20200922"
dateformat2 <- "09-22-2020"
dateformat3 <- "22/09/2020"
dateformat4 <- "09-22-2020 17:00:00"
dateformat5 <- "20200922 170000"
```

## Practice 
1. Convert each of the examples above using lubridate.
```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

## Export for part 2  
We need to save our final data to a `.csv` so we can use it in part 2. 
```{r}
#write.csv(spiders_with_locs, file = "spiders_with_locs.csv", row.names = FALSE)
```

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)