---
title: "Lab 2.1"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Define an object in R.  
2. Use objects to perform calculations.  
3. Explain the difference between data classes in R.  
4. Use R to identify the class of specific data.  
5. Define NA in R.  
6. Determine whether or not data have NA values.  

## Objects
In order to access the potential of R we need to assign values or other types of data to `objects`. There is a specific format that I want you to follow, so please pay close attention.  

Assign a value to object 'x'. The '<-' symbol is read as 'gets'. In this case, x gets 42. Make sure that you are in the environment panel and you should see the value associated with 'x'. On a mac, you can push `option` and `-` to automatically generate the gets symbol.  
```{r}

```

To print the object to the screen, just type x.  
```{r}

```

Assign a value of 30 to a new object y.
```{r}

```

The = symbol works, but is not a convention followed by most programmers.
```{r}

```

Once objects have been created, you can do things with them.  
```{r}

```

```{r}

```

Make two new objects, treatment and control. The value of treatment is 36 and the value of control is 38.
```{r}

```

What is the sum of treatment and control?
```{r}

```

Here we make a new object `my_experiment` that is the sum of the treatment and control. Notice that I use `_` and not spaces. 
```{r}

```

We can also use the function sum to do the same thing. Notice that if I give a new object the same name as an existing object, the old one is replaced.
```{r}

```

To learn more about the sum function, use the help command `?`.
```{r}

```

## Nomenclature
We need to be careful about nomenclature when we write code. R allows us to give almost any name we want to an object, but there are exceptions. For example, we don't want to give a name to an object that is the same as a function in R.  
```{r}
#else <- 12
```

We get an error here because `else` is a function in R. You also don't want to give names that might get confused with functions; i.e. you can assign a value to 'mean' but this could become confusing because mean is also a function.  
```{r}

```

What is the mean of 2+8+2? (Do the math in your head first). Does the following code match your prediction?
```{r}
2+8+2/3
```

Order of operations applies, so we need to tell R exactly what we want.
```{r}

```

Here we use the mean function. Notice that we use `c` which stands for concatenate. More on this later.  
```{r}

```

Why does this not work?
```{r}

```

## Practice
1. Create three new objects, `venom_GT`, `chiron`, and `veyron`. These are the fastest cars in the world. Assign each car to its top speed. The venom_GT can go 270, chiron is 261, and veyron is 268. 
```{r}

```

2. Use arithmetic to calculate the mean top speed for the cars.
```{r}

```

3. Use the function `mean()` to calculate the mean top speed for the cars.
```{r}

```

## Types of Data
There are four frequently used `classes` of data: 1. numeric, 2. integer, 3. character, 4. logical.
```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe"
my_logical <- FALSE
```

To find out what type of data you are working with, use the `class()` function. This is important because sometimes we will need to change the type of data to perform certain analyses.
```{r}

```

```{r}

```

You can use the `is()` and `as()` functions to clarify or specify a type of data.
```{r}

```

Let's convert my_numeric to an integer.
```{r}

```

```{r}

```

## Missing Data
R has a special way to designate missing data, the NA. NA values in R have specific properties which are very useful if your data contains any missing values. Later this quarter we will have a lab focused on dealing with NAs.  

NA values are used to designate missing data. `is.na` or `anyNA` are useful functions when dealing with NAs in data. 
```{r}

```

```{r}

```

```{r}

```

## Practice  
1. Let's create a vector that includes some missing data (we will discuss vectors more in part 2). For now, run the following code chunk.  
```{r}

```

2. Calculate the mean of `new_vector`.
```{r}

```

3. How do you interpret this result? What does this mean about NAs? 

4. Recalculate the mean using the following code chunk. Why is the useful?  
```{r}

```

-->[Home](https://jmledford3115.github.io/datascibiol/)  