---
title: "Vectors and Data Matrices"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Define data structures and provide examples of vectors and data matrices.  
2. Build a new vector and call elements within it.  
3. Combine a series of vectors into a data matrix.  
4. Name columns and rows in a data matrix.  
5. Select values and use summary functions in a data matrix.  

## Data Structures
In addition to classes of data, R also organizes data in different ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. For now, we will focus on `vectors`.  

## Vectors
Vectors are a common way of organizing data in R.  We create vectors using the `c` command. The `c` stands for concatenate.  

A numeric vector.
```{r}
my_vector <- c(10, 20, 30)
```

A character vector. Characters always have quotes and may be referred to as "strings".
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thrusday", "Friday", "Saturday", "Sunday")
```

A convenient trick for creating a vector is to generate a sequence of numbers.
```{r}
my_vector_sequence <- c(1:100)
my_vector_sequence
```

## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.
```{r}
days_of_the_week[3]
```

## Practice
Work through these examples and make adjustments to the values to experiment.

1. Evaluate all numerics in `my_vector_sequence` to determine which has a value of 15.  


2. We can use operators such as <, >, ==, <==, etc. Show all values in `my_vector_sequence` that are less than 10.  How about less than or equal to 10?  


3. If you use `[]` then you only get the values, not the logical evaluation of the entire vector. Experiment with this by adjusting the chunk below.    


## Data Matrices
Data matrices are a series of stacked vectors, similar to a data table. In the example below, we build a new data matrix using the matrix command.  

Box office earnings for Harry Potter movies (in millions!). Notice that these are separate vectors.  
```{r}
Philosophers_Stone <- c(317.5, 657.1)
Chamber_of_Secrets <- c(261.9, 616.9)
Prisoner_of_Azkaban <- c(249.5, 547.1)
Goblet_of_Fire <- c(290.0, 606.8)
Order_of_the_Phoenix <- c(292.0, 647.8)
Half_Blood_Prince <- c(301.9, 632.4)
Deathly_Hallows_1 <- c(295.9, 664.3)
Deathly_Hallows_2 <- c(381.0, 960.5)
```

Create a new object called `box_office`. Here we are using the `c` command to combine the vectors into one.
```{r}
box_office <- c(Philosophers_Stone, Chamber_of_Secrets, Prisoner_of_Azkaban, Goblet_of_Fire, Order_of_the_Phoenix, Half_Blood_Prince, Deathly_Hallows_1, Deathly_Hallows_2)
box_office
```

Create `harry_potter_matrix` using the `matrix()` command. We need to tell R how to organize the `box_office` vector using the `nrow` and `byrow` commands.
```{r}
harry_potter_matrix <- matrix(box_office, nrow = 8, byrow = T)
harry_potter_matrix
```
## Name the rows and columns
Vectors `region` and `titles`, used for naming.
```{r}
region <- c("US", "non-US")
region
```

```{r}
titles <- c("Philosophers_Stone", "Chamber_of_Secrets", "Prisoner_of_Azkaban", "Goblet_of_Fire", "Order_of_the_Phoenix", "Half_Blood_Prince", "Deathly_Hallows_1", "Deathly_Hallows_2")
titles
```

Name the columns using `colnames()` with the vector region.
```{r}
colnames(harry_potter_matrix) <- region
```

Name the rows using `rownames()` with the vector titles.
```{r}
rownames(harry_potter_matrix) <- titles
```

Print `harry_potter_matrix`.
```{r}
harry_potter_matrix
```

## Using a data matrix
Once you have a data matrix, you can perform lots of different analyses. For example, you can calculate the total earnings of each movie.
```{r}
global <- rowSums(harry_potter_matrix)
global
```

And even add a new column to reflect this calculation. `cbind()` adds columns.
```{r}
all_harry_potter_matrix <- cbind(harry_potter_matrix, global)
all_harry_potter_matrix
```

## Practice
1. What are the total earnings for the US and non-US regions?  


2. Add this information to the data matrix. Hint: you are adding a row, not a column.  


## A few Extras
The same methods of selecting elements in a vector apply to data matrices. We use `[]`. The following selects the value in the first column, second row.
```{r}
harry_potter_matrix[2,1]
```

Adding a colon `:` selects the specified elements in a column.  
```{r}
harry_potter_matrix[1:4]
```

We can also select values in an entire row or column. This can be useful for calculations. Here we calculate the mean of the entire second column. 
```{r}
non_us_earnings <- all_harry_potter_matrix[ ,2]
mean(non_us_earnings)
```

## Wrap-up
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)