---
title: "Data Structures: Vectors & Data Matrices"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Define data structures and provide examples of vectors, lists, matrices, and data frames.  
2. Build a new vector and call elements within it.  
3. Combine a series of vectors into a data matrix.  
4. Name columns and rows in a data matrix.  
5. Select values and use summary functions in a data matrix.  
6. Explain the difference between a data matrix and a data frame.  

## Setup
At the beginning of each class, please make sure that you complete the following:  
1. Login to the lab computer.  
2. Open Github Desktop (it should be pre-installed) and login.  
3. Clone your repository to the desktop or somewhere convenient to access.  
4. Navigate to github.com and login.  
5. Go to the class repository `BIS15-W20-DataScienceBiologists`, click the `Clone or download` button then `Download zip` to the desktop.  
6. Copy the files for today's lab from the class repository and paste them into **your** repository.  
7. Open today's lab in RStudio.  

## Data Structures
In addition to classes of data, R also organizes data in different ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. For now, we will focus on `vectors`.  

## Vectors
Vectors are a common way of organizing data in R.  We create vectors using the `c` command.  

A numeric vector.
```{r}
my_vector <- c(10, 20, 30)
```

A character vector. Characters always have quotes and may be referred to as "strings".
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thrusday", "Friday", "Saturday", "Sunday")
```

What if our numeric vector above had a missing value?
```{r}
my_vector <- c(10, NA, 30)
anyNA(my_vector)
```

Certain functions require us to specify how we want to deal with NAs. Often if NA is included in an operation, the result is NA.
```{r}
mean(my_vector)
```

Let's remove the NA value.
```{r}
mean(my_vector, na.rm = TRUE)
```

A convenient trick for creating a vector is to generate a sequence of numbers.
```{r}
my_vector_sequence <- c(1:10)
my_vector_sequence
```

`seq` allows more control.
```{r}
my_vector_sequence2 <- seq(1, 4, by=0.25)
my_vector_sequence2
```

## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.
```{r}
days_of_the_week[3]
```

## Practice
Work through these examples and make adjustments to the values to experiment.

1. Create a new numeric vector of values 1 through 100
```{r}
new_vector <- c(1:100)
```

2. We can evaluate all numerics in our vector to determine which has a value of 15.
```{r}
new_vector == 15
```

3. We can also use operators such as <, >, ==, <==, etc.
```{r}
new_vector < 10
```

4. If you use `[]` then you only get the values, not the logical evaluation of the entire vector.
```{r}
new_vector[new_vector <= 10]
```

## Data Matrices
Data matrices are a series of stacked vectors, similar to a data table. In the example below, we will build a new data matrix using the matrix command.  

Box office earnings for Harry Potter movies (in millions!). Notice that these are separate vectors.
```{r}
Philosophers_Stone <- c(317.5, 657.1)
Chamber_of_Secrets <- c(261.9, 616.9)
Prisoner_of_Azkaban <- c(249.5, 547.1)
Goblet_of_Fire <- c(290.0, 606.8)
Order_of_the_Phoenix <- c(292.0, 647.8)
Half_Blood_Prince <- c(301.9, 632.4)
Deathly_Hallows_1 <- c(295.9, 664.3)
Deathly_Hallows_2 <- c(381.0, 960.5)
```

Create a new object called `box_office`. Here we are using the `c` command to combine the vectors into one.
```{r}
box_office <- c(Philosophers_Stone, Chamber_of_Secrets, Prisoner_of_Azkaban, Goblet_of_Fire, Order_of_the_Phoenix, Half_Blood_Prince, Deathly_Hallows_1, Deathly_Hallows_2)
box_office
```

Create `harry_potter_matrix` using the `matrix()` command. We need to tell R how to organize the `box_office` vector using the `nrow` and `byrow` commands.
```{r}
harry_potter_matrix <- matrix(box_office, nrow = 8, byrow = T)
harry_potter_matrix
```

## Name the rows and columns
Vectors `region` and `titles`, used for naming.
```{r}
region <- c("US", "non-US")
titles <- c("Philosophers_Stone", "Chamber_of_Secrets", "Prisoner_of_Azkaban", "Goblet_of_Fire", "Order_of_the_Phoenix", "Half_Blood_Prince", "Deathly_Hallows_1", "Deathly_Hallows_2")
```

Name the columns using `colnames()` with the vector region.
```{r}
colnames(harry_potter_matrix) <- region
```

Name the rows using `rownames()` with the vector titles.
```{r}
rownames(harry_potter_matrix) <- titles
```

Print `harry_potter_matrix`.
```{r}
harry_potter_matrix
```

## Using a data matrix
Once you have a data matrix, you can perform lots of different analyses. For example, you can calculate the total earnings of each movie.
```{r}
global <- rowSums(harry_potter_matrix)
global
```

And even add a new column to reflect this calculation. `cbind()` adds columns.
```{r}
all_harry_potter_matrix <- cbind(harry_potter_matrix, global)
all_harry_potter_matrix
```

## Practice
1. What are the total earnings for each of the regions?  
```{r}
total_earnings <- colSums(all_harry_potter_matrix)
total_earnings
```

2. Add this information to the data matrix. Hint: you are adding a row, not a column.  
```{r}
final_harry_potter_matrix <- rbind(all_harry_potter_matrix, total_earnings)
final_harry_potter_matrix
```

## Select elements in a data matrix
The same methods of selecting elements in a vector apply to data matrices. We use `[]`. The following selects the value in the first column, second row.
```{r}
final_harry_potter_matrix[1,2]
```

Adding a colon `:` selects the specified elements in a column.  
```{r}
final_harry_potter_matrix[1:4]
```

We can also select values in an entire row or column. This can be useful for calculations. Here we calculate the mean of the entire second column. 
```{r}
non_us_earnings <- all_harry_potter_matrix[ ,2]
mean(non_us_earnings)
```

## Practice
Below are data collected by three scientists (Jill, Steve, Susan in order) measuring temperatures of eight hot springs.
```{r}
spring_1 <- c(36.25, 35.40, 35.30)
spring_2 <- c(35.15, 35.35, 33.35)
spring_3 <- c(30.70, 29.65, 29.20)
spring_4 <- c(39.70, 40.05, 38.65)
spring_5 <- c(31.85, 31.40, 29.30)
spring_6 <- c(30.20, 30.65, 29.75)
spring_7 <- c(32.90, 32.50, 32.80)
spring_8 <- c(36.80, 36.45, 33.15)
```

1. Build a data matrix that has the springs as rows and the columns as scientists.  
```{r}
springs <- c(spring_1, spring_2, spring_3, spring_4, spring_5, spring_6, spring_7, spring_8)
springs_matrix <- matrix(springs, nrow=8, byrow = T)
springs_matrix
```

Let's name the rows and columns. Start by making two new vectors with the names.
```{r}
scientists <- c("Jill", "Steve", "Susan")
springs <- c("Bluebell Spring", "Opal Spring", "Riverside Spring", "Too Hot Spring", "Mystery Spring", "Emerald Spring", "Black Spring", "Pearl Spring")
```

2. Use `colnames()` and `rownames()` to name the columns and rows.  
```{r}
colnames(springs_matrix) <- scientists
rownames(springs_matrix) <- springs
springs_matrix
```

Calculate the mean temperature of all three springs. Notice the command `rowMeans`.
```{r}
mean_vector <- rowMeans(springs_matrix)
mean_vector
```

3. Add `mean_vector` as a new column.  
```{r}
springs_matrix2 <- cbind(springs_matrix, mean_vector)
springs_matrix2
```

## That's it, let's take a break!  
-->[Home](https://jmledford3115.github.io/datascibiol/)