---
title: "Lab 2.2"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Define data structure.    
2. Build a new vector and call elements within it.  
3. Combine a series of vectors into a data frame.  
4. Name columns and rows in a data frame.  
5. Select columns and rows and use summary functions.   
6. Write your data frame to a csv file!  

## Load the tidyverse
A library is a collection of R functions and data sets. The tidyverse is a collection of R packages designed for data science. For this course, we will be using many of the packages in the tidyverse. We load the tidyverse with the command below.
```{r}
library("tidyverse")
```

## Data Structures
In addition to classes of data, R also organizes data in different ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. Here, we will introduce vectors and data frames.  

## Vectors
Vectors are a common way of organizing data in R.  We create vectors using the `c` command. The `c` stands for concatenate. We used this in the first part of today's lab.  

A numeric vector.
```{r}
my_vector <- c(10, 20, 30)
```

A character vector.
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thrusday", "Friday", "Saturday", "Sunday")
```

A convenient trick for creating a vector to play around with is to generate a sequence of numbers.
```{r}

```

## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.
```{r}

```

```{r}

```

## Practice
1. Use `[]` to determine which element in `my_vector_sequence` has a value of 15.
```{r}

```

2. We can use operators such as <, >, ==, <==, etc. Show all values in `my_vector_sequence` that are less than or equal to 10. 
```{r}

```

3. If you use `[]` then you only get the values, not the logical evaluation of the entire vector. Experiment with this by adjusting the chunk below.    
```{r}

```

## Data Frames
The data frame is the most common way to organize data within R. A data frame can store data of many different classes. We usually don't build data frames in RStudio, but this example will show you how they are structured.  

Let's build separate vectors that include length (in), weight (oz), and sex of three [ruby-throated hummingbirds](https://www.audubon.org/field-guide/bird/ruby-throated-hummingbird).  
```{r}
Sex <- c("male", "female", "male")
Length <- c(3.2, 3.7, 3.4)
Weight <- c(2.9, 4.0, 3.1)
```

Since we work in the tidyverse, we use the `tibble()` function to create a data frame. A tibble is a modern take on data frames. Tibbles are data frames, but they tweak some older behaviors to make life a little easier.
```{r}

```

Notice that not only are the data neat and clean looking, there is also information provided about the class of data. `dbl` means that the value is a type of numeric [double precision floating point](http://uc-r.github.io/integer_double/). 
```{r}

```

What are the column names of our data frame? Notice that R defaulted to using the names of our vectors, but we could name them something else when creating the data frame, or rename them later.
```{r}

```

What are the dimensions of the `hbirds` data frame? The `dim()` and `str()` commands provide this information.   
```{r}

```

```{r}

```

Let's use lowercase names when we create the data frame. We just changed to lowercase here, but we could use any names we wish.  
```{r}

```

## Accessing Data Frame Columns and Rows 
The same methods of selecting elements in vectors and data matrices apply to data frames. We use `[]`. We have two positions where the first applies to the rows, and the second to the columns.  

The first row.  
```{r}

```

The third column.  
```{r}

```

## Calculations
We can use the `$` to access a column (variable) in a data frame. Here we calculate the mean length of the hummingbirds.  
```{r}

```

## Writing Data to File
We should save our hbirds data frame so we can use it again later! There are many ways to save data in R, here we write our data frame to a csv file. We use `row.names = FALSE` to avoid row numbers from printing out. 
```{r}

```

## Wrap-up
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  

-->[Home](https://jmledford3115.github.io/datascibiol/)