---
title: "Data Matrices"
author: "Joel Ledford"
date: "Winter 2020"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Combine a series of vectors into a data matrix.  
2. Name columns and rows in a data matrix.  
3. Select values and use summary functions in a data matrix.  
4. Explain the difference between a data matrix and a data frame.  

## Setup
At the beginning of each class, please make sure that you complete the following:  
1. Login to the lab computer.  
2. Open Github Desktop (it should be pre-installed) and login.  
3. Clone your repository to the desktop or somewhere convenient to access.  
4. Navigate to github.com and login.  
5. Go to the class repository `BIS15-W20-DataScienceBiologists`, click the `Clone or download` button then `Download zip` to the desktop.  
6. Copy the files for today's lab from the class repository and paste them into **your** repository.  
7. Open today's lab in RStudio.  

## Class Discussion Page  
+ [BIS15 Piazza](https://piazza.com/uc_davis/winter2020/bis015l/home)  

## Homework 1 Review  
Let's take a few minutes to compare solutions with your lab  partner. You should be especially mindful of alternate solutions; generally, the code that is the shortest in terms of lines is best. Make sure that your code is knitted and in your Github repository, please.  

## Data Matrix
Last time, you learned how to work with vectors. Today we will organize the vectors into a new type of data structure called a data matrix. Like vectors, data matrices are restricted to *data of the same type*. In the short example below, we will build a new data matrix using the matrix command.  

Box office earnings for Harry Potter movies (in millions!). Notice that these are separate vectors.
```{r}
Philosophers_Stone <- c(317.5, 657.1)
Chamber_of_Secrets <- c(261.9, 616.9)
Prisoner_of_Azkaban <- c(249.5, 547.1)
Goblet_of_Fire <- c(290.0, 606.8)
Order_of_the_Phoenix <- c(292.0, 647.8)
Half_Blood_Prince <- c(301.9, 632.4)
Deathly_Hallows_1 <- c(295.9, 664.3)
Deathly_Hallows_2 <- c(381.0, 960.5)
```

Create a new object called `box_office`. Here we are using the `c` command to combine the vectors into one.
```{r}
box_office <- c(Philosophers_Stone, Chamber_of_Secrets, Prisoner_of_Azkaban, Goblet_of_Fire, Order_of_the_Phoenix, Half_Blood_Prince, Deathly_Hallows_1, Deathly_Hallows_2)
box_office
```

Create `harry_potter_matrix` using the `matrix()` command. We need to tell R how to organize the `box_office` vector using the `nrow` and `byrow` commands.
```{r}
harry_potter_matrix <- matrix(box_office, nrow = 8, byrow = T)
harry_potter_matrix
```

## Name the rows and columns
Vectors `region` and `titles`, used for naming.
```{r}
region <- c("US", "non-US")
titles <- c("Philosophers_Stone", "Chamber_of_Secrets", "Prisoner_of_Azkaban", "Goblet_of_Fire", "Order_of_the_Phoenix", "Half_Blood_Prince", "Deathly_Hallows_1", "Deathly_Hallows_2")
```

Name the columns using `colnames()` with the vector region.
```{r}
colnames(harry_potter_matrix) <- region
```

Name the rows using `rownames()` with the vector titles.
```{r}
rownames(harry_potter_matrix) <- titles
```

Print `harry_potter_matrix`.
```{r}
harry_potter_matrix
```

## Using a data matrix
Once you have a data matrix, you can perform lots of different analyses. For example, you can calculate the total earnings of each movie.
```{r}
global <- rowSums(harry_potter_matrix)
global
```

And even add a new column to reflect this calculation. `cbind()` adds columns.
```{r}
all_harry_potter_matrix <- cbind(harry_potter_matrix, global)
all_harry_potter_matrix
```

## Practice
1. What are the total earnings for each of the regions?  

2. Add this information to the data matrix. Hint: you are adding a row, not a column.  

## Select elements in a data matrix
The same methods of selecting elements in a vector apply to data matrices. We use `[]`. The following selects the value in the first column, second row.
```{r}
all_harry_potter_matrix[1,2]
```

Adding a colon `:` selects the specified elements in a column.  
```{r}
all_harry_potter_matrix[1:4]
```

We can also select values in an entire row or column. This can be useful for calculations. Here we calculate the mean of the entire second column. 
```{r}
non_us_earnings <- all_harry_potter_matrix[ ,2]
mean(non_us_earnings)
```

## Practice
Below are data collected by three scientists (Jill, Steve, Susan in order) measuring temperatures of eight hot springs.
```{r}
spring_1 <- c(36.25, 35.40, 35.30)
spring_2 <- c(35.15, 35.35, 33.35)
spring_3 <- c(30.70, 29.65, 29.20)
spring_4 <- c(39.70, 40.05, 38.65)
spring_5 <- c(31.85, 31.40, 29.30)
spring_6 <- c(30.20, 30.65, 29.75)
spring_7 <- c(32.90, 32.50, 32.80)
spring_8 <- c(36.80, 36.45, 33.15)
```

1. Build a data matrix that has the springs as rows and the columns as scientists.  
```{r}

```


Let's name the rows and columns. Start by making two new vectors with the names.
```{r}
scientists <- c("Jill", "Steve", "Susan")
springs <- c("Bluebell Spring", "Opal Spring", "Riverside Spring", "Too Hot Spring", "Mystery Spring", "Emerald Spring", "Black Spring", "Pearl Spring")
```

2. Use `colnames()` and `rownames()` to name the columns and rows.  
```{r}

```

Calculate the mean temperature of all three springs. Notice the command `rowMeans`.
```{r}
#mean_vector <- rowMeans(springs_matrix)
#mean_vector
```

3. Add `mean_vector` as a new column.  
```{r}

```


## That's it, let's take a break!   

--> On to [part 2](https://jmledford3115.github.io/datascibiol/lab2_2.html)  

[gh]: www.github.com  
[gd]: https://desktop.github.com/
[BIS15]: https://github.com/jmledford3115/BIS15-W20-DataScienceBiologists