---
title: "Data Structures: Vectors and Data Frames"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Breakout Rooms  
Please take 5-8 minutes to check over your answers to HW 1 in your group. If you are stuck, please remember that you can check the key in [Joel's repository](https://github.com/jmledford3115/BIS15LW2021_jledford).  

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Combine a series of vectors into a data frame.  
2. Name columns and rows in a data frame.  
3. Select columns and rows and use summary functions.   
4. Change data classes of columns in the data frame.  
5. Write your data frame to a csv file!  

## Data Structures
In addition to classes of data, R also organizes data in different ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. For now, we will focus on `vectors`.  

## Vectors
Vectors are a common way of organizing data in R.  We create vectors using the `c` command. The `c` stands for concatenate.  

A numeric vector.
```{r}
my_vector <- c(10, 20, 30)
```

A character vector. Characters always have quotes and may be referred to as "strings".
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thrusday", "Friday", "Saturday", "Sunday")
```

What if our numeric vector above had a missing value?
```{r}
my_vector <- c(10, NA, 30)
anyNA(my_vector)
```

Certain functions require us to specify how we want to deal with NAs. Often if NA is included in an operation, the result is NA.
```{r}
mean(my_vector)
```

Let's remove the NA value.
```{r}
mean(my_vector, na.rm = TRUE)
```

A convenient trick for creating a vector is to generate a sequence of numbers.
```{r}
my_vector_sequence <- c(1:10)
my_vector_sequence
```

`seq` allows more control.
```{r}
my_vector_sequence2 <- seq(1, 4, by=0.25)
my_vector_sequence2
```

## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.
```{r}
days_of_the_week[3]
```

## Practice
Work through these examples and make adjustments to the values to experiment.

1. Create a new numeric vector of values 1 through 100
```{r}
new_vector <- c(1:100)
```

2. We can evaluate all numerics in our vector to determine which has a value of 15.
```{r}
new_vector == 15
```

3. We can also use operators such as <, >, ==, <==, etc.
```{r}
new_vector < 10
```

4. If you use `[]` then you only get the values, not the logical evaluation of the entire vector.
```{r}
new_vector[new_vector <= 10]
```





## Load tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Data Frame
During the first part of lab, you learned how to work with vectors. Today we will organize the vectors into a new type of data structure called a **data frame**. The data frame is the most common way to organize data within R. You can think of a data frame as similar to an Excel spreadsheet. A data frame can store data of different types. 

Let's build three separate vectors of length (in), weight (oz), and sex of three ruby throated hummingbirds.
```{r}
Sex <- c("Male", "Female", "Male")
Length <- c(3.2, 3.7, 3.4)
Weight <- c(2.9, 4.0, 3.1)
```

Here we combine our three vectors to create a data frame with the function `data.frame()`.
```{r}
hbirds <- data.frame(Sex, Length, Weight)
hbirds
```

Notice that not only are the data neat and clean looking, there is also information provided about the type of data in the frame. `dbl` means that the value is a type of numeric [double precision floating point](http://uc-r.github.io/integer_double/). 

What are the column names of our data frame? Notice that R defaulted to using the names of our vectors, but we could name them something else when creating the data frame, or rename them later.
```{r}
names(hbirds)
```

Let's use lowercase names when we create the data frame. We just changed to lowercase here, but we could use any names we wish.
```{r}
hbirds <- data.frame(sex = Sex,length_in = Length, weight_oz = Weight) #renaming will become more helpful in later labs
names(hbirds)
```

```{r}
hbirds
```

## Accessing Data Frame Columns and Rows 
The same methods of selecting elements in a vector apply to data frames. We use `[]`. But now we have two positions where the first applies to the rows, and the second to the columns.  

The first row.  
```{r}
hbirds[1,]
```

The third column.  
```{r}
hbirds[,3]
```

We can also select values in an entire column using the `$` sign. This can be useful for calculations. Here we calculate the mean of all of the birds weights. 
```{r}
w <- hbirds$weight_oz
mean(w)
```

## Summaries
We can also summarize our entire data frame with the`summary()` function.
```{r}
summary(hbirds)
```

## Glimpse
`glimpse()` is another useful summary function.
```{r}
glimpse(hbirds)
```

## Adding Rows
We can add a new bird's data with the `rbind()` function, known as row bind. It will bind our new vector to our data frame row-wise.
```{r}
new_bird <- c("Female", 3.6, 3.9)
new_bird
```

```{r}
hbirds<- rbind(hbirds, new_bird)
hbirds
```

## Adding Columns
What if we come across more data and we want to add it to our data frame? We can do that with the `$` operator.
```{r}
hbirds$neighborhood <- c("Lakewood", "Brentwood", "Lakewood", "Scenic Heights")
hbirds
```

## Changing Column Classes
Great, we added our new column, but it's a character and not a factor. A factor is a special type of character vector with predefined levels. 
```{r}
class(hbirds$neighborhood)
hbirds$neighborhood <- as.factor(hbirds$neighborhood)
```

```{r}
class(hbirds$neighborhood)
```

`levels()` will give you a list of the different levels within a factor.  
```{r}
levels(hbirds$neighborhood)
```

## Writing Data to File
We should save our hbirds data frame so we can use it again later! There are many ways to save data in R, here we write our data frame to a csv file. We use  `row.names = FALSE` to avoid row numbers from printing out. 
```{r}
write.csv(hbirds, "hbirds_data.csv", row.names = FALSE)
```

## Practice
Below are data collected by three scientists (Jill, Steve, Susan in order) measuring temperatures of three hot springs near Mono Lake.
```{r}
temp <- c(36.25, 35.40, 35.30, 35.15, 35.35, 33.35, 30.70, 29.65, 29.20)
name <- c("Jill","Susan", "Steve", "Jill","Susan", "Steve", "Jill","Susan", "Steve")
spring <- c("Buckeye", "Buckeye", "Buckeye", "Benton", "Benton", "Benton", "Travertine", "Travertine", "Travertine")
```

1. Build a data frame called `hsprings` with the above data. Name the temperature column `temp_C`. Print out the data frame.  
```{r}
hsprings <- data.frame(temp_C = temp, name, spring)
hsprings
```

2. Change the column titled `name` to `scientist`, leave the other column names the same. Print out the data frame names.  
```{r}
names(hsprings)[2] <- c("scientist")
hsprings
```

3. Our scientists forgot to record the depth data for each spring. Here it is as a vector, add it as a new column called depth_ft: `c(4.15, 4.13, 4.12, 3.21, 3.23, 3.20, 5.67, 5.65, 5.66)`. Print out the data frame.
```{r}
hsprings$depth_ft <- c(4.15, 4.13, 4.12, 3.21, 3.23, 3.20, 5.67, 5.65, 5.66)
hsprings
```

4. Calculate the mean temperature of all of the temperature measurements.
```{r}
mean(hsprings$temp_C)
```

5. What is the class of the scientist column? Use an R command to see all the unique levels.
```{r}
class(hsprings$scientist)
levels(hsprings$scientist)
```

6. Save your hot springs data as a csv file! Do not allow row names. 
```{r}
write.csv(hsprings, "hsprings_data.csv", row.names = FALSE)
```

## Wrap-up
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)