---
title: "Lab 2 Homework"
author: "Please Add Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Practice
Below are data collected by three scientists (Jill, Steve, Susan in order) measuring temperatures of eight hot springs.
```{r}
spring_1 <- c(36.25, 35.40, 35.30)
spring_2 <- c(35.15, 35.35, 33.35)
spring_3 <- c(30.70, 29.65, 29.20)
spring_4 <- c(39.70, 40.05, 38.65)
spring_5 <- c(31.85, 31.40, 29.30)
spring_6 <- c(30.20, 30.65, 29.75)
spring_7 <- c(32.90, 32.50, 32.80)
spring_8 <- c(36.80, 36.45, 33.15)
```

1. Build a data matrix that has the springs as rows and the columns as scientists.  
```{r}
springs <- c(spring_1, spring_2, spring_3, spring_4, spring_5, spring_6, spring_7, spring_8)
springs_matrix <- matrix(springs, nrow=8, byrow = T)
springs_matrix
```

Let's name the rows and columns. Start by making two new vectors with the names.
```{r}
scientists <- c("Jill", "Steve", "Susan")
springs <- c("Bluebell Spring", "Opal Spring", "Riverside Spring", "Too Hot Spring", "Mystery Spring", "Emerald Spring", "Black Spring", "Pearl Spring")
```

2. Use `colnames()` and `rownames()` to name the columns and rows.  
```{r}
colnames(springs_matrix) <- scientists
rownames(springs_matrix) <- springs
springs_matrix
```

Calculate the mean temperature of all three springs. Notice the command `rowMeans`.
```{r}
mean_vector <- rowMeans(springs_matrix)
mean_vector
```

3. Add `mean_vector` as a new column.  
```{r}
springs_matrix2 <- cbind(springs_matrix, mean_vector)
springs_matrix2
```