---
title: "App for Animal Crrossing Presentation"
author: "DQ"
date: "2023-03-01"
output: html_document
---

## Loading libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
library(shiny)
library(shinydashboard)
library(lubridate)
```

## Loading the data
```{r message=FALSE, warning=FALSE}
bugs <- clean_names(read_csv(here("data","insects.csv")))
fish <- clean_names(read_csv(here("data","fish.csv")))
```

## Fixing error in data
```{r}
bugs[65,34] <- "Ins45"
```

## Cleaning up the "bugs" data frame
```{r}
bugs <- bugs %>% 
  select(-"critterpedia_filename", -"furniture_filename", -contains("id"), -contains("sh"), -contains("color"), -"number") %>% 
  mutate(category = "bugs")
```

## Cleaning up the "fish" data frame
```{r}
fish <- fish %>% 
  select(-"critterpedia_filename", -"furniture_filename", -contains("id"), -contains("sh"), -contains("color"), -"number", -"lighting_type", -"rain_snow_catch_up", -"size") %>% 
  mutate(category = "fish")
```

## Combining "fish" and "bugs" into the "museum" data frame
```{r message=FALSE, warning=FALSE}
museum <- full_join(fish, bugs)
```

## Pivoting the months & times into 2 columns
```{r}
 museum <- museum %>%
  pivot_longer(nh_jan:nh_dec,
               names_to = "month",
               names_prefix = "nh_",
               values_to = "times")
```

## Separating the "times" column into "start times" and "end times"
```{r message=FALSE, warning=FALSE}
museum <- museum %>% 
  filter(times != "NA") %>% 
  mutate_all(~ gsub("\\ ", " ", .)) %>% 
  replace(. == "All day", "0 AM – 24 AM") %>% 
  separate(times, into = c("start_times","end_times"), sep = " – ") %>% 
  separate(start_times, into = c("start_times","start_td"), sep = " ") %>% 
  separate(end_times, into = c("end_times","end_td"), sep = " ")
```

## Converting "start times" and "end times" to numerics
```{r}
museum$start_times <- as.numeric(museum$start_times)
museum$end_times <- as.numeric(museum$end_times)
```

## Converting all times to 24 hours
```{r}
museum <- museum %>% 
  mutate(start_times = ifelse(start_times == 12, 0, start_times)) %>% 
  mutate(start_times = ifelse(start_td == "PM",start_times+12,start_times)) %>%
  mutate(end_times = ifelse(end_times == 12, 0, end_times)) %>% 
  mutate(end_times = ifelse(end_td == "PM",end_times+12,end_times)) %>% 
  select(-"start_td",-"end_td")
```

## Converting "months" to factors based on the built in "month.abb" levels
```{r}
month.abb.l <- tolower(month.abb)
museum$month <- factor(museum$month, levels = month.abb.l)
```

## Creating a color palatte corresponding to each month
```{r}
myColors <- c("#7BAFF7", "#A4D2F1", "#27A34D", "#A9F364", "#FCBBD3", "#F72D2D", "#FF800F", "#FFBE13", "#815800", "#B9923E", "#D6D1B2", "#6B87AE")
```

## Adding ".png" to the every "icon_filename" value
```{r}
museum <- museum %>% 
  mutate(icon_filename = paste(museum$icon_filename, ".png", sep = ""))
```

## The finished app
```{r}
ui <- dashboardPage(skin = "purple",
  dashboardHeader(title = "Animal Crossing Museum Guide", titleWidth = 330),
  dashboardSidebar(disable = T),
  dashboardBody(
    fluidRow(
      box(title = "Hemisphere", width = 2, height = "150px",
        radioButtons("hemi", "Select Hemisphere", 
                    choices = c("Northern","Southern"))
      ),
      box(title = "Category", width = 2, height = "150px",
        radioButtons("cat", "Select Category", 
                    choices = c("Fish","Bugs")),
      ),
      box(title = "Species", width = 3, height = "150px",
        selectInput("spec", "Select Species", 
                    choices = NULL)
      ),
      box(title = NULL, width = 2, height = "150px", align = "center",
        imageOutput("icon")
      ),
      box(title = "Catch Info", width = 3, height = "150px",
          textOutput("catch_info"),
          textOutput("weather_info"),
          textOutput("sell_info")
      )    
    ),
    fluidRow(
      box(title = "Catching Times by Month", width = 8, align = "center",
          plotOutput("plot", 
             width = "600px", 
             height = "400px")
      ),
      box(style='height:420px;overflow-y: scroll;', title = "Currently Available to Catch", width = 4,
          tableOutput("available")
      )
    )
  )
)

server <- function(input, output, session) { 
  observe({
    new_choices <- unique(filter(museum, category == tolower(input$cat))$name)
    updateSelectInput(session, "spec", 
                      choices = new_choices)
  })
  output$plot <- renderPlot({
    if (input$hemi == "Southern"){
      mod_museum <- mutate(museum, month = month.abb.l[(as.integer(month) + 6 - 1) %% 12 + 1])
    } else {
      mod_museum <- museum
    }
    ggplot(filter(mod_museum, mod_museum$name == input$spec), 
           aes_string(x = "month", 
                      ymin = "start_times", 
                      ymax = "end_times", 
                      color = "month"),
           xlim = c("jan","dec")) +
      geom_linerange(size = 14) +
      theme_classic(base_size = 18) +
      labs(x = NULL,
           y = "Times (24hrs)") +
      scale_color_manual(values = myColors, drop = T, limits = levels(museum$month)) +
      theme(axis.title.y = element_text(size = 24),
            panel.grid.major.y = element_line(),
            axis.line = element_blank()) +
      scale_y_continuous(limits = c(0,24), breaks = seq(0, 24, by = 1)) +
      xlim("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec")
  })
  output$catch_info <- renderText({
    paste("Where/How: ", filter(museum, name == input$spec)$where_how[1])
  })
  output$weather_info <- renderText({
    weather <- filter(museum, name == input$spec)$weather[1]
    if (is.na(weather)) {
        NULL
    } else {
        paste("Weather: ", weather)
    }
  })
  output$sell_info <- renderText({
    paste("Sell price: ", filter(museum, name == input$spec)$sell[1])
  })
  output$available <- renderTable({
    if (input$hemi == "Southern"){
      mod_museum <- mutate(museum, month = month.abb.l[(as.integer(month) + 6 - 1) %% 12 + 1])
    } else {
      mod_museum <- museum
    }
    select(filter(mod_museum, hour(now())<end_times & hour(now())>start_times & tolower(month(now(),label=T)) == month), "name","where_how") %>% 
      arrange(name) %>%
      rename("Where/How"=where_how,"Name"=name)
  })
  output$icon <- renderImage({
    filename <- here("images", filter(museum, name == input$spec)$icon_filename[1])
    list(src = filename, width = "118px", height = "118px")
  },
  deleteFile = FALSE
  )

  session$onSessionEnded(stopApp)
}

shinyApp(ui, server)
```


-->[Home](https://jmledford3115.github.io/datascibiol/)
