---
title: "Homework 3"
author: "Type Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run.  

## Load the tidyverse
```{r}
library("tidyverse")
options(scipen=999) # turn off scientific notation
```

## Data
For the homework, we will use data about vertebrate home range sizes. The data are in the class folder and the reference is below.  

**Database of vertebrate home range sizes.**. 
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

**1. Load the data into a new object called `homerange`.**
```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```

**2. What are the dimensions of the dataframe?**  
```{r}
dim(homerange)
```

**3. Are there any NA's in the dataframe? Try using summary to determine which variables have more or less NA's.**
```{r}
summary(homerange)
```

```{r}
glimpse(homerange)
```

**4. What are the names of the columns in the dataframe?**
```{r}
names(homerange)
```

**5. Based on the summary output, do you see anything in the data that looks strange? Think like a biologist and consider the variables.**  

The mean mass is zero for some animal? Not really, the minimum is 0.22. There are also some negative numbers for log10.mass and log10.hra.

```{r}
filter(homerange, mean.mass.g == 0)
```

**6. The `min` and `max` functions can be used to find the minimum and maximum values in a vector. Use these functions to find the minimum and maximum values for the variable `mean.mass.g`.**  
```{r}
min(homerange$mean.mass.g)
```

```{r}
max(homerange$mean.mass.g)
```

**7. Change the class of the variables `taxon` and `order` to factors and display their levels.**  
```{r}
homerange$taxon <- as.factor(homerange$taxon)
homerange$order <- as.factor(homerange$order)
```

```{r}
levels(homerange$taxon)
```

```{r}
levels(homerange$order)
```

**8. Use `select` to pull out the variables taxon and common.name.**  
```{r}
select(homerange, taxon, common.name)
```

**9. Use `filter` to pull out all mammals from the data.**
```{r}
mammals <- filter(homerange, taxon == "mammals")
```

**10. What is the largest `mean.hra.m2` for mammals? This is a very large number! Which animal has this homerange? Look them up online and see if you can figure out why this number is so large.**
```{r}
max(mammals$mean.hra.m2)
```

```{r}
filter(mammals, mean.hra.m2==3550830977)
```

## Knit and Upload
Please knit your work as a .pdf or .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  