---
title: "Data Frames"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Combine a series of vectors into a data frame.  
2. Name columns and rows in a data frame.  
3. Select columns and rows and use summary functions.   
4. Change data classes of columns in the data frame.  
5. Write your data frame to a csv file!  

## Load tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Data Frames
During lab 2, you learned how to work with vectors and data matrices. Today we will organize the vectors into a new type of data structure called a **data frame**. The data frame is the most common way to organize data within R. You can think of a data frame as similar to a spreadsheet. A data frame can store data of many different classes.  

Let's build separate vectors that include length (in), weight (oz), and sex of three [ruby-throated hummingbirds](https://www.audubon.org/field-guide/bird/ruby-throated-hummingbird).  
```{r}
Sex <- c("male", "female", "male")
Length <- c(3.2, 3.7, 3.4)
Weight <- c(2.9, 4.0, 3.1)
```

Here we combine our three vectors to create a data frame with the function `data.frame()`.
```{r}
hbirds <- data.frame(Sex, Length, Weight)
hbirds
```

OK, stop and think- how does this differ from making a data matrix? Are the data treated differently?  
```{r}
hummingbirds <- c(Sex, Length, Weight)
hummingbirds_matrix <- matrix(hummingbirds, ncol=3, byrow = F)
hummingbirds_matrix
```

Notice that not only are the data neat and clean looking, there is also information provided about the class of data. `dbl` means that the value is a type of numeric [double precision floating point](http://uc-r.github.io/integer_double/). 

What are the column names of our data frame? Notice that R defaulted to using the names of our vectors, but we could name them something else when creating the data frame, or rename them later.
```{r}

```

What are the dimensions of the `hbirds` data frame? The `dim()` and `str()` commands provide this information.   
```{r}

```

```{r}

```

Let's use lowercase names when we create the data frame. We just changed to lowercase here, but we could use any names we wish.  
```{r}

```

```{r}

```

## Accessing Data Frame Columns and Rows 
The same methods of selecting elements in vectors and data matrices apply to data frames. We use `[]`. We have two positions where the first applies to the rows, and the second to the columns.  

The first row.  
```{r}
hbirds[1,]
```

The third column.  
```{r}
hbirds[ ,3]
```

We can also select values in an entire column using the `$` sign. This can be useful for calculations. Here we calculate the mean of all of the birds weights. 
```{r}

```

## Adding Rows
We can add a new bird's data with the `rbind()` function, known as row bind. It will bind our new vector to our data frame row-wise.
```{r}
new_bird <- c("female", 3.6, 3.9)
new_bird
```

```{r}
hbirds<- rbind(hbirds, new_bird)
hbirds
```

## Adding Columns
What if we come across more data and we want to add it to our data frame? We can also do that with the `$` operator.
```{r}
hbirds$neighborhood <- c("lakewood", "brentwood", "lakewood", "scenic Heights")
hbirds
```

## Writing Data to File
We should save our hbirds data frame so we can use it again later! There are many ways to save data in R, here we write our data frame to a csv file. We use `row.names = FALSE` to avoid row numbers from printing out. 
```{r}
write.csv(hbirds, "hbirds_data.csv", row.names = FALSE)
```

## Practice
Below are data collected by three scientists (Jill, Steve, Susan in order) measuring temperatures of three hot springs near Mono Lake.
```{r}
temp <- c(36.25, 35.40, 35.30, 35.15, 35.35, 33.35, 30.70, 29.65, 29.20)
name <- c("Jill", "Susan", "Steve", "Jill", "Susan", "Steve", "Jill", "Susan", "Steve")
spring <- c("Buckeye", "Buckeye", "Buckeye", "Benton", "Benton", "Benton", "Travertine", "Travertine", "Travertine")
```

1. Build a data frame called `hsprings` with the above data. Name the temperature column `temp_C`. Print out the data frame.  
```{r}

```

2. Change the column titled `name` to `scientist`, leave the other column names the same. Print out the data frame names.  
```{r}

```

3. Our scientists forgot to record the depth data for each spring. Here it is as a vector, add it as a new column called depth_ft: `c(4.15, 4.13, 4.12, 3.21, 3.23, 3.20, 5.67, 5.65, 5.66)`. Print out the data frame.
```{r}

```

4. Calculate the mean temperature of all of the temperature measurements.
```{r}

```

5. Save your hot springs data as a `.csv` file! Do not allow row names. 
```{r}

```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  