---
title: "Importing Data Frames"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Import .csv files as data frames using `read_csv()`.  
2. Use summary functions to explore the dimensions, structure, and contents of a data frame.  
3. Use the `select()` command of dplyr to sort data frames.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Practice
In your lab 3 folder there is another folder titled `data`. Inside the `data` folder there is a `.csv` titled `Gaeta_etal_CLC_data.csv`. Open this data and store them as an object called `fish`.  

The data are from: Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.  [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
```{r}
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")
```

Notice that when the data are imported, you are presented with a message that tells you how R interpreted the column classes. This is also where error messages will appear if there are problems.  

## Summary functions
Once data have been uploaded, you may want to get an idea of its structure, contents, and dimensions. I routinely run one or more of these commands when data are first imported.  

We can summarize our data frame with the`summary()` function.  
```{r}
#summary(fish)
```

`glimpse()` is another useful summary function.
```{r}
#glimpse(fish)
```

`nrow()` gives the numbers of rows.
```{r}
#nrow(fish)
```

`ncol` gives the number of columns.
```{r}
#ncol(fish)
```

`dim()` gives the dimensions.
```{r}
#dim(fish)
```

`names` gives the column names.
```{r}
#names(fish)
```

`head()` prints the first n rows of the data frame.
```{r}
#head(fish)
```

`tail()` prinst the last n rows of the data frame.
```{r}
#tail(fish)
```

`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable. We will come back to this later... 
```{r}
#table(fish$lakeid)
```

## Select
Select is a way of sorting data frames by pulling out variables (columns) of interest.
```{r}
#select(fish, lakeid, length)
```

## Filter
Filter is a way of pulling out observations that meet specific criteria in a variable. We will work a lot more with this in the next lab.  
```{r}
#filter(fish, length<=100)
```

## Practice
1. Load the data `mammal_lifehistories_v2.csv` and place it into a new object called `mammals`.

2. Provide the dimensions of the data frame.

3. Check the column names in the data frame. 

4. Use `str()` to show the structure of the data frame and its individual columns; compare this to `glimpse()`. 

5. Try the `table()` command to produce counts of mammal order, family, and genus.  

6. Use the `filter()` command to pull out mammals that have a gestation period greater than or equal to one year.  

## Wrap-up
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)