---
title: "Tidyverse: `select()`, pipes ` %>%`, and `mutate()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Use the functions of dplyr (filter, select, arrange) to organize and sort data frames.  
2. Use `mutate()` to calculate a new column from existing columns.  

## Review
In the last lab you were introduced to the tidyverse and the core function `filter()`. Filter allows us to separate variables with specific attributes of interest from large, complex data. `filter()` works by looking through rows, but what about columns? We need the new function `select()` to separate columns of interest. As you may expect, we will use `filter()` and `select()` together to focus on the variables and values of interest.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Load the fish scale data
```{r message=FALSE, warning=FALSE}
fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")
```

```{r}
names(fish)
```

## dplyr: select()
Select allows you to build a new data frame by selecting your variable(s) of interest. Our fish data has six variables, but this should give you some ideas especially when you have a large data frames with lots of variables.  

What if we are only interested in lakeid and scalelength.
```{r}
select(fish, lakeid, scalelength)
```

The - operator is useful in select. It allows us to select everything except the specified variables.
```{r}
select(fish, -fish_id, -annnumber, -length, -radii_length_mm)
```

For very large data frames with lots of variables, `select()` utilizes lots of different operators to make things easier. Let's say we are only interested in the variables that deal with length.
```{r}
select(fish, contains("length"))
```

Some additional options to select columns based on a specific criteria include:  
1. ends_with() = Select columns that end with a character string  
2. contains() = Select columns that contain a character string  
3. matches() = Select columns that match a regular expression  
4. one_of() = Select columns names that are from a group of names  

## Practice
1. What are the names of the columns in the `fish` data?

2. We are only interested in the variables `lakeid`, `length`, and `scalelength`. Use `select()` to build a new dataframe focused on these variables.

3. Repeat #2 above using the `-` operator.  

## dplyr: Can we combine filter() and select()?
Absolutely. This is one of the strengths of the tidyverse, it uses the same grammar to specify commands.
```{r}
fish2 <- select(fish, lakeid, scalelength)
filter(fish2, lakeid == "AL")
```

## dplyr: Pipes %>% 
The code above works fine but there is a more efficient way. We need to learn pipes `%>%`. Pipes allow you to feed the output from one function to the input of another function. We are going to use pipes from here on to keep things cleaner. (command+shift+m)
```{r}
fish %>%
  select(lakeid, scalelength) %>%
  filter(lakeid == "AL")
```

## Practice
1. Build a dataframe focused on the lakes `AL` and `AR` and looking at radii length between 2 and 4 only. Use pipes!  

## dplyr: arrange()
The `arrange()` command is a bit like a sort command in excel.
```{r}
fish %>% 
  arrange(scalelength)
```

To sort in decreasing order, wrap the variable name in `desc()`.
```{r}
fish %>% 
  arrange(desc(scalelength))
```

`arrange()` can be very helpful in combination with the other functions.
```{r}
fish %>% 
  select(lakeid, length, fish_id, scalelength) %>% 
  filter(lakeid=="AL") %>% 
  arrange(fish_id)
```

## dplyr: mutate()  
Mutate allows us to create a new variable from existing variables in our data frame. Let's convert the length variable from cm to inches and create a new variable called length_in.  
```{r}
fish %>% 
  mutate(length_in = length * 0.394) %>% 
  select(fish_id, length, length_in)
```

Note that if we had called our new variable length, R would overwrite the old variable. 

## Practice
1. Use `mutate()` to calculate the half length of each fish: length_half = length/2. Select only fish_id, length, and length_half, use pipes.

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)