---
title: "Transforming data 2: pipes, `filter()`, `mutate()`, and `summarize()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Use the functions of dplyr (filter, select, arrange) to organize and sort data frames.  
2. Use `mutate()` to calculate a new column from existing columns.  

## Review  
In the previous lab, we used `select()` to extract columns of interest from a data frame. This helps us focus our attention on the variables relevant to our question. However, it doesn't allow us to extract observations from within the data frame. The `filter()` function allows us to extract data that meet specific criterea. When combined with `select()`, we have the power to transform, shape, and explore data with the potential to make new discoveries.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Load the data
```{r}
fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")
mammals <- readr::read_csv("data/mammal_lifehistories_v2.csv")
```

## `filter()`
Unlike `select()`, `filter()` allows us to extract data that meet specific criterea within a data frame. Let's say that we are interested only in the fish that occur in lake "AL". We can use `filter()` to extract these observations.
```{r}
filter(fish, lakeid == "AL")
```

Similarly, if we are only interested in fish with a length greater than or equal to 350 we can use `filter()` to extract these observations.  
```{r}
filter(fish, length >= 350)
```

+ `filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).  

Using the `!` operator allows for the exclusion of specific obeservations.
```{r}
filter(fish, lakeid != "AL")
```

Filtering multiple values within the same variable requires the `%in%` operator.  
```{r}
filter(fish, length %in% c(167, 175))
```

Alternatively, you can use `between` if you are looking for a range of values.  
```{r}
filter(fish, between(length, 167, 175))
```

You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
filter(fish, near(radii_length_mm, 2, tol = 0.2))
```

## Practice
1. Filter the `fish` data to include the samples from lake "BO".
```{r}
filter(fish, lakeid == "BO")
```

2. Filter the data to include all lakes except "AR".
```{r}
filter(fish, lakeid != "AR")
```

3. Filter the fish data to include all fish with a scalelength within 0.25 of 8.
```{r}
filter(fish, near(scalelength, 8, tol = 0.5))
```

4. Filter the fish data to include fish with a scalelength between 2 and 4.
```{r}
filter(fish, between(scalelength, 2, 4))
```

## `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract only the fish that have lakeid "AL" and length >350.
```{r}
filter(fish, lakeid == "AL", length > 350)
```

Notice that the `|` operator generates a different result.
```{r}
filter(fish, lakeid == "AL" | length > 350)
```

Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met.  
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 and/or condition 2 is met.  
+ `filter(xor(condition1, condition2)` will return all rows where only one of the conditions is met, and not when both conditions are met.  

In this case, we filter out the fish with a length over 400 and a scale length over 11 or a radii length over 8.
```{r}
filter(fish, length > 400, (scalelength > 11 | radii_length_mm > 8))
```

## Practice
1.  From the `mammals` data, filter all members of the family Bovidae with a mass greater than 450000.
```{r}
filter(mammals, family=="Bovidae", mass>=450000)
```

2. From the `mammals` data, filter only Carnivora and Primates.
```{r}
filter(mammals, order %in% c("Carnivora", "Primates"))
```

## Pipes `%>%` 
In order to start combining `select()` and `filter()` efficiently, we need to learn pipes. Pipes feed the output from one function to the input of another function. This helps us keep our code sequential and clean.  

Here we call up the fish data, select only lakeid and scalelength then filter that output only for lakes "AL". Notice that we only need to call the data one time.
```{r}
fish %>%
  select(lakeid, scalelength) %>%
  filter(lakeid == "AL")
```

## Practice
1. Build a dataframe focused on lakeid and radii_length but restricted to the fish from the lakes "AL" and "AR" with a radii length between 2 and 4. Use pipes!  
```{r}
fish %>% 
  select(lakeid, radii_length_mm) %>% 
  filter(lakeid=="AL" | lakeid=="AR", (between(radii_length_mm, 2, 4)))
```

## `arrange()`
The `arrange()` command is a bit like a sort command in excel.
```{r}
fish %>% 
  arrange(scalelength)
```

To sort in decreasing order, wrap the variable name in `desc()`.
```{r}
fish %>% 
  arrange(desc(scalelength))
```

`arrange()` can be very helpful in combination with the other functions.
```{r}
fish %>% 
  select(lakeid, length, fish_id, scalelength) %>% 
  filter(lakeid=="AL") %>% 
  arrange(fish_id)
```

## `mutate()`  
Mutate allows us to create a new column from existing columns in a data frame. Let's convert the length variable from cm to millimeters and create a new variable called length_mm.  
```{r}
fish %>% 
  mutate(length_mm = length*100) %>% 
  select(fish_id, length, length_mm)
```

## Practice
1. Use `mutate()` to calculate the half length of each fish: length_half = length/2. Select only fish_id, length, and length_half, use pipes.
```{r}
fish %>% 
  mutate(length_half=length/2) %>% 
  select(fish_id, length, length_half)
```

## `if_else()`
We will briefly introduce `if_else()` here because it allows us to use `mutate()` but not the entire column affected in the same way. In a sense, this can function like find and replace in a spreadsheet program. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  

Have a look at the data from mammals below. Notice that the values for newborn include `-999.00`. This is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.
```{r}
mammals %>% 
  select(Genus, species, newborn) %>% 
  arrange(newborn)
```

```{r}
mammals %>% 
  select(Genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))
```

## `mutate_all()`
This last function is super helpful when cleaning data. With "wild" data, there are often mixed entries (upper and lowercase), blank spaces, odd characters, etc. These all need to be dealt with before analysis.  

Here is an example that changes all entries to lowercase (if present).  
```{r}
mammals %>%
  mutate_all(tolower)
```

Here I take the mammals data and add some random characters as an example.
```{r}
mammals_ohno <- 
  mammals %>%
  mutate_all(~paste(., "  /n  "))

mammals_ohno[,1:4]
```

```{r}
mammals_corr <- 
  mammals_ohno %>%
  mutate_all(~str_replace_all(., "/n", "")) %>%
  mutate_all(str_trim)

mammals_corr[,1:4]
```

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)