---
title: "Lab 3.2"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Import .csv files as data frames using `read_csv()`.  
2. Understand the importance of paths and working directories to import data.  
2. Use summary functions to explore the dimensions, structure, and contents of a data frame.  

## Load tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Data Frames
In part 1, you learned how to work with vectors and data frames. For the remainder of the course, we will work exclusively with data frames. Recall that data frames store multiple classes of data. Last time, you were shown how to build data frames by combining multiple vectors using  `tibble()`.  

## Importing Data
Scientists often make their data available as supplementary material associated with a publication. This is excellent scientific practice as it insures repeatability by showing exactly how analyses were performed. As data scientists, we capitalize on this by importing data directly into R.  

R allows us to import a wide variety of data types. The most common type of file is a .csv file which stands for comma separated values. Spreadsheets are often developed in Excel then saved as .csv files for use in R. There are packages that allow you to open excel files and many other formats but .csv is the most common.  

To import a file, first make sure that you are in the correct working directory. If you are not in the correct directory, R will not "see" the file.
```{r}
getwd()
```

## Load the data
Here we import a .csv file. Since we are using the tidyverse, we import the file using `read_csv()`. `readr` is included in the tidyverse set of packages.  
```{r}
hot_springs <- read_csv("hsprings.csv")
```

Notice that when the data are imported, you are presented with a message that tells you how R interpreted the column classes. This is also where error messages will appear if there are problems.  

Use the `str()` function to get an idea of the data structure of `hot_springs`.  
```{r}

```

## Change column classes
When making plots or performing statistical modeling, it may be important to convert column classes. For example, if a column contains categorical data (e.g., scientist names, treatment groups, etc.) it may be best to convert that column to a factor class.  

Recall that factors are categorical variables that have a fixed number of possible values called levels.  

What is the class of the scientist column? Let's change it to factor and then show the levels of that factor.  
```{r}

```

```{r}

```

Did our change work? We can also check the levels of each column.
```{r}

```

## Practice
1. Change the class of the springs column to factor.
```{r}

```

2. What are the levels of the springs column?
```{r}

```

## Importing Data from a Published Study
In your lab 3 folder there is another folder titled `data`. Inside the `data` folder there is a `.csv` titled `Gaeta_etal_CLC_data.csv`. Open this data and store them as an object called `fish`.  

The data are from: Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.  [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
```{r}

```

## Additional summary functions
Once data have been loaded, you need to get an idea of its structure, contents, and dimensions. I routinely run one or more of these commands when data are first imported.  

We can summarize our data frame with the`summary()` function.  
```{r}

```

`glimpse()` is another useful summary function.
```{r}

```

`nrow()` gives the numbers of rows.
```{r}

```

`ncol` gives the number of columns.
```{r}

```

`dim()` gives the dimensions.
```{r}

```

`names` gives the column names.
```{r}

```

`head()` prints the first n rows of the data frame.
```{r}

```

`tail()` prinst the last n rows of the data frame.
```{r}

```

`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable.
```{r}

```

## Practice
1. Load the data `mammal_lifehistories_v2.csv` and place it into a new object called `mammals`.
```{r}

```

2. Provide the dimensions of the data frame.
```{r}

```

3. Check the column names in the data frame. 
```{r}

```

4. Use `str()` to show the structure of the data frame and its individual columns; compare this to `glimpse()`. 
```{r}

```

5. Try the `table()` command to produce counts of mammal order.  
```{r}

```