---
title: "`mutate()`, and `if_else()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use `mutate()` to add columns in a dataframe.  
2. Use `mutate()` and `if_else()` to replace values in a dataframe.  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
For this lab, we will use the following two datasets:  

1. 1. Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.   [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  

2. S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.   [link](http://esapubs.org/archive/ecol/E084/093/)  

## Pipes `%>%` 
Recall that we use pipes to connect the output of code to a subsequent function. This makes our code cleaner and more efficient. One way we can use pipes is to attach the `clean_names()` function from janitor to the `read_csv()` output.  
```{r}
fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")
```

```{r}
mammals <- read_csv("data/mammal_lifehistories_v2.csv") %>% clean_names()
```

## `mutate()`  
Mutate allows us to create a new column from existing columns in a data frame. We are doing a small introduction here and will add some additional functions later. Let's convert the length variable from cm to millimeters and create a new variable called length_mm.  
```{r}
fish %>% 
  mutate(length_mm = length*10) %>% 
  select(fish_id, length, length_mm)
```

## Practice
1. Use `mutate()` to make a new column that is the half length of each fish: length_half = length/2. Select only fish_id, length, and length_half.
```{r}

```

## `mutate_all()`
This last function is super helpful when cleaning data. With "wild" data, there are often mixed entries (upper and lowercase), blank spaces, odd characters, etc. These all need to be dealt with before analysis.  

Here is an example that changes all entries to lowercase (if present).  
```{r}
mammals
```

```{r}
mammals %>%
  mutate_all(tolower)
```

Using the across function we can specify individual columns.
```{r}
mammals %>% 
  mutate(across(c("order", "family"), tolower))
```

## `if_else()`
We will briefly introduce `if_else()` here because it allows us to use `mutate()` but not have the entire column affected in the same way. In a sense, this can function like find and replace in a spreadsheet program. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  

Have a look at the data from mammals below. Notice that the values for newborn include `-999.00`. This is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.  
```{r}
mammals %>% 
  select(genus, species, newborn) %>% 
  arrange(newborn)
```

```{r}
mammals %>% 
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>% 
  arrange(newborn)
```

## Practice
1. We are interested in the family, genus, species and max life variables. Because the max life span for several mammals is unknown, the authors have use -999 in place of NA. Replace all of these values with NA in a new column titled `max_life_new`. Finally, sort the date in descending order by max_life_new. Which mammal has the oldest known life span?
```{r}

```









## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  