---
title: "summarize practice, `count()`, `across()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce clear, concise summaries using a variety of functions in `dplyr` and `janitor.`  
2. Use the `across()` operator to produce summaries across multiple variables.  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("skimr")
library("palmerpenguins")
```

## Review
The summarize() and group_by() functions are powerful tools that we can use to produce clean summaries of data. Especially when used together, we can quickly group variables of interest and save time. Let's do some practice with the [palmerpenguins(https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data to refresh our memory.
```{r}
glimpse(penguins)
```

As biologists, a good question that we may ask is how do the measured variables differ by island (on average)?
```{r}

```

Why do we have NA here? Do all of these penguins lack data?
```{r}

```

Well, that won't work so let's remove the NAs and recalculate.
```{r}

```

What if we are interested in the number of observations (penguins) by species and island?
```{r}
names(penguins)
```

```{r}

```

## Counts
Although these summary functions are super helpful, oftentimes we are mostly interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  

`count()` is an easy way of determining how many observations you have within a column. It acts like a combination of `group_by()` and `n()`.
```{r}
penguins %>% 
  count(island, sort = T) #sort=T sorts the column in descending order
```

Compare this with `summarize()` and `group_by()`.
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(n=n())
```

You can also use `count()` across multiple variables.
```{r}
penguins %>% 
  count(island, species, sort = T) # sort=T will arrange in descending order
```

For counts, I often prefer `tabyl()`. Lots of options are supported in [tabyl](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html)
```{r}
penguins %>% 
  tabyl(island, species)
```

```{r}
penguins %>% 
  tabyl(species, island) %>% 
  adorn_percentages() %>%
  adorn_pct_formatting(digits = 2)
```

## Practice
1. Produce a summary of the mean for bill_length_mm, bill_depth_mm, flipper_length_mm, and body_mass_g within Adelie penguins only. Be sure to provide the number of samples.
```{r}

```

2. How does the mean of `bill_length_mm` compare between penguin species?
```{r}

```

3. For some penguins, their sex is listed as NA. Where do these penguins occur?
```{r}

```

## `across()`
What about using `filter()` and `select()` across multiple variables? There is a function in dplyr called `across()` which is designed to work across multiple variables. Have a look at Rebecca Barter's awesome blog [here](http://www.rebeccabarter.com/blog/2020-07-09-across/) as I am following her below.  

What if we wanted to apply summarize in order to produce distinct counts over multiple variables; i.e. species, island, and sex? Although this isn't a lot of coding you can image that with a lot of variables it would be cumbersome.
```{r}
penguins %>%
  summarize(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_sex = n_distinct(sex))
```

```{r}
names(penguins)
```

By using `across()` we can reduce the clutter and make things cleaner. 
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```

This is very helpful for continuous variables.
```{r}
penguins %>%
  summarize(across(contains("mm"), mean, na.rm=T))
```

`group_by` also works.
```{r}
penguins %>%
  group_by(sex) %>% 
  summarize(across(contains("mm"), mean, na.rm=T))
```

Here we summarize across all variables.
```{r}
penguins %>%
  summarise_all(mean, na.rm=T)
```

Operators can also work, here I am summarizing across all variables except `species`, `island`, `sex`, and `year`.
```{r}
penguins %>%
  summarise(across(!c(species, island, sex, year), 
                   mean, na.rm=T))
```

All variables that include "bill"...all of the other dplyr operators also work.
```{r}
penguins %>%
  summarise(across(starts_with("bill"), mean, na.rm=T))
```

## Practice
1. Produce separate summaries of the mean and standard deviation for bill_length_mm, bill_depth_mm, and flipper_length_mm for each penguin species. Be sure to provide the number of samples.  
mean
```{r}

```

## Wrap-up  

Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)