---
title: "Data Visualization: `ggplot` part 3"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Build stacked barplots of categorical variables.  
2. Build side-by-side barplots using `position= "dodge"`.  
3. Build histograms and density plots.  
4. Create new categories with `case_when()`.

## Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

## Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
options(scipen=999) #cancels the use of scientific notation for the session
```

## Data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

```{r message=FALSE, warning=FALSE}
homerange <- readr::read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", na = c("", "NA", "\\"))
```

Note that when making this tutorial, we noticed a stray backtick in a cell under the primary method category. We want to set this to NA, so we include it in the list of NA indicators we have in our data (`"\\"`).
```{r}
homerange %>% 
  count(primarymethod)
```

## Barplots and multiple variables
Last time we explored the `fill` option in boxplots as a way to bring color to the plot; i.e. we filled by the same variable that we were plotting. What happens when we fill by a different categorical variable?  

Let's start by counting how many obervations we have in each taxonomic group.
```{r}
homerange %>% 
  count(taxon, sort=T)
```

Now let's make a barplot of these data.
```{r}
homerange %>% 
  ggplot(aes(x = taxon)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group")
```

By specifying `fill=trophic.guild` we build a stacked bar plot that shows the proportion of a given taxonomic group that is an herbivore or carnivore.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill = "Trophic Guild")
```

We can also have counts of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill = "Trophic Guild")
```

Here is the same plot oriented vertically.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")
```

We can also scale all bars to a percentage.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild))+
  geom_bar(position = position_fill())+ 
  scale_y_continuous(labels = scales::percent)+
  coord_flip()
```

## Practice
1. Make a barplot that shows locomotion type by `primarymethod`. Try both a stacked barplot and `position="dodge"`.


## Histograms and Density Plots
Histograms are frequently used by biologists; they show the distribution of continuous variables. As students, you almost certainly have seen histograms of grade distributions. Without getting into the statistics, a histogram `bins` the data and you specify the number of bins that encompass a range of observations. For something like grades, this is easy because the number of bins corresponds to the grades A-F. By default, R uses a formula to calculate the number of bins but some adjustment is often required.  

What does the distribution of body mass look like in the homerange data?
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram(alpha = 0.6, color = "black")+
  labs(title = "Distribution of Body Mass")
```

`Density plots` are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_density(fill="steelblue", alpha  =0.6, color = "black")+
  labs(title = "Distribution of Body Mass")
```

I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r message=FALSE}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(aes(y = ..density..), binwidth = .4, fill = "steelblue", alpha = 0.8, color = "black")+
  geom_density(color = "red")+
  labs(title = "Distribution of Body Mass")
```

## Practice
1. Make a histogram of `log10.hra`. Make sure to add a title.


2. Now plot the same variable using `geom_density()`.


3. Combine them both!


## Create Categories with mutate and case_when()
`case_when()` is a very handy function from `dplyr` which allows us to calculate a new variable from other variables. We use `case_when()` within `mutate()` to do this.`case_when()` allows us to specify multiple conditions. Let's reclassify the body mass variable into a new factor variable with small, medium, and large animals. 
```{r}
homerange <- homerange %>% 
             mutate(mass_category = case_when(log10.mass <= 1.75 ~ "small",
                                              log10.mass > 1.75 & log10.mass <= 2.75 ~ "medium",
                                              log10.mass > 2.75 ~ "large"))
```

Here we check how the newly created body mass categories compare across trophic guild. 
```{r}
homerange %>% 
  ggplot(aes(x = mass_category, fill = trophic.guild)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Observations by Taxon and Mass Category in Homerange Data",
       x = "Mass Category",
       fill = "Trophic Guild")
```

## Practice
1. Use case_when() to make a new column `range_category` that breaks down `log10.hra` into very small, small, medium, and large classes based on quartile.

2. How many and which taxonomic classes are represented in each of the `range_category` classes?


3. Isolate the small `range_category` and plot the range of `log10.mass` by taxonomic class.

## Faceting: `facet_wrap()`
Faceting is one of the most amazing features of ggplot. It allows you to make multi-panel plots for easy comparison. Here, I am making histograms of log10.mass for every taxon. We read the `~` in the `facet_wrap()` layer as `by`.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram() +
  facet_wrap(~taxon)
```

## Faceting: `facet_grid()`
As you can imagine, there are lots of options for faceting. Another useful facet type is `facet_grid`. This can be helpful when you want to facet by multiple categorical variables.  
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram() +
  facet_grid(~thermoregulation + locomotion)
```

## Practice
1. Use faceting to produce density distributions of log10.mass by taxonomic class.
```{r}
homerange %>% 
  ggplot(aes(x=log10.mass))+
  geom_density(fill="steelblue", alpha=0.4)+
  facet_wrap(~class)
```

## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)