---
title: "`summarize()` and `count()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce clean summaries of data using `summarize()`.  
4. Use `group_by()` in combination with `summarize()` to produce grouped summaries of data.  

## Load the tidyverse and janitor
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
For this lab, we will use the built-in data on mammal sleep patterns. From: _V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007_.
```{r}
msleep <- msleep
```

## `summarize()`
`summarize()` will produce summary statistics for a given variable in a data frame. For example, if you are asked to calculate the mean of `sleep_total` for large and small mammals you could do this using a combination of commands, but it isn't very efficient or clean. We can do better!  

For example, if we define "large" as having a `bodywt` greater than 200 then we get the following:
```{r}
large <- msleep %>% 
  filter(bodywt > 200) %>% 
  arrange(desc(bodywt))
large
```

```{r}
mean(large$sleep_total)
```

We can accomplish the same task using the `summarize()` function.  

Large mammals
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarize(mean_sleep_lg=mean(sleep_total))
```

Small mammals
```{r}
msleep %>% 
  filter(bodywt<10)%>% 
  summarize(mean_sleep_sm=mean(sleep_total))
```

You can also combine functions to make summaries for multiple variables.
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarize(mean_sleep_lg=mean(sleep_total),
            min_sleep_lg=min(sleep_total),
            max_sleep_lg=max(sleep_total),
            sd_sleep_lg=sd(sleep_total),
            total=n())
```

Maybe you want to summarize all of the numeric variables in the data frame. You can use the `where()` function to select only the numeric variables.
```{r}
msleep %>% 
  select(where(is.numeric)) %>% 
  summarize(across(everything(), mean, na.rm=TRUE))
```

## Practice
1. What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.
```{r}

```

`n_distinct()` is a handy way of cleanly presenting the number of distinct observations. Notice that there are multiple genera with over 100 in body weight.
```{r}
msleep %>% 
  filter(bodywt > 100)
```

Here we show the number of distinct genera over 100 in body weight.
```{r}
msleep %>% 
  filter(bodywt > 100) %>% 
  summarize(n_genera=n_distinct(genus))
```

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of a column), first() (returns first value in a column), last() (returns last value in a column) and n_distinct() (number of distinct values in a column).

## Practice
1. How many genera are represented in the msleep data frame?
```{r}

```

2. What are the min, max, and mean `sleep_total` for all of the mammals? Be sure to include the total n.
```{r}

```

## Counts
Although `summarize()` is helpful, oftentimes we are only interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  

For example, if we wanted to count the number of observations for each `vore` category in the `msleep` data frame we could do the following:
```{r}
msleep %>% 
  count(vore)
```

You can even combine multiple variables to get counts. For example, if we wanted to count the number of observations for each `vore` category within each `order` category we could do the following:
```{r}
msleep %>% 
  count(vore, order)
```

The tabyl() function from the janitor package is also very useful for producing counts and percentages.  
```{r}
msleep %>% 
  tabyl(vore)
```

## Practice
1. In the taxonomic order Carnivora, count the number of observations for each conservation status.
```{r}

```

2. Among herbivores, which taxonomic order is most represented?
```{r}

```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  