---
title: "Data Visualization: `ggplot` part 2"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce boxplots using `ggplot.`  
2. Customize labels on axes using `labs` and `themes`.  
3. Use `color`, `fill`, and `group` to customize plots and improve overall aesthetics.  

## Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)

## Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Review
In the last lab you were introduced to `ggplot`. Let's build on this by exploring more geom types and adding some aesthetics.  

## Data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1
```{r message=FALSE, warning=FALSE}
homerange <- readr::read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```

Remember to be mindful of NAs in the data; know how they are represented and how R has interpreted them.
```{r}
anyNA(homerange)
glimpse(homerange)
```

## Box Plots
Boxplots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range. In the case below, we are plotting `log10.mass` by taxonomic class in the homerange data. `geom_boxplot()` is the geom type for a standard boxplot. The center line in each box represents the median, not the mean.

Let's look at the variable `log10.mass` grouped by taxonomic class.
```{r}
homerange %>% 
  group_by(class) %>% 
  summarize(min_log10.mass=min(log10.mass),
            max_log10.mass=max(log10.mass),
            mean_log10.mass=mean(log10.mass),
            median_log10.mass=median(log10.mass))
```

```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass)) +
  geom_boxplot()
```

## Practice
1. There are more herbivores than carnivores in the homerange data, but how do their masses compare? Make a summary and boxplot that compares their masses. Use `log10.mass`.


2. Have a closer look at carnivorous mammals. Summarize and visualize the range of log10.mass by family.


3. Now use a boxplot to visualize the range of body mass by family of carnivore.


## Aesthetics: Labels
Now that we have practiced scatterplots, barplots, and boxplots we need to learn how to adjust their appearance to suit our needs. Let's start with labelling x and y axes.  

In this example from part 1, we explored the relationship between body mass and homerange.
```{r}
ggplot(data = homerange, mapping = aes(x = log10.mass, y = log10.hra)) +
  geom_point()
```

The plot looks clean, but it is incomplete. A reader unfamiliar with the data might have a difficult time interpreting the current labels. To add custom labels, we use the `labs` command.
```{r}
ggplot(data = homerange, mapping = aes(x = log10.mass, y = log10.hra)) +
  geom_point() +
  labs(title = "Mass vs. Homerange",
       x = "Mass (log10)",
       y = "Homerange (log10)")
```

We can improve the plot further by adjusting the size and face of the text. We do this using `theme()`.
```{r}
ggplot(data = homerange, mapping = aes(x = log10.mass, y = log10.hra)) +
  geom_point() +
  labs(title = "Mass vs. Homerange",
       x = "Mass (log10)",
       y = "Homerange (log10)")+
  theme(plot.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12))
```

The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` allows control of title position.
```{r}
ggplot(data = homerange, mapping = aes(x = log10.mass, y = log10.hra)) +
  geom_point() +
  labs(title = "Mass vs. Homerange",
       x = "Mass (log10)",
       y = "Homerange (log10)")+ 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```

## Practice
1. Make a barplot that shows the number of individuals per locomotion type. Be sure to provide a title and label the axes appropriately.


## Other Aesthetics
There are lots of options for aesthtics. An aesthetic can be assigned to either numeric or categorical data. `color` is a common option. Because color is included within the main `ggplot` command, it will color the points based on the category to which they belong and add a key.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = locomotion)) +
  geom_point()
```

Adding color within `geom_point()` changes the color of all of the points.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra)) +
  geom_point(color = "orange")
```

`size` adjusts the size of points relative to a continuous variable.
```{r}
options(scipen = 999) #disable scientific notation

homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, size = mean.mass.g)) +
  geom_point()
```

Here I am plotting `class` on the x-axis and `log10.mass` on the y-axis. I use `group` to make individual box plots for each taxon. I also use `fill` so I can associate the different taxa with a color coded key.
```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass, group = taxon, fill = taxon)) +
  geom_boxplot()
```

## Practice
1. Make a barplot that shows counts of ectotherms and endotherms. Label the axes, provide a title, and fill by thermoregulation type.


2. Make a boxplot that compares thermoregulation type by log10.mass. Group and fill by class. Label the axes and provide a title.


## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)