---
title: "`mutate()`, and `if_else()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use `mutate()` to add columns in a dataframe.  
2. Use `mutate()` and `if_else()` to replace values in a dataframe.  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
For this lab, we will use the following dataset:  
S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.   [link](http://esapubs.org/archive/ecol/E084/093/)  

## Pipes `%>%` 
Recall that we use pipes to connect the output of code to a subsequent function. This makes our code cleaner and more efficient. One way we can use pipes is to attach the `clean_names()` function from janitor to the `read_csv()` output.  
```{r}
mammals <- read_csv("data/mammal_lifehistories_v2.csv") %>% clean_names()
```

## `mutate()`  
Recall that mutate allows us to create a new column from existing columns in a data frame. Use `mutate()` to make a new column that converts gestation to years. Which animal has the longest gestation period?
```{r}

```

## `mutate()` and across()
This last function is super helpful when cleaning data. With "wild" data, there are often mixed entries (upper and lowercase), blank spaces, odd characters, etc. These all need to be dealt with before analysis.  

Here is an example that changes all entries to lowercase (if present).  
```{r}
mammals
```

```{r}
mammals %>% 
  mutate(across(everything(), tolower))
```

We can specify individual columns.
```{r}
mammals %>% 
  mutate(across(c("order", "family"), tolower))
```

Or, leave some out. This is an example of a lambda function ` ~ ` that applies `tolower`() to each column. The `.` represents the current column being processed.
```{r}
mammals %>%
  mutate(across(-1, ~ tolower(.))) #apply tolower to all columns except the first
```

## `if_else()`
We will briefly introduce `if_else()` here because it allows us to use `mutate()` but not have the entire column affected in the same way. In a sense, this can function like find and replace in a spreadsheet program. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  

Have a look at the data from mammals below. Notice that the values for newborn include `-999.00`. This is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.  
```{r}
mammals %>% 
  select(genus, species, newborn) %>% 
  arrange(newborn)
```

```{r}
mammals %>% 
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>% 
  arrange(-newborn)
```

## Practice
1. We are interested in the family, genus, species and max life variables. Because the max life span for several mammals is unknown, the authors have use -999 in place of NA. Replace all of these values with NA in a new column titled `max_life_new`. Then convert `max_life_new` into years. Finally, sort the date in descending order by max_life_new. Which mammal has the longest life span?
```{r}

```

2. Build a new data frame `msleep24` from the `msleep` data that: contains the `name` and `vore` variables along with a new column called `sleep_total_24` which is the amount of time a species sleeps expressed as a proportion of a 24-hour day. Restrict the `sleep_total_24` values to less than or equal to 0.3. Arrange the output in descending order.  
```{r}

```

Did `dplyr` do what we expected? How do we check our output? Remember, just because your code runs it doesn't mean that it did what you intended.
```{r}
#summary(msleep24)
```

Histograms are also a quick way to check the output.
```{r}
#hist(msleep24$sleep_total)
```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  