---
title: "`summarize()` and `group_by()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use a combination of `select()`, `filter()`, and `mutate()` to transform data frames.  
2. Use the `skimr` package to produce summaries of data.  
3. Produce clean summaries of data using `summarize()`.  
4. Use `group_by()` in combination with `summarize()` to produce grouped summaries of data.  

## Load the tidyverse and janitor
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Install `skimr`
```{r}
#install.packages("skimr")
library("skimr")
```

## Load the data
For this lab, we will use the built-in data on mammal sleep patterns. From: _V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007_.
```{r}
?msleep
```

We will also use [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data in the second part of lab.
```{r}
library("palmerpenguins")
```

## dplyr Practice
Let's do a bit more practice to make sure that we understand `select()`, `filter()`, and `mutate()`. Start by building a new data frame `msleep24` from the `msleep` data that: contains the `name` and `vore` variables along with a new column called `sleep_total_24` which is the amount of time a species sleeps expressed as a proportion of a 24-hour day. Restrict the `sleep_total_24` values to less than or equal to 0.3. Arrange the output in descending order.  
```{r}
names(msleep)
msleep <- msleep #stores msleep as a new object, but this isn't necessary
```

```{r}
msleep24 <- msleep %>% 
  mutate(sleep_total_24=sleep_total/24) %>% 
  select(name, vore, sleep_total_24, sleep_total) %>% 
  filter(sleep_total_24<=0.3) %>% 
  arrange(desc(sleep_total_24))
msleep24
```

Did `dplyr` do what we expected? How do we check our output? Remember, just because your code runs it doesn't mean that it did what you intended.
```{r}
summary(msleep24)
```

Try out the new function `skim()` as part of the `skimr` package.
```{r}
skim(msleep24)
```

Histograms are also a quick way to check the output.
```{r}
hist(msleep24$sleep_total_24)
```

## `summarize()`
`summarize()` will produce summary statistics for a given variable in a data frame. For example, if you are asked to calculate the mean of `sleep_total` for large and small mammals you could do this using a combination of commands, but it isn't very efficient or clean. We can do better!  
```{r}
head(msleep)
```

For example, if we define "large" as having a `bodywt` greater than 200 then we get the following:
```{r}
large <- msleep %>% 
  select(name, genus, bodywt, sleep_total) %>% 
  filter(bodywt > 200) %>% 
  arrange(desc(bodywt))
large
```

```{r}
mean(large$sleep_total)
```

We can accomplish the same task using the `summarize()` function to make things cleaner.  

Large mammals
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarize(mean_sleep_lg=mean(sleep_total))
```

Small mammals
```{r}
msleep %>% 
  filter(bodywt<10)%>% 
  summarize(mean_sleep_sm=mean(sleep_total))
```

You can also combine functions to make useful summaries for multiple variables.
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarize(mean_sleep_lg=mean(sleep_total),
            min_sleep_lg=min(sleep_total),
            max_sleep_lg=max(sleep_total),
            sd_sleep_lg=sd(sleep_total),
            total=n())
```

## Practice
1. What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.
```{r}
msleep %>% 
  filter(order=="Primates") %>% 
  summarize(mean_bodywt=mean(bodywt),
            min_bodywt=min(bodywt),
            max_bodywt=max(bodywt),
            total=n())
```

`n_distinct()` is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 100 in body weight.  

Notice that there are multiple genera with over 100 in body weight.
```{r}
msleep %>% 
  filter(bodywt > 100)
```

n_distinct() is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 100 in body weight.
```{r}
msleep %>% 
  filter(bodywt > 100) %>% 
  summarize(n_genera=n_distinct(genus))
```

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of a column), first() (returns first value in a column), last() (returns last value in a column) and n_distinct() (number of distinct values in a column).

## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>% 
  summarize(n_genera=n_distinct(genus))
```

Here is another option.
```{r}
msleep %>% 
  tabyl(genus)
```

2. What are the min, max, and mean `sleep_total` for all of the mammals? Be sure to include the total n.
```{r}
msleep%>% 
  summarize(mean_sleep_total=mean(sleep_total),
            minsleep_total=min(sleep_total),
            max_sleep_total=max(sleep_total),
            total=n())
```

## `group_by()`
The `summarize()` function is most useful when used in conjunction with `group_by()`. Although producing a summary of body weight for all of the mammals in the data set is helpful, what if we were interested in body weight by feeding ecology?
```{r}
msleep %>%
  group_by(vore) %>% #we are grouping by feeding ecology, a categorical variable
  summarize(min_bodywt = min(bodywt),
            max_bodywt = max(bodywt),
            mean_bodywt = mean(bodywt),
            total=n())
```

## Practice
1. Calculate mean brain weight by taxonomic order in the msleep data.
```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brain_wt=mean(brainwt))
```

2. What does `NA` mean? How are NA's being treated by the summarize function?
```{r}
msleep %>% 
  filter(order=="Carnivora") %>% 
  select(order, genus, brainwt)
```

3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea? Compare this to Carnivora. 
NaN is a special value that is returned when a mathematical operation is undefined. In this case, the mean of a column with NA's is undefined.  
```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brain_wt=mean(brainwt, na.rm=T))
```

```{r}
msleep %>% 
  filter(order=="Cetacea") %>% 
  select(order, genus, brainwt)
```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  