---
title: "Midterm review and `naniar()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Identify and manage NA's is data sets where NA's are represented in multiple ways.    
2. Use the `naniar` and `visdat` packages to help manage NA's in large data sets.  

## Breakout Rooms  
Please take 5-8 minutes to check over your answers to HW 6 in your group. If you are stuck, please remember that you can check the key in [Joel's repository](https://github.com/jmledford3115/BIS15LW2022_jledford).  

## Midterm 1 Review
Let's briefly review the questions from midterm 1 so you can get an idea of how I was thinking about the problems. Remember, there is more than one way to get at these answers, so don't worry if yours looks different than mine!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("skimr")
```

## Review
When working with outside or "wild" data, dealing with NA's is a fundamental part of the data cleaning or tidying process. Data scientists spend most of their time cleaning and transforming data- including managing NA's. There isn't a single approach that will always work so you need to be careful about using replacements strategies across an entire data set. And, as the data sets become larger NA's can become trickier to deal with.  

For the following, we will use life history data for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from:  
**S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.**  

## Practice
1. Load the mammals life history data and clean the names.  
```{r}
life_history <- read_csv("data/mammal_lifehistories_v3.csv") %>% clean_names()
```

2. Use one or more of the functions from the last lab to determine if there are NA's in the data, how they are represented, and where they are located.
```{r}

```

```{r}

```


3. Can we use a single approach to deal with NA's in this data set? Given what you learned in the previous lab, how would you manage the NA values?
```{r}

```

## `naniar`
`naniar` is a package that is built to manage NA's. Many of the functions it performs can also be performed using tidyverse functions, but it does provide some interesting alternatives.  

`miss_var_summary` provides a clean summary of NA's across the data frame.
```{r}
naniar::miss_var_summary(life_history)
```

Notice that `max_life` has no NA's. Does that make sense in the context of this data?
```{r}
hist(life_history$max_life)
```

```{r}
life_history <- 
  life_history %>% 
  mutate(max_life=na_if(max_life, 0))
```

```{r}
naniar::miss_var_summary(life_history)
```

We can also use `miss_var_summary` with `group_by()`. This helps us better evaluate where NA's are in the data.
```{r}
life_history %>%
  group_by(order) %>%
  select(order, wean_mass) %>% 
  naniar::miss_var_summary(order=T)
```

`naniar` also has a nice replace function which will allow you to precisely control which values you want replaced with NA's in each variable.
```{r}
life_history %>% 
  naniar::replace_with_na(replace = list(newborn = "not measured", weaning= -999, wean_mass= -999, afr= -999, max_life= 0, litter_size= -999, gestation= -999, mass= -999)) %>% 
  naniar::miss_var_summary()
```

## Practice
Let's practice evaluating NA's in a large data set. The data are compiled from [CITES](https://cites.org/eng). This is the international organization that tracks trade in endangered wildlife. You can find information about the data [here](https://www.kaggle.com/cites/cites-wildlife-trade-database).  

Some key information:  
[country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2)  

1. Import the data and do a little exploration. Be sure to clean the names if necessary.
```{r}

```

2. Use `naniar` to summarize the NA's in each variable.
```{r}

```

3. Try using `group_by()` with `naniar`. Look specifically at class and `exporter_reported_quantity`. For which taxonomic classes do we have a high proportion of missing export data?
```{r}

```

## Visualizing NAs
There is another package `visdat` that can be used to visualize the proportion of different classes of data, including missing data. But, it is limited by size.
```{r}
library(visdat)
```

```{r}
vis_dat(life_history) #classes of data
```

```{r}
vis_miss(life_history)
```

## Dealing with NA's in advance
If you are sure that you know how NA's are treated in the data, then you can deal with them in advance using `na()` as part of the `readr` package.
```{r}
life_history_advance <- 
  readr::read_csv(file = "data/mammal_lifehistories_v3.csv", 
                  na = c("NA", " ", ".", "-999")) #all NA, blank spaces, .,and -999 are treated as NA
```

```{r}
naniar::miss_var_summary(life_history_advance)
```

## Wrap-up  
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)